# (PART) Appendixes {-}

# Functions for Primary and Foreign Keys {-}

The chapter [Relational data](#relational-data) describes primary key and foreign key relations.
However, the dplyr package does not provide functions to check for these.

The `any_missing()` function is a helper function that will be used later. 
```{r}
any_missing <- function(x) any(is.na(x))
```

The `is_primary_key()` function checks whether variables are 
```{r}
#' Check whether variables are a primary key
#' 
#' Check whether a set of variables is a primary key for a data frame.
#' Unlike SQL databases, R data frames do not enforce a primary key 
#' constraint. This function checks whether a set of variables uniquely
#' identify a row.
#'
#' @param tbl A tbl.
#' @param ... One or more unquoted expressions separated by commas. 
#'   You can treat variable names like they are positions. This uses
#'   The same semantics as [dplyr::select()].
#' @return A logical vector of length one that is `TRUE` if the 
#'   variables are primary key, and `FALSE` otherwise.
is_primary_key <- function(tbl, ...) {
  variables <- quos(...)
  # no elements can be missing
  has_nulls <- summarise_at(tbl, vars(UQS(variables)), any_missing)
  if (any(as.logical(has_nulls))) {
    return(FALSE)
  }
  nrow(distinct(tbl, !!!variables)) == nrow(tbl)
}

foo <- tribble(
  ~a, ~b, ~c,
  1, NA, 1,
  2, 2, 1,
  3, 3, 3
)

is_key(foo, a)
is_key(foo, b)
is_key(foo, c)

is_key(foo, 1:3)
```

The `is_foreign_key()` function checks whether 
```{r is_foreign_key}
# check that columns in y are are foreign key of x
is_foreign_key <- function(x, y, by = NULL) {
  # check that y is a primary key
  if (!rlang::eval_tidy(quo(is_primary_key(y, !!!by)))) {
    return(FALSE)
  }
  # check that all x are found in y
  !nrow(anti_join(x, y, by = by))
}

```
