---
title: "Ch 10: Tibbles"
output: html_notebook
---

## Creating Tibbles

Why might we want to create non-syntactic variable names? Since variable names are often used as in plots (e.g. axis-titles) or headers in tables, where having spaces or other characters that are invalid R variable names is useful.
Those functions will have ways to use text other than the column.



## Tibbles vs. data.frame

Discuss the definition of a data frame.

What is the traditional R `data.frame`?

In general, discuss how this "dialect" of R relates to base R and other R that they will see.

Also, need to discuss types of variables.


```{r}
nycflights13::flights %>% 
  print(n = 10, width = Inf)
```

If this were printed in the console it would be much worse

```{r eval=FALSE}
as.data.frame(nycflights13::flights) 
```


### Subsetting

- **Note** Warnings about partial matching! What is it and why is it dangerous.

```{r}
df <- tibble(x = runif(5), y = rnorm(5))
df$x
df[["x"]]
df[[1]]
df %>% .$x
df %>% .[["x"]]
```

## Interacting with older code

**Note** Not all older functions work with tibbles (an example is Amelia); usually because they rely on quirks in `data.frame` behavior that `tibbles` "fix". Use `as.data.frame()` to turn a tibble back into a `data.frame`. This is usually because of `[` and the way it inconsistenly returns a vector or a data frame.
With tibbles `[` always returns a data frame

### Exercises

**TODO**





