---
title: "Ch 13: Relational Data"
output: html_notebook
---


## Prerequisites

```{r message=FALSE}
library(tidyverse)
library(nycflights13)
```

## nycflights13

**NOTES**

[nycflights13](https://cran.r-project.org/web/packages/nycflights13/index.html) is an example of a **data-only** R package. R packages can contain both functions and data.
Since data-sets can get large, often they can be packaged as their own dataset. These sorts of data-only R packages make it convenient for R users to access your data, but it should not be the only way you provide your research data. Not everyone uses R, so the original data should be provided in a program agnostic format (e.g. csv files). This also holds for those using Stata; they should not be distributing data in `.dta` format files specific to Stata (even if as we saw earlier, other programs can read that data.)
Another example of a data-only R package is [gapminder](https://cran.r-project.org/package=gapminder).

How does Hadley create his diagrams? 

The four tables in the **nycflights13** package:
```{r}
airlines
airports
planes
weather
```

### Exercises

**TODO**

## Keys

```{r}
planes %>% 
  count(tailnum) %>%
  filter(n > 1)
```

```{r}
weather %>%
  count(year, month, day, hour, origin) %>%
  filter(n > 1)
```



### Exercises
