---
title: "Ch 15: Factors"
output: html_notebook
---

Some notes

- Maybe ordered categorical factors like democracy?

Functions and packages

- **forcats**
- `factor`
- `fct_inorder`
- `levels`
- `readr::parse_factor`
- `fct_reorder`
- `fct_relevel`

## Introduction

```{r}
library(tidyverse)
library(forcats)
```


## Creating Factors

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
y1 <- factor(x1, levels = month_levels)
```

```{r}
y1 <- factor(x1, levels = month_levels)
y1
sort(y1)
```

```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
y2 <- factor(x2, levels = month_levels)
y2
```

No warnings with factor. Use `readr::parse_factor` for warning.

```{r}
y2 <- parse_factor(x2, levels = month_levels)
```

Change order of levels, e.g. first appearance:
```{r}
f1 <- factor(x1, levels = unique(x1))
f1
f2 <- x1 %>% factor() %>% fct_inorder()
f2
```

```{r}
levels(f2)
```

## GSS

```{r}
gss_cat
```

```{r}
gss_cat %>%
  count(race)
```

```{r}
ggplot(gss_cat, aes(race)) +
  geom_bar()
```

```{r}
ggplot(gss_cat, aes(race)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```

Use `drop = FALSE` to not drop valid values.

### Exercise

1. Explore the distribution of `rincome` (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

```{r}
rincome_plot <-
  gss_cat %>%
  ggplot(aes(rincome)) +
  geom_bar()
rincome_plot
```

The default bar chart labels are too squished to read.
One solution is to change the angle of the labels,
```{r}
rincome_plot +
  theme(axis.text.x = element_text(angle = 90))
```

But that's not natural either, because text is vertical, and we read horizontally.
So with long labels, it is better to flip it.
```{r}
rincome_plot +
  coord_flip()
```

This is better, but it unituively goes from low to high. It would help if the
scale is reversed. Also, if all the missing factors were differentiated.


3. What is the most common `relig` in this survey? What’s the most common `partyid`?

The most common `relig` is "Protestant"
```{r}
gss_cat %>%
  count(relig) %>%
  arrange(-n) %>%
  head(1)
```

The most common `partyid` is "Independent"
```{r}
gss_cat %>%
  count(partyid) %>% 
  arrange(-n) %>%
  head(1)
```


3. Which `relig` does `denom` (denomination) apply to? How can you find out with a table? How can you find out with a visualisation?

```{r}
levels(gss_cat$denom)
```

From the context it is clear that `denom` refers to "Protestant" (and unsurprising given that it is the largest category in `freq`).
Let's filter out the non-responses, no answers, others, not-applicable, or
no denomination, to leave only answers to denominations.
After doing that, the only remaining responses are "Protestant".
```{r}
gss_cat %>%
  filter(!denom %in% c("No answer", "Other", "Don't know", "Not applicable",
                       "No denomination")) %>%
  count(relig)

  
```

This is also clear in a scatter plot of `relig` vs. `denom` where the points are
proportional to the size of the number of answers (since otherwise there would be overplotting).
```{r}
gss_cat %>%
  count(relig, denom) %>%
  ggplot(aes(x = relig, y = denom, size = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```

## Modifying factor order

```{r}
relig <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )
ggplot(relig, aes(tvhours, relig)) + geom_point()
```

```{r}
ggplot(relig, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```

Move most data analysis out of the ggplot function call
```{r}
rincome <- gss_cat %>%
  group_by(rincome) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(rincome, aes(age, fct_reorder(rincome, age))) + geom_point()
```

