# Dates and Times


## Introduction

```{r message=FALSE}
library(tidyverse)
library(lubridate)
library(nycflights13)
```

## Creating date/times

This code is needed by exercises.
```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))
```

### Exercise 1 {.exercise}

> What happens if you parse a string that
> contains invalid dates?

```{r}
ret <- ymd(c("2010-10-10", "bananas"))
print(class(ret))
ret
```

It produces an `NA` and an warning message.

### Exercise 2 {.exercise}

What does the `tzone` argument to `today()` do? Why is it important?

It determines the time-zone of the date. Since different time-zones can have different dates, the value of `today()` can vary depending on the time-zone specified.

### Exercise 3 {.exercise}

> Use the appropriate **lubridate** function to
> parse each of the following dates:

```{r}
d1 <- "January 1, 2010"
mdy(d1)
d2 <- "2015-Mar-07"
ymd(d2)
d3 <- "06-Jun-2017"
dmy(d3)
d4 <- c("August 19 (2015)", "July 1 (2015)")
mdy(d4)
d5 <- "12/30/14" # Dec 30, 2014
mdy(d5)
```


## Date-Time Components

The following code from the chapter is used 

```{r}
sched_dep <- flights_dt %>%
  mutate(minute = minute(sched_dep_time)) %>%
  group_by(minute) %>%
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n())
```
In the previous code, the difference between rounded and un-rounded dates provides the within-period time.


### Exercise 1 {.exercise}

> How does the distribution of flight times 
> within a day change over the course of the year?

Let's try plotting this by month:
```{r}
flights_dt %>%
  mutate(time = hour(dep_time) * 100 + minute(dep_time),
         mon = as.factor(month
                         (dep_time))) %>%
  ggplot(aes(x = time, group = mon, color = mon)) +
  geom_freqpoly(binwidth = 100)
```

This will look better if everything is normalized within groups. The reason
that February is lower is that there are fewer days and thus fewer flights.
```{r}
flights_dt %>%
  mutate(time = hour(dep_time) * 100 + minute(dep_time),
         mon = as.factor(month
                         (dep_time))) %>%
  ggplot(aes(x = time, y = ..density.., group = mon, color = mon)) +
  geom_freqpoly(binwidth = 100)
```

At least to me there doesn't appear to much difference in within-day distribution over the year, but I maybe thinking about it incorrectly.

### Exercise 2 {.exercise}

Compare `dep_time`, `sched_dep_time` and `dep_delay`. Are they consistent? Explain your findings.

If they are consistent, then `dep_time = sched_dep_time + dep_delay`.

```{r}
flights_dt %>%
  mutate(dep_time_ = sched_dep_time + dep_delay * 60) %>%
  filter(dep_time_ != dep_time) %>%
  select(dep_time_, dep_time, sched_dep_time, dep_delay)
```

There exist discrepancies. It looks like there are mistakes in the dates.
These are flights in which the actual departure time is on the *next* day relative to the scheduled departure time. We forgot to account for this when creating the date-times. The code would have had to check if the departure time is less than the scheduled departure time. Alternatively, simply adding the delay time is more robust because it will automatically account for crossing into the next day.

### Exercise 3 {.exercise}

Compare `air_time` with the duration between the departure and arrival. Explain your findings. 

```{r}
flights_dt %>%
  mutate(flight_duration = as.numeric(arr_time - dep_time),
         air_time_mins = air_time,
         diff = flight_duration - air_time_mins) %>%
  select(origin, dest, flight_duration, air_time_mins, diff)
```

### Exercise 4 {.exercise}

How does the average delay time change over the course of a day? Should you use `dep_time` or `sched_dep_time`? Why?

Use `sched_dep_time` because that is the relevant metric for someone scheduling a flight. Also, using `dep_time` will always bias delays to later in the day since delays will push flights later.

```{r}
flights_dt %>%
  mutate(sched_dep_hour = hour(sched_dep_time)) %>%
  group_by(sched_dep_hour) %>%
  summarise(dep_delay = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay, x = sched_dep_hour)) +
  geom_point() +
  geom_smooth()
```

### Exercise 5 {.exercise}

On what day of the week should you leave if you want to minimize the chance of a delay?

Sunday has the lowest average departure delay time and the lowest average arrival delay time.

```{r}
flights_dt %>%
  mutate(dow = wday(sched_dep_time)) %>%
  group_by(dow) %>%
  summarise(dep_delay = mean(dep_delay),
            arr_delay = mean(arr_delay, na.rm = TRUE))
```

### Exercise 6 {.exercise}

What makes the distribution of `diamonds$carat` and `flights$sched_dep_time` similar?

```{r}
ggplot(diamonds, aes(x = carat)) + 
  geom_density()
```

In both `carat` and `sched_dep_time` there are abnormally large numbers of values are at nice "human" numbers. In `sched_dep_time` it is at 00 and 30 minutes. In carats, it is at 0, 1/3, 1/2, 2/3, 

```{r}
ggplot(diamonds, aes(x = carat %% 1 * 100)) +
  geom_histogram(binwidth = 1)
```

In scheduled departure times it is 00 and 30 minutes, and minutes
ending in 0 and 5.

```{r}
ggplot(flights_dt, aes(x = minute(sched_dep_time))) +
  geom_histogram(binwidth = 1)
```

### Exercise 7 {.exercise}

Confirm my hypothesis that the early departures of flights in minutes 20-30 and 50-60 are caused by scheduled flights that leave early. Hint: create a binary variable that tells you whether or not a flight was delayed.

At the minute level, there doesn't appear to be anything:
```{r}
flights_dt %>%
  mutate(early = dep_delay < 0,
         minute = minute(sched_dep_time)) %>%
  group_by(minute) %>%
  summarise(early = mean(early)) %>%
  ggplot(aes(x = minute, y = early)) +
  geom_point()
```


But if grouped in 10 minute intervals, there is a higher proportion of early flights during those minutes.

```{r}
flights_dt %>%
  mutate(early = dep_delay < 0,
         minute = minute(sched_dep_time) %% 10) %>%
  group_by(minute) %>%
  summarise(early = mean(early)) %>%
  ggplot(aes(x = minute, y = early)) +
  geom_point()
```


## Time Spans

### Exercise 1 {.exercise}

> Why is there `months()` but no `dmonths()`? 

There is no direct unambiguous value of months in seconds since months have differing numbers of days. 

- 31 days: January, March, May, July, August, October
- 30 days: April, Jun, September, November, December
- 28 or 29 days: February

Though in the past, in the pre-computer era, for arithmetic convenience, bankers adopted a 360 day year with 30 day months.

### Exercise 2 {.exercise}

> Explain `days(overnight * 1)` to someone who has just started learning R. How does it work? 

The variable `overnight` is equal to `TRUE` or `FALSE`.
If it is an overnight flight, this becomes 1 day, and if not, then overnight = 0, and no days are added to the date.

### Exercise 3 {.exercise}

> Create a vector of dates giving the first day of every month in 2015. Create a vector of dates giving the first day of every month in the current year.

A vector of the first day of the month for every month in 2015:
```{r}
ymd("2015-01-01") + months(0:11)
```

To get the vector of the first day of the month for *this* year, we first need to figure out what this year is, and get January 1st of it.
I can do that by taking `today()` and truncating it to the year using `floor_date`:
```{r}
floor_date(today(), unit = "year") + months(0:11)
```


### Exercise 4 {.exercise}

> Write a function that given your birthday (as a date), returns how old you are in years.

```{r}
age <- function(bday) {
  (bday %--% today()) %/% years(1)
}
age(ymd("1990-10-12"))
```

### Exercise 5 {.exercise}

> Why can’t `(today() %--% (today() + years(1)) / months(1)` work?

It appears to work. Today is a date. Today + 1 year is a valid endpoint for an interval. And months is period that is defined in this period.
```{r}
(today() %--% (today() + years(1))) %/% months(1)
(today() %--% (today() + years(1))) / months(1)
```


## Time Zones

No exercises. 

