# Graphics for communication

## Introduction

```{r}
library("tidyverse")
library("modelr")
library("lubridate")
```

## Label

### Exercises

#### Exercise 1  {.exercise}


`r BeginQuestion()`
Create one plot on the fuel economy data with customized `title`,
`subtitle`, `caption`, `x`, `y`, and `colour` labels.
`r EndQuestion()`


`r BeginAnswer()`

```{r}
ggplot(data = mpg,
       mapping = aes(x = reorder(class, hwy, median), y = hwy)) + 
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Compact Cars have > 10 Hwy MPG than Pickup Trucks",
    subtitle = "Comparing the median highway mpg in each class",
    caption = "Data from fueleconomy.gov",
    x = "Car Class",
    y = "Highway Miles per Gallon"
  )
```


`r EndAnswer()`

### Exercise 3 {.exercise}


`r BeginQuestion()`
The `geom_smooth()` is somewhat misleading because the `hwy` for large engines is skewed upwards due to the inclusion of lightweight sports cars with big engines. 
Use your modeling tools to fit and display
`r EndQuestion()`


`r BeginAnswer()`
a better model.

```{r}
ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point(aes(colour = class)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Subcompact cars show the greatest sensitivity to engine size",
    caption = "Data from fueleconomy.gov"
  )
```


```{r}
mod <- lm(hwy ~ class, data = mpg)
mpg %>%
  add_residuals(mod) %>%
  ggplot(aes(displ, resid)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Fuel efficiency decreases with engine size",
    subtitle = "Highway MPG for Cars After Subtracing Mean MPG of their Class",
    caption = "Data from fueleconomy.gov",
    x = "Highway MPG Relative to Class",
    y = "Engine Displacement"
  )
```

`r EndAnswer()`

#### Exercise 3 


`r BeginQuestion()`
Take an exploratory graphic that you've created in the last month, and add informative titles to make it easier for others to understand.
`r EndQuestion()`


`r BeginAnswer()`

This exercise is by is intrinsically left to readers.

`r EndAnswer()`

## Annotations

### Exercises

#### Exercise 1 {.exercise}


`r BeginQuestion()`
Use `geom_text()` with infinite positions to place text at the four corners of the plot.
`r EndQuestion()`


`r BeginAnswer()`

I can use similar code as the example in the text.
However, I need to use `vjust` and `hjust` in order for the text to appear in the plot, and these need to be different for each corner.
But, `geom_text` takes `hjust` and `vjust` as aesthetics, I can add them to the data and mappings, and use a single `geom_text` call instead of four different `geom_text` calls with four different data arguments, and four different values of `hjust` and `vjust` arguments.
```{r}
label <- tribble(
  ~displ, ~hwy, ~label, ~vjust, ~hjust,
     Inf,  Inf,    "Top right", "top", "right",
     Inf, -Inf,    "Bottom right", "bottom", "right",
    -Inf,  Inf,    "Top left", "top", "left",
    -Inf, -Inf,    "Bottom left", "bottom", "left"
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label, vjust = vjust, hjust = hjust), data = label)
```


`r EndAnswer()`

#### Exercise 2 {.exercise}


`r BeginQuestion()`
Read the documentation for `annotate()`. How can you use it to add a text label to a plot without having to create a tibble?
`r EndQuestion()`


`r BeginAnswer()`

With annotate you use what would be aesthetic mappings directly as arguments:
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  annotate("text", x = Inf, y = Inf, 
           label = "Increasing engine size is \nrelated to decreasing fuel economy.", vjust = "top", hjust = "right")
```


`r EndAnswer()`

#### Exercise 3 {.exercise}


`r BeginQuestion()`
How do labels with `geom_text()` interact with faceting? 
How can you add a label to a single facet?
How can you put a different label in each facet?
(Hint: think about the underlying data.)
`r EndQuestion()`


`r BeginAnswer()`

If the facet variable is not specified, the text is drawn in all facets.
```{r}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  label = "Increasing engine size is \nrelated to decreasing fuel economy."
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right",
            size = 2) +
  facet_wrap(~ class)
```

To draw the label in only one facet, add a column to the label data frame with the value of the faceting variable(s) in which to draw it.
```{r}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  class = "2seater",
  label = "Increasing engine size is \nrelated to decreasing fuel economy."
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right",
            size = 2) +
  facet_wrap(~ class)
```


To draw labels in different plots, simply have the facetting variable(s):
```{r}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  class = unique(mpg$class),
  label = stringr::str_c("Label for ", class)
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right",
            size = 3) +
  facet_wrap(~ class)
```

`r EndAnswer()`

#### Exercise 4 {.exercise}


`r BeginQuestion()`
What arguments to `geom_label()` control the appearance of the background box?
`r EndQuestion()`


`r BeginAnswer()`

- `label.padding`: padding around label
- `label.r`: amount of rounding in the corners
- `label.size`: size of label border


`r EndAnswer()`

#### Exercise 5 {.exercise}


`r BeginQuestion()`
What are the four arguments to `arrow()`? How do they work? 
Create a series of plots that demonstrate the most important options.
`r EndQuestion()`


`r BeginAnswer()`

The four arguments are: (from the help for [arrow](https://www.rdocumentation.org/packages/grid/versions/3.3.2/topics/arrow))
- `angle` : angle of arrow head
- `length` : length of the arrow head
- `ends`: ends of the line to draw arrow head
- `type`: `"open"` or `"close"`: whether the arrow head is a closed or open triangle


`r EndAnswer()`

## Scales

### Exercises

#### Exercise 1 {.exercise}


`r BeginQuestion()`
Why doesn’t the following code override the default scale?
`r EndQuestion()`


`r BeginAnswer()`

```{r}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_colour_gradient(low = "white", high = "red") +
  coord_fixed()
```

It does not override the default scale because the colors in `geom_hex` are set by the `fill` aesthetic, not the `color` aesthetic.

```{r}
ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_fill_gradient(low = "white", high = "red") +
  coord_fixed()
```


`r EndAnswer()`

#### Exercise 2 {.exercise}


`r BeginQuestion()`
The first argument to every scale is the label for the scale.
It is equivalent to using the `labs` function.
`r EndQuestion()`


`r BeginAnswer()`

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  scale_x_continuous("Engine displacement (L)") + 
  scale_y_continuous("Highway fuel economy (mpg)") +
  scale_colour_discrete("Car type")
```

`r EndAnswer()`

#### Exercise 3 {.exercise}


`r BeginQuestion()`
Change the display of the presidential terms by:

1.  Combining the two variants shown above.
2.  Improving the display of the y axis.
3.  Labeling each term with the name of the president.
4.  Adding informative plot labels.
5.  Placing breaks every 4 years (this is trickier than it seems!).

`r EndQuestion()`


`r BeginAnswer()`
```{r}
years <- lubridate::make_date(seq(year(min(presidential$start)),
             year(max(presidential$end)),
             by = 4), 1, 1)

presidential %>%
  mutate(id = 33 + row_number(),
         name_id = stringr::str_c(name, " (", id, ")"),
         name_id = factor(name_id, levels = name_id)) %>%
  ggplot(aes(start, name_id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = name_id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue")) +
    scale_y_discrete(NULL) +
    scale_x_date(NULL, breaks = years, date_labels = "'%y") +
    theme(panel.grid.minor = element_blank())
```

`r EndAnswer()`

#### Exercise 4 {.exercise}


`r BeginQuestion()`
Use `override.aes` to make the legend on the following plot easier to see.
`r EndQuestion()`


`r BeginAnswer()`

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20)
```

The problem with the legend is that the `alpha` value make the colors hard to see. So I'll override the alpha value to make the points solid in the legend.
```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20)  +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 1, override.aes = list(alpha = 1)))  
```

`r EndAnswer()`

