# Graphics for communication

## Introduction

```{r}
library("tidyverse")
```

## Label

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

### Exercises

**Ex 28.2.1**  Create one plot on the fuel economy data with customised `title`,
`subtitle`, `caption`, `x`, `y`, and `colour` labels.

```{r}
ggplot(data = mpg,
       mapping = aes(x = reorder(class, hwy, median), y = hwy)) + 
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Compact Cars have > 10 Hwy MPG than Pickup Trucks",
    subtitle = "Comparing the median highway mpg in each class",
    caption = "Data from fueleconomy.gov",
    x = "Car Class",
    y = "Highway Miles per Gallon"
  )
```


**Ex 28.2.3** The `geom_smooth()` is somewhat misleading because the `hwy` for
large engines is skewed upwards due to the inclusion of lightweight
sports cars with big engines. Use your modelling tools to fit and display
a better model.

```{r}
ggplot(mpg, aes(displ, hwy, colour = class)) +
  geom_point(aes(colour = class)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Subcompact cars show the greatest sensitivity to engine size",
    caption = "Data from fueleconomy.gov"
  )
```


```{r}
mod <- lm(hwy ~ class, data = mpg)
mpg %>%
  add_residuals(mod) %>%
  ggplot(mpg, aes(displ, resid, colour = class)) +
  geom_point(aes(colour = class)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Subcompact cars show the greatest sensitivity to engine size",
    caption = "Data from fueleconomy.gov"
  )
```

**Ex 28.2.3.**  Take an exploratory graphic that you've created in the last month, and add informative titles to make it easier for others to understand.

*Ignoring*
