# Introduction

## Acknowledgments {-}

All the credit should go to Garrett Grolemund and Hadley Wickham for writing the truly fantastic *R for Data Science* book,
without which these solutions would not exist---literally.

These solutions have benefitted from those who fixed problems and contributed 
solutions. Thank you to all of those who contributed on
[GitHub](https://github.com/jrnold/r4ds-exercise-solutions/graphs/contributors) 
(in alphabetical order):
```{r contributions,echo=FALSE,cache=FALSE,message=FALSE,results='asis'}
# adapted from usethis:::github_repo_spec
github_repo_spec <- function(path = here::here()) {
  stringr::str_c(gh::gh_tree_remote(path), collapse = "/")
}

# copied from usethis:::parse_repo_spec
parse_repo_spec <- function (repo_spec) {
    repo_split <- str_split(repo_spec, "/")[[1]]
    if (length(repo_split) != 2) {
        stop("`repo_spec` must be of the form 'owner/repo'")
    }
    list(owner = repo_split[[1]], repo = repo_split[[2]])
}

# copied from usethis:::spec_owner
spec_owner <- function(repo_spec) {
  parse_repo_spec(repo_spec)$owner
}

# copied from usethis:::spec_repo
spec_repo <- function(repo_spec) {
  parse_repo_spec(repo_spec)$repo
}

# adapted from from usethis:::use_tidy_thanks
create_thanks <- function(repo_spec = github_repo_spec(), 
                          excluded = NULL) {
  excluded <- excluded %||% spec_owner(repo_spec)
    res <- gh::gh("/repos/:owner/:repo/issues", 
                  owner = spec_owner(repo_spec), 
                  repo = spec_repo(repo_spec), state = "all", 
                  filter = "all", .limit = Inf)
    if (identical(res[[1]], "")) {
        message("No matching issues/PRs found.")
        return(invisible())
    }
    contributors <- purrr::map_chr(res, c("user", "login")) %>%
      discard(~ .x %in% excluded) %>%
      unique() %>%
      sort()
    contributors <- 
      glue::glue("[@{contributors}](https://github.com/{contributors})") %>%
      glue::glue_collapse(sep = ", ", width = Inf, last = ", and")
    glue::glue("(in alphabetical order) {contributors}.")
}
create_thanks()
```

## Organization {-}

The solutions are organized in the same order, and with the 
same numbers as in *R for Data Science*. Sections without
exercises are given a placeholder.

Like *R for Data Science*, packages used in each chapter are loaded in a code chunk at the start of the chapter in a section titled "Prerequisites".
If a package is used infrequently in solutions it may not 
be loaded, and functions using it will be called using the 
package name followed by two colons, as in `dplyr::mutate()` (see the *R for Data Science* [Introduction](http://r4ds.had.co.nz/introduction.html#running-r-code)).
We will also use `::` to be explicit about the package of a
function.

## Dependencies {-}

You can install all packages used in the solutions with the 
following line of code.
```{r eval=FALSE}
devtools::install_github("jrnold/r4ds-exercise-solutions")
```

## Bugs/Contributing {-}

If you find any typos, errors in the solutions, have an alternative solution,
or think the solution could be improved, I would love your contributions.
Please open an issue at <https://github.com/jrnold/r4ds-exercise-solutions/issues> or a pull request at
<https://github.com/jrnold/r4ds-exercise-solutions/pulls>.

## Updates {-}

```{r updates,results='asis',echo=FALSE,cache.extras=tools::md5sum(here::here("NEWS.md"))}
readr::read_file(here::here("NEWS.md")) %>%
  stringr::str_replace_all(stringr::regex("^## (.*)$", multiline = TRUE), 
                           "### \\1 {-}") %>%
  cat()
```

## Colophon {-}

HTML and PDF versions of this book are available at <https://jrnold.github.io/r4ds-exercise-solutions>.
The book is powered by [bookdown](https://bookdown.org) which makes it easy to turn R markdown files into HTML, PDF, and EPUB.

This book was built with:
```{r colonophon}
devtools::session_info()
```

<!-- match unopened div --><div>
