# Introduction

## Acknowledgments {-}

All the credit should go to Garrett Grolemund and Hadley Wickham for writing the truly fantastic *R for Data Science* book,
without which these solutions would not exist---literally.

This book was written in the open, with some people contributed pull requests to fix problems.
Thank you to all who contributed via [GitHub](https://github.com/jrnold/r4ds-exercise-solutions/graphs/contributors).

```{r contributions,echo=FALSE,cache=FALSE,message=FALSE,results='asis'}
library("dplyr")
clean_author <- function(x) {
  if_else(stringr::str_detect(x, "@users\\.noreply\\.github\\.com"),
          stringr::str_replace(x, "(.*?)@users\\.noreply\\.github\\.com", 
                               "\\\\@\\1"),
          stringr::str_c("<", x, ">"))
}

excluded_authors <- c("jeffrey.arnold@gmail.com")
contribs <- git2r::contributions(by = "author") %>%
  filter(!author %in% excluded_authors) %>%
  mutate(author = clean_author(author)) %>%
  arrange(author) 

cat("Thanks go to all contributers in alphabetical order: ")
cat(paste0(contribs$author, collapse = ", "))
cat(".\n")
```

## Organization {-}

The solutions are organized in the same order, and with the 
same numbers as in *R for Data Science*. Sections without
exercises are given a placeholder.

Like *R for Data Science*, packages used in each chapter are loaded in a code chunk at the start of the chapter in a section titled "Prerequisites".
If a package is used infrequently in solutions it may not 
be loaded, and functions using it will be called using the 
package name followed by two colons, as in `dplyr::mutate()` (see the *R for Data Science* [Introduction](http://r4ds.had.co.nz/introduction.html#running-r-code)).
We will also use `::` to be explicit about the package of a
function.

## Dependencies {-}

You can install all packages used in the solutions with the 
following line of code.
```{r eval=FALSE}
devtools::install_github("jrnold/r4ds-exercise-solutions")
```

## Bugs/Contributing {-}

If you find any typos, errors in the solutions, have an alternative solution,
or think the solution could be improved, I would love your contributions.
Please open an issue at <https://github.com/jrnold/r4ds-exercise-solutions/issues> or a pull request at
<https://github.com/jrnold/r4ds-exercise-solutions/pulls>.

## Updates {-}

```{r updates,results='asis',echo=FALSE,cache.extras=tools::md5sum(here::here("NEWS.md"))}
readr::read_file(here::here("NEWS.md")) %>%
  stringr::str_replace_all(stringr::regex("^## (.*)$", multiline = TRUE), 
                           "### \\1 {-}") %>%
  cat()
```

## Colophon {-}

HTML and PDF versions of this book are available at <https://jrnold.github.io/r4ds-exercise-solutions>.
The book is powered by [bookdown](https://bookdown.org) which makes it easy to turn R markdown files into HTML, PDF, and EPUB.

This book was built with:
```{r colonophon}
devtools::session_info()
```

<!-- match unopened div --><div>
