# Iteration

## Prerequisites
 
```{r}
library(tidyverse)
```

## For Loops

### Exercises


Write for loops to:

1. Compute the mean of every column in `mtcars`.
2. Determine the type of each column in `nycflights13::flights`.
3. Compute the number of unique values in each column of `iris`.
4. Generate 10 random normals for each of $\mu = -10$, 0, 10, and 100.

Think about the output, sequence, and body before you start writing the loop.

To compute the mean of every column in `mtcars`.
```{r}
output <- vector("double", ncol(mtcars))
names(output) <- names(mtcars)
for (i in names(mtcars)) {
  output[i] <- mean(mtcars[[i]])
}
output
```

Determine the type of each column in `nycflights13::flights`.
Note that we need to use a `list`, not a character vector, since the class can have multiple values.
```{r}
data("flights", package = "nycflights13")
output <- vector("list", ncol(flights))
names(output) <- names(flights)
for (i in names(flights)) {
  output[[i]] <- class(flights[[i]])
}
output
```


```{r}
data(iris)
iris_uniq <- vector("double", ncol(iris))
names(iris_uniq) <- names(iris)
for (i in names(iris)) {
  iris_uniq[i] <- length(unique(iris[[i]]))
}
iris_uniq
```

```{r}
# number to draw
n <- 10
# values of the mean
mu <- c(-10, 0, 10, 100)
normals <- vector("list", length(mu))
for (i in seq_along(normals)) {
  normals[[i]] <- rnorm(n, mean = mu[i])
}
normals
```

However, we don't need a `for` loop for this since `rnorm` recycles means.
```{r}
matrix(rnorm(n * length(mu), mean = mu), ncol = n)
```

2. Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
```

`str_c` already works with vectors, so simply use `str_c` with the `collapse` argument to return a single string.
```{r}
stringr::str_c(letters, collapse = "")
```

For this I'm going to rename the variable `sd` to something different because `sd` is the name of the function we want to use.
```{r}
x <- sample(100)
sd. <- 0
for (i in seq_along(x)) {
  sd. <- sd. + (x[i] - mean(x)) ^ 2
}
sd. <- sqrt(sd. / (length(x) - 1))
sd.
```

We could simply use the `sd` function.
```{r}
sd(x)
```
Or if there was a need to use the equation (e.g. for pedagogical reasons), then
the functions `mean` and `sum` already work with vectors:
```{r}
sqrt(sum((x - mean(x)) ^ 2) / (length(x) - 1))
```


```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out
```
The code above is calculating a cumulative sum. Use the function `cumsum`
```{r}
all.equal(cumsum(x),out)
```
