# Iteration

## Introduction

```{r message=FALSE}
library("tidyverse")
library("stringr")
```
The package **microbenchmark** is used for timing code
```{r}
library("microbenchmark")
```


## For Loops

### Exercise 1 {.exercise}

> Write for loops to:
> 
> 1. Compute the mean of every column in `mtcars`.
> 2. Determine the type of each column in `nycflights13::flights`.
> 3. Compute the number of unique values in each column of `iris`.
> 4. Generate 10 random normals for each of $\mu = -10$, 0, 10, and 100.

> Think about the output, sequence, and body before you start writing the loop.

To compute the mean of every column in `mtcars`.
```{r}
output <- vector("double", ncol(mtcars))
names(output) <- names(mtcars)
for (i in names(mtcars)) {
  output[i] <- mean(mtcars[[i]])
}
output
```

Determine the type of each column in `nycflights13::flights`.
Note that we need to use a `list`, not a character vector, since the class can have multiple values.
```{r}
data("flights", package = "nycflights13")
output <- vector("list", ncol(flights))
names(output) <- names(flights)
for (i in names(flights)) {
  output[[i]] <- class(flights[[i]])
}
output
```


```{r}
data(iris)
iris_uniq <- vector("double", ncol(iris))
names(iris_uniq) <- names(iris)
for (i in names(iris)) {
  iris_uniq[i] <- length(unique(iris[[i]]))
}
iris_uniq
```

```{r}
# number to draw
n <- 10
# values of the mean
mu <- c(-10, 0, 10, 100)
normals <- vector("list", length(mu))
for (i in seq_along(normals)) {
  normals[[i]] <- rnorm(n, mean = mu[i])
}
normals
```

However, we don't need a `for` loop for this since `rnorm` recycles means.
```{r}
matrix(rnorm(n * length(mu), mean = mu), ncol = n)
```

### Exercise 2 {.exercise} 

> Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:

```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
```

`str_c` already works with vectors, so simply use `str_c` with the `collapse` argument to return a single string.
```{r}
stringr::str_c(letters, collapse = "")
```

For this I'm going to rename the variable `sd` to something different because `sd` is the name of the function we want to use.


```{r}
x <- sample(100)
sd. <- 0
for (i in seq_along(x)) {
  sd. <- sd. + (x[i] - mean(x)) ^ 2
}
sd. <- sqrt(sd. / (length(x) - 1))
sd.
```

We could simply use the `sd` function.
```{r}
sd(x)
```
Or if there was a need to use the equation (e.g. for pedagogical reasons), then
the functions `mean` and `sum` already work with vectors:
```{r}
sqrt(sum((x - mean(x)) ^ 2) / (length(x) - 1))
```


```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out
```

The code above is calculating a cumulative sum. Use the function `cumsum`
```{r}
all.equal(cumsum(x),out)
```

### Exercise 3 {.exercise}

> Combine your function writing and for loop skills:
>
>     1. Write a for loop that `prints()` the lyrics to the children's song  "Alice the camel".
>
>    1. Convert the nursery rhyme "ten in the bed" to a function. Generalise 
>       it to any number of people in any sleeping structure.

>    1. Convert the song "99 bottles of beer on the wall" to a function.
>       Generalise to any number of any vessel containing any liquid on 
>       any surface.
    
The lyrics for ``Alice the Camel'' can be found at <http://www.kididdles.com/lyrics/a012.html>.

We'll look from five to no humps, and print out a different last line if there are no humps. This uses `cat` instead of `print`, so it looks nicer.
```{r}
humps <- c("five", "four", "three", "two", "one", "no")
for (i in humps) {
  cat(str_c("Alice the camel has ", rep(i, 3), " humps.",
             collapse = "\n"), "\n")
  if (i == "no") {
    cat("Now Alice is a horse.\n")
  } else {
    cat("So go, Alice, go.\n")
  }
  cat("\n")
}
```

The lyrics for [Ten in the Bed](http://supersimplelearning.com/songs/original-series/one/ten-in-the-bed/):
```{r}
numbers <- c("ten", "nine", "eight", "seven", "six", "five",
             "four", "three", "two", "one")
for (i in numbers) {
  cat(str_c("There were ", i, " in the bed\n"))
  cat("and the little one said\n")
  if (i == "one") {
    cat("I'm lonely...")
  } else {
    cat("Roll over, roll over\n")
    cat("So they all rolled over and one fell out.\n")
  }
  cat("\n")
}
```

For the bottles of beer, I define a helper function to correctly print the number of bottles.
```{r}
bottles <- function(i) {
  if (i > 2) {
   bottles <- str_c(i - 1, " bottles")
  } else if (i == 2) {
   bottles <- "1 bottle"
  } else {
   bottles <- "no more bottles"
  }
  bottles
}

beer_bottles <- function(n) {
  # should test whether n >= 1.
  for (i in seq(n, 1)) {
     cat(str_c(bottles(i), " of beer on the wall, ", bottles(i), " of beer.\n"))
     cat(str_c("Take one down and pass it around, ", bottles(i - 1),
                " of beer on the wall.\n\n"))
  }
  cat("No more bottles of beer on the wall, no more bottles of beer.\n")
  cat(str_c("Go to the store and buy some more, ", bottles(n), " of beer on the wall.\n"))
}
beer_bottles(3)
```


#### Exercise 4 {.exercise} 

> It's common to see for loops that don't preallocate the output and instead increase the length of a vector at each step:
    
```{r, eval = FALSE}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```

> I'll use the package **microbenchmark** to time this.
> The `microbenchmark` function will run an R expression a number of times and time it.

Define a function that appends to an integer vector.
```{r}
add_to_vector <- function(n) {
  output <- vector("integer", 0)
  for (i in seq_len(n)) {
    output <- c(output, i)
  }
  output  
}
microbenchmark(add_to_vector(10000), times = 3)
```

And one that pre-allocates it.
```{r}
add_to_vector_2 <- function(n) {
  output <- vector("integer", n)
  for (i in seq_len(n)) {
    output[[i]] <- i
  }
  output
}
microbenchmark(add_to_vector_2(10000), times = 3)
```

The pre-allocated vector is about **100** times faster!
You may get different answers, but the longer the vector and the bigger the objects, the more that pre-allocation will outperform appending.


## For loop variations


### Exercise 1 {.exercise} 

> Imagine you have a directory full of CSV files that you want to read in.
> You have their paths in a vector, 
`files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now
want to read each one with `read_csv()`. Write the for loop that will 
load them into a single data frame. 

I will pre-allocate a list, read each file as data frame into an element in that list.
This creates a list of data frames.
I then use `bind_rows` to create a single data frame from the list of data frames.
```{r eval = FALSE}
df <- vector("list", length(files))
for (fname in seq_along(files)) {
  df[[i]] <- read_csv(files[[i]])
}
df <- bind_rows(df)
```


### Exercise 2 {.exercise} 

> What happens if you use `for (nm in names(x))` and `x` has no names?
> What if only some of the elements are named?
> What if the names are not unique?

Let's try it out and see what happens.

When there are no names for the vector, it does not run the code in the loop (it runs zero iterations of the loop):
```{r}
x <- 1:3
print(names(x))
for (nm in names(x)) {
  print(nm)
  print(x[[nm]])
}
```
Note that the length of `NULL` is zero:
```{r}
length(NULL)
```

If there only some names, then we get an error if we try to access an element without a name. 
However, oddly, `nm == ""` when there is no name.
```{r}
x <- c(a = 1, 2, c = 3)
names(x)
```
```{r error = TRUE}
for (nm in names(x)) {
  print(nm)
  print(x[[nm]])
}
```

Finally, if there are duplicate names, then `x[[nm]]` will give the *first* element with that name.
There is no way to access elements with duplicate names.
```{r}
x <- c(a = 1, a = 2, c = 3)
names(x)
```
```{r}
for (nm in names(x)) {
  print(nm)
  print(x[[nm]])
}
```


### Exercise 3 {.exercise}  


> Write a function that prints the mean of each numeric column in a data  frame, along with its name. 
> For example, `show_mean(iris)` would print:
    
```{r, eval = FALSE}
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
```
    
> (Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

There may be other functions to do this, but I'll use `str_pad`, and `str_length` to ensure that the space given to the variable names is the same.
I messed around with the options to `format` until I got two digits .
```{r}
show_mean <- function(df, digits = 2) {
  # Get max length of any variable in the dataset
  maxstr <- max(str_length(names(df)))
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      cat(str_c(str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
                format(mean(df[[nm]]), digits = digits, nsmall = digits),
                sep = " "),
          "\n")
    }
  }
}
show_mean(iris)
```

### Exercise 4 {.exercise} 

> What does this code do? How does it work?

```{r}
trans <- list( 
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
```
```{r eval=FALSE}
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
```

This code mutates the `disp` and `am` columns:

- `disp` is  multiplied by 0.0163871
- `am` is replaced by a factor variable.

The code works by looping over a named list of functions.
It calls the named function in the list on the column of `mtcars` with the same name, and replaces the values of that column.

E.g. this is a function:
```{r eval=FALSE}
trans[["disp"]]
```
This applies the function to the column of `mtcars` with the same name
```{r eval=FALSE}
trans[["disp"]](mtcars[["disp"]])
```



## For loops vs. functionals


### Exercise 1 {.exercise}

> Read the documentation for `apply()`. In the 2d case, what two for loops does it generalize.

It generalizes looping over the rows or columns of a matrix or data-frame. 

### Exercise 2 {.exercise}

> Adapt `col_summary()` so that it only applies to numeric columns.
> You might want to start with an `is_numeric()` function that returns a logical vector that has a `TRUE` corresponding to each numeric column.

The original `col_summary()` function is,
```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
```

The adapted version is,
```{r}
col_summary2 <- function(df, fun) {
  # test whether each colum is numeric
  numeric_cols <- vector("logical", length(df))
  for (i in seq_along(df)) {
    numeric_cols[[i]] <- is.numeric(df[[i]])
  }
  # indexes of numeric columns
  idxs <- seq_along(df)[numeric_cols]
  # number of numeric columns
  n <- sum(numeric_cols)
  out <- vector("double", n)
  for (i in idxs) {
    out[i] <- fun(df[[i]])
  }
  out
}
```

Let's test that it works,
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = letters[1:10],
  d = rnorm(10)
)
col_summary2(df, mean)
```


## The map functions

### Exercise 1 {.exercise}

> Write code that uses one of the map functions to:
> 
>     1. Compute the mean of every column in `mtcars`.
>    1. Determine the type of each column in `nycflights13::flights`.
>    1. Compute the number of unique values in each column of `iris`.
>    1. Generate 10 random normals for each of $\mu = -10$, $0$, $10$, and $100$.
    
To calculate the mean of every column in `mtcars`:
```{r}
map_dbl(mtcars, mean)
```

To calculate the type of every column in `nycflights13::flights`. 
```{r}
map(nycflights13::flights, class)
```
I had to use `map` rather than `map_chr` since the class
Though if by type, `typeof` is meant:
```{r}
map_chr(nycflights13::flights, typeof)
```

To calculate the number of unique values in each column of `iris`:
```{r}
map_int(iris, ~ length(unique(.)))
```

To generate 10 random normals for each of $\mu = -10$, $0$, $10$, and $100$:
```{r}
map(c(-10, 0, 10, 100), rnorm, n = 10)
```



### Exercise 2 {.exericse}  


> How can you create a single vector that for each column in a data frame indicates whether or not it's a factor?

Use `map_lgl` with the function `is.factor`,
```{r}
map_lgl(mtcars, is.factor)
```


### Exercise 3 {.exercise} 

> What happens when you use the map functions on vectors that aren't lists? 
> What does `map(1:5, runif)` do? 
> Why?

The function `map` applies the function to each element of the vector.
```{r}
map(1:5, runif)
```

    
### Exercise 4 {.exercise} 

> What does `map(-2:2, rnorm, n = 5)` do? Why? >
> What does `map_dbl(-2:2, rnorm, n = 5)` do?
> Why?

This takes samples of `n = 5` from normal distributions of means -2, -1, 0, 1, and 2, and returns a list with each element a numeric vectors of length 5.
```{r}
map(-2:2, rnorm, n = 5)
```

However, if we use `map_dbl` it throws an error. `map_dbl` expects the function 
to return a numeric vector of length one.
```{r error = TRUE}
map_dbl(-2:2, rnorm, n = 5)
```

If we wanted a numeric vector, we could use `map` followed by `flatten_dbl`,
```{r}
flatten_dbl(map(-2:2, rnorm, n = 5))
```


### Exercise 5 {.exercise} 

> Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the anonymous function. 

```{r}
map(list(mtcars), ~ lm(mpg ~ wt, data = .))
```

## Dealing with Failure

No exercises

## Mapping over multiple arguments

No exercises

## Walk

No exercises

## Other patterns of for loops


### Exercise 1 {.exercise}

> Implement your own version of `every()` using a for loop. 
> Compare it with `purrr::every()`. 
> What does purrr's version do that your version doesn't?


```{r}
# Use ... to pass arguments to the function
every2 <- function(.x, .p, ...) {
  for (i in .x) {
    if (!.p(i, ...)) {
      # If any is FALSE we know not all of then were TRUE
      return(FALSE)
    }
  }
  # if nothing was FALSE, then it is TRUE
  TRUE  
}

every2(1:3, function(x) {x > 1})
every2(1:3, function(x) {x > 0})
```

The function `purrr::every` does fancy things with `.p`, like taking a logical vector instead of a function, or being able to test part of a string if the elements of `.x` are lists.

### Exercise 2 {.exercise} 

> Create an enhanced `col_sum()` that applies a summary function to every numeric column in a data frame.

**Note:** this question has a typo. It is referring to `col_summary`. 

I will use `map` to apply the function to all the columns, and `keep` to only select numeric columns.
```{r}
col_sum2 <- function(df, f, ...) {
  map(keep(df, is.numeric), f, ...)
}
```

```{r}
col_sum2(iris, mean)
```



### Exercise 3 {.exercise} 

> Create possible base R equivalent of `col_sum()` is:

```{r}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]
  sapply(df_num, f)
}
```

> But it has a number of bugs as illustrated with the following inputs:

```{r, eval=FALSE}
df <- tibble(
  x = 1:3, 
  y = 3:1,
  z = c("a", "b", "c")
)
# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```

> What causes these bugs? 


The problem is that `sapply` does not always return numeric vectors.
If no columns are selected, instead of returning an empty numeric vector, it returns an empty list. 
This causes an error since we can't use a list with `[`.
```{r}
sapply(df[0], is.numeric)
```
```{r}
sapply(df[1], is.numeric)
```
```{r}
sapply(df[1:2], is.numeric)
```
