# Iteration

## Introduction

```{r message=FALSE}
library("tidyverse")
library("stringr")
```
The package **microbenchmark** is used for timing code
```{r}
library("microbenchmark")
```

## For Loops

### Exercise <span class="exercise-number">21.2.1.1</span> {.unnumbered .exercise}

<div class="question">

Write for-loops to:

1.  Compute the mean of every column in `mtcars`.
1.  Determine the type of each column in `nycflights13::flights`.
1.  Compute the number of unique values in each column of `iris`.
1.  Generate 10 random normals for each of $\mu = -10$, 0, 10, and 100.

</div>

<div class="answer">
The answers for each part are below.

1.  To compute the mean of every column in `mtcars`.
    ```{r}
    output <- vector("double", ncol(mtcars))
    names(output) <- names(mtcars)
    for (i in names(mtcars)) {
      output[i] <- mean(mtcars[[i]])
    }
    output
    ```

1.  Determine the type of each column in `nycflights13::flights`.
    Note that we need to use a `list`, not a character vector, since the class can have multiple values.
    ```{r}
    data("flights", package = "nycflights13")
    output <- vector("list", ncol(flights))
    names(output) <- names(flights)
    for (i in names(flights)) {
      output[[i]] <- class(flights[[i]])
    }
    output
    ```

1.  To compute the number of unique values in each column of the `iris` dataset.
    ```{r}
    data("iris")
    iris_uniq <- vector("double", ncol(iris))
    names(iris_uniq) <- names(iris)
    for (i in names(iris)) {
      iris_uniq[i] <- length(unique(iris[[i]]))
    }
    iris_uniq
    ```

1.  To generate 10 random normals for each of $\mu = -10$, 0, 10, and 100.
    ```{r}
    # number to draw
    n <- 10
    # values of the mean
    mu <- c(-10, 0, 10, 100)
    normals <- vector("list", length(mu))
    for (i in seq_along(normals)) {
      normals[[i]] <- rnorm(n, mean = mu[i])
    }
    normals
    ```

However, we don't need a for loop for this since `rnorm()` recycle the `mean` argument.
```{r}
matrix(rnorm(n * length(mu), mean = mu), ncol = n)
```
</div>

### Exercise <span class="exercise-number">21.2.1.2</span> {.unnumbered .exercise}

<div class="question">
Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:
</div>

<div class="answer">

```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}
out
```

Since `str_c()` already works with vectors, use `str_c()` with the `collapse` argument to return a single string.
```{r}
stringr::str_c(letters, collapse = "")
```

For this I'm going to rename the variable `sd` to something different because `sd` is the name of the function we want to use.

```{r}
x <- sample(100)
sd. <- 0
for (i in seq_along(x)) {
  sd. <- sd. + (x[i] - mean(x)) ^ 2
}
sd. <- sqrt(sd. / (length(x) - 1))
sd.
```

We could simply use the `sd` function.
```{r}
sd(x)
```
Or if there was a need to use the equation (e.g. for pedagogical reasons), then
the functions `mean()` and `sum()` already work with vectors:
```{r}
sqrt(sum((x - mean(x)) ^ 2) / (length(x) - 1))
```

```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}
out
```

The code above is calculating a cumulative sum. Use the function `cumsum()`
```{r}
all.equal(cumsum(x),out)
```

</div>

### Exercise <span class="exercise-number">21.2.1.3</span> {.unnumbered .exercise}

<div class="question">

Combine your function writing and for loop skills:

1.  Write a for loop that `prints()` the lyrics to the children's song  "Alice the camel".

1.  Convert the nursery rhyme "ten in the bed" to a function. Generalize it to any number of people in any sleeping structure.

1.  Convert the song "99 bottles of beer on the wall" to a function.
Generalize to any number of any vessel containing any liquid on  surface.

</div>

<div class="answer">

The answers to each part follow.

1.  The lyrics for [Alice the Camel](http://www.kididdles.com/lyrics/a012.html) are:

    > Alice the camel has five humps. \
    > Alice the camel has five humps. \
    > Alice the camel has five humps. \
    > So go, Alice, go.

    This verse is repeated, each time with one fewer hump,
    until there are no humps.
    The last verse, with no humps, is:

    > Alice the camel has no humps. \
    > Alice the camel has no humps. \
    > Alice the camel has no humps. \
    > Now Alice is a horse.

    We'll iterate from five to no humps, and print out a different last line if there are no humps.
    ```{r}
    humps <- c("five", "four", "three", "two", "one", "no")
    for (i in humps) {
      cat(str_c("Alice the camel has ", rep(i, 3), " humps.",
                 collapse = "\n"), "\n")
      if (i == "no") {
        cat("Now Alice is a horse.\n")
      } else {
        cat("So go, Alice, go.\n")
      }
      cat("\n")
    }
    ```

1.  The lyrics for [Ten in the Bed](https://www.kididdles.com/lyrics/t003.html) are:

    > Here we go! \
    > There were ten in the bed \
    > and the little one said, \
    > “Roll over, roll over.” \
    > So they all rolled over and one fell out.

    This verse is repeated, each time with one fewer in the bed, until there is one left. That last verse is:

    > One!
    > There was one in the bed \
    > and the little one said, \
    > “I’m lonely...”

    ```{r}
    numbers <- c("ten", "nine", "eight", "seven", "six", "five",
                 "four", "three", "two", "one")
    for (i in numbers) {
      cat(str_c("There were ", i, " in the bed\n"))
      cat("and the little one said\n")
      if (i == "one") {
        cat("I'm lonely...")
      } else {
        cat("Roll over, roll over\n")
        cat("So they all rolled over and one fell out.\n")
      }
      cat("\n")
    }
    ```

1.  The lyrics of [Ninety-Nine Bottles of Beer on the Wall](https://en.wikipedia.org/wiki/99_Bottles_of_Beer) are

    > 99 bottles of beer on the wall, 99 bottles of beer. \
    > Take one down, pass it around, 98 bottles of beer on the wall

    This verse is repeated, each time with one few bottle, until
    there are no more bottles of beer. The last verse is

    > No more bottles of beer on the wall, no more bottles of beer. \
    > We've taken them down and passed them around; now we're drunk and passed out!

    For the bottles of beer, I define a helper function to correctly print the number of bottles.

    ```{r}
    bottles <- function(i) {
      if (i > 2) {
       bottles <- str_c(i - 1, " bottles")
      } else if (i == 2) {
       bottles <- "1 bottle"
      } else {
       bottles <- "no more bottles"
      }
      bottles
    }

    beer_bottles <- function(n) {
      # should test whether n >= 1.
      for (i in seq(n, 1)) {
         cat(str_c(bottles(i), " of beer on the wall, ", bottles(i), " of beer.\n"))
         cat(str_c("Take one down and pass it around, ", bottles(i - 1),
                    " of beer on the wall.\n\n"))
      }
      cat("No more bottles of beer on the wall, no more bottles of beer.\n")
      cat(str_c("Go to the store and buy some more, ", bottles(n), " of beer on the wall.\n"))
    }
    beer_bottles(3)
    ```

</div>

#### Exercise <span class="exercise-number">21.2.1.4</span> {.unnumbered .exercise}

<div class="question">
It's common to see for loops that don't preallocate the output and instead increase the length of a vector at each step:

```{r, eval = FALSE}
output <- vector("integer", 0)
for (i in seq_along(x)) {
  output <- c(output, lengths(x[[i]]))
}
output
```
How does this affect performance? Design and execute an experiment.
</div>

<div class="answer">
I'll use the package **microbenchmark** to time this.
The `microbenchmark()` function will run an R expression a number of times and time it.

Define a function that appends to an integer vector.
```{r}
add_to_vector <- function(n) {
  output <- vector("integer", 0)
  for (i in seq_len(n)) {
    output <- c(output, i)
  }
  output  
}
microbenchmark(add_to_vector(10000), times = 3)
```

And one that pre-allocates it.
```{r}
add_to_vector_2 <- function(n) {
  output <- vector("integer", n)
  for (i in seq_len(n)) {
    output[[i]] <- i
  }
  output
}
microbenchmark(add_to_vector_2(10000), times = 3)
```

The pre-allocated vector is about **10** times faster!
You may get different answers, but the longer the vector and the bigger the objects, the more that pre-allocation will outperform appending.

</div>

## For loop variations

### Exercise <span class="exercise-number">21.3.5.1</span> {.unnumbered .exercise}

<div class="question">
Imagine you have a directory full of CSV files that you want to read in.
You have their paths in a vector,
`files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now
want to read each one with `read_csv()`. Write the for loop that will
load them into a single data frame.

</div>

<div class="answer">

First, I pre-allocate a list. Then, I read each file into a data frame, and assign it to an element in that list.
The result is a list of data frames.
I then use `bind_rows()` to combine the list of data frames into a single data frame.
```{r eval = FALSE}
df <- vector("list", length(files))
for (fname in seq_along(files)) {
  df[[i]] <- read_csv(files[[i]])
}
df <- bind_rows(df)
```

</div>

### Exercise <span class="exercise-number">21.3.5.2</span> {.unnumbered .exercise}

<div class="question">
What happens if you use `for (nm in names(x))` and `x` has no names?
What if only some of the elements are named?
What if the names are not unique?
</div>

<div class="answer">

Let's try it out and see what happens.

When there are no names for the vector, it does not run the code in the loop (it runs zero iterations of the loop):
```{r}
x <- 1:3
print(names(x))
for (nm in names(x)) {
  print(nm)
  print(x[[nm]])
}
```
Note that the length of `NULL` is zero:
```{r}
length(NULL)
```

If there only some names, then we get an error if we try to access an element without a name.
However, oddly, `nm == ""` when there is no name.
```{r}
x <- c(a = 1, 2, c = 3)
names(x)
```
```{r error = TRUE}
for (nm in names(x)) {
  print(nm)
  print(x[[nm]])
}
```

Finally, if there are duplicate names, then `x[[nm]]` will give the *first* element with that name.
There is no way to access elements with duplicate names.
```{r}
x <- c(a = 1, a = 2, c = 3)
names(x)
```
```{r}
for (nm in names(x)) {
  print(nm)
  print(x[[nm]])
}
```

</div>

### Exercise <span class="exercise-number">21.3.5.3</span> {.unnumbered .exercise}  

<div class="question">
Write a function that prints the mean of each numeric column in a data  frame, along with its name.
For example, `show_mean(iris)` would print:
</div>

<div class="answer">

```{r, eval = FALSE}
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width:  3.06
#> Petal.Length: 3.76
#> Petal.Width:  1.20
```

> (Extra challenge: what function did I use to make sure that the numbers lined up nicely, even though the variable names had different lengths?)

There may be other functions to do this, but I'll use `str_pad()`, and `str_length()` to ensure that the space given to the variable names is the same.
I messed around with the options to `format()` until I got two digits .
```{r}
show_mean <- function(df, digits = 2) {
  # Get max length of all variable names in the dataset
  maxstr <- max(str_length(names(df)))
  for (nm in names(df)) {
    if (is.numeric(df[[nm]])) {
      cat(str_c(str_pad(str_c(nm, ":"), maxstr + 1L, side = "right"),
                format(mean(df[[nm]]), digits = digits, nsmall = digits),
                sep = " "),
          "\n")
    }
  }
}
show_mean(iris)
```

</div>

### Exercise <span class="exercise-number">21.3.5.4</span> {.unnumbered .exercise}

<div class="question">
What does this code do? How does it work?
</div>

<div class="answer">

```{r}
trans <- list(
  disp = function(x) x * 0.0163871,
  am = function(x) {
    factor(x, labels = c("auto", "manual"))
  }
)
```
```{r eval=FALSE}
for (var in names(trans)) {
  mtcars[[var]] <- trans[[var]](mtcars[[var]])
}
```

This code mutates the `disp` and `am` columns:

-   `disp` is  multiplied by 0.0163871
-   `am` is replaced by a factor variable.

The code works by looping over a named list of functions.
It calls the named function in the list on the column of `mtcars` with the same name, and replaces the values of that column.

E.g. this is a function:
```{r eval=FALSE}
trans[["disp"]]
```
This applies the function to the column of `mtcars` with the same name
```{r eval=FALSE}
trans[["disp"]](mtcars[["disp"]])
```

</div>

## For loops vs. functionals

### Exercise <span class="exercise-number">21.4.1.1</span> {.unnumbered .exercise}

<div class="question">
Read the documentation for `apply()`. In the 2d case, what two for loops does it generalize.
</div>

<div class="answer">

It generalizes looping over the rows or columns of a matrix or data-frame.

</div>

### Exercise <span class="exercise-number">21.4.1.2</span> {.unnumbered .exercise}

<div class="question">
Adapt `col_summary()` so that it only applies to numeric columns.
You might want to start with an `is_numeric()` function that returns a logical vector that has a `TRUE` corresponding to each numeric column.
</div>

<div class="answer">

The original `col_summary()` function is,
```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
```

The adapted version is,
```{r}
col_summary2 <- function(df, fun) {
  # test whether each colum is numeric
  numeric_cols <- vector("logical", length(df))
  for (i in seq_along(df)) {
    numeric_cols[[i]] <- is.numeric(df[[i]])
  }
  # indexes of numeric columns
  idxs <- seq_along(df)[numeric_cols]
  # number of numeric columns
  n <- sum(numeric_cols)
  out <- vector("double", n)
  for (i in idxs) {
    out[i] <- fun(df[[i]])
  }
  out
}
```

Let's test that it works,
```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = letters[1:10],
  d = rnorm(10)
)
col_summary2(df, mean)
```

</div>

## The map functions

### Exercise <span class="exercise-number">21.5.3.1</span> {.unnumbered .exercise}

<div class="question">
Write code that uses one of the map functions to:

1.  Compute the mean of every column in `mtcars`.
1.  Determine the type of each column in `nycflights13::flights`.
1.  Compute the number of unique values in each column of `iris`.
1.  Generate 10 random normals for each of $\mu = -10$, $0$, $10$, and $100$.

</div>

<div class="answer">

1.  To calculate the mean of every column in `mtcars`, apply the function
    `mean()` to each column, and use `map_dbl`, since the results are numeric.
    ```{r}
    map_dbl(mtcars, mean)
    ```

1.  To calculate the type of every column in `nycflights13::flights` apply
    the function `typeof()`, discussed in the section on [Vector basics](http://r4ds.had.co.nz/vectors.html#vector-basics), 
    and use `map_chr()`, since the results are character.
    ```{r}
    map_chr(nycflights13::flights, typeof)
    ```
    

1.  There is no function that directly calculates the number of unique values 
    in a vector. For a single column, the number of unique values of a vector
    can be calculated like so,
    ```{r}
    length(unique(iris$Species))
    ```
    To apply this to all columns, we can provide the map funtion anonymous function.
    We can write anonymous function using the standard R syntax, like this,
    ```{r}
    map_int(iris, function(x) length(unique(x)))
    ```
    or using the more compact, one-sided formula shortcut that **purrr** provides,
    ```{r}
    map_int(iris, ~ length(unique(.)))
    ```
    The `map_int` function is used since `length()` returns an integer.
    However, `map_dbl` will also work,
    ```{r results='hide'}
    map_dbl(iris, ~ length(unique(.)))    
    ```

1.  To generate 10 random normals for each of $\mu = -10$, $0$, $10$, and $100$:
    The result is a list of numeric vectors.
    ```{r}
    map(c(-10, 0, 10, 100), ~ rnorm(n = 10, mean = .))
    ```
    Since a single call of `rnorm()` returns a numeric vector with a length greater
    than one we cannot use `map_dbl`, which requires the function to return a numeric
    vector that is only length one (see [Exercise 21.5.3.4](#exercise-21.5.3.4)
    The map functions pass any additional arguments to the function being called.

</div>

### Exercise <span class="exercise-number">21.5.3.2</span> {.unnumbered .exercise}  

<div class="question">
How can you create a single vector that for each column in a data frame indicates whether or not it's a factor?
</div>

<div class="answer">

The function `is.factor` indicates whether a vector is a factor. For example,
```{r}
is.factor(iris$Species)
is.factor(iris$Sepal.Length)
```
To apply this to all columns in a data frame, use `map_lgl` since the result
of `is.factor` is logical. For example,
```{r}
map_lgl(iris, is.factor)
```

</div>

### Exercise <span class="exercise-number">21.5.3.3</span> {.unnumbered .exercise}

<div class="question">
What happens when you use the map functions on vectors that aren't lists?
What does `map(1:5, runif)` do?
Why?
</div>

<div class="answer">

Map functions work with any vectors, not just lists. 
As with lists, the map functions will apply the function to each element of the vector.

The expression `map(1:5, runif)` is equivalent to running `runif(1)`, `runif(2)`,
..., `runif(5)`, and collecting the results in a list.
The result is a length five list with numeric vectors of sizes one through five.
```{r}
map(1:5, runif)
```


</div>

### Exercise <span class="exercise-number">21.5.3.4</span> {.unnumbered .exercise}

<div class="question">
What does `map(-2:2, rnorm, n = 5)` do? Why?

What does `map_dbl(-2:2, rnorm, n = 5)` do? Why?
</div>

<div class="answer">

Consider the first expression.
```{r}
map(-2:2, rnorm, n = 5)
```
This takes samples size five from five normal distributions, each with a different mean of (-2, -1, 0, 1, and 2) but all have the same standard deviation (1). It returns a list with each element a numeric vectors of length 5.

However, if we use `map_dbl` it throws an error. 
```{r error = TRUE}
map_dbl(-2:2, rnorm, n = 5)
```
This is because `map_dbl` expects the function it applies to each element to 
return a numeric vector of length one.

If we wanted a numeric vector, we could use `map()` followed by `flatten_dbl()`,
```{r}
flatten_dbl(map(-2:2, rnorm, n = 5))
```

</div>

### Exercise <span class="exercise-number">21.5.3.5</span> {.unnumbered .exercise}

<div class="question">
Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the anonymous function.
</div>

<div class="answer">

```{r}
map(list(mtcars), ~ lm(mpg ~ wt, data = .))
```

</div>

## Dealing with Failure

No exercises

## Mapping over multiple arguments

No exercises

## Walk

No exercises

## Other patterns of for loops

### Exercise <span class="exercise-number">21.9.3.1</span> {.unnumbered .exercise}

<div class="question">
Implement your own version of `every()` using a for loop.
Compare it with `purrr::every()`.
What does purrr's version do that your version doesn't?
</div>

<div class="answer">

```{r}
# Use ... to pass arguments to the function
every2 <- function(.x, .p, ...) {
  for (i in .x) {
    if (!.p(i, ...)) {
      # If any is FALSE we know not all of then were TRUE
      return(FALSE)
    }
  }
  # if nothing was FALSE, then it is TRUE
  TRUE  
}

every2(1:3, function(x) {x > 1})
every2(1:3, function(x) {x > 0})
```

The function `purrr::every()` does fancy things with `.p`, like taking a logical vector instead of a function, or being able to test part of a string if the elements of `.x` are lists.
</div>

### Exercise <span class="exercise-number">21.9.3.2</span> {.unnumbered .exercise}

<div class="question">
Create an enhanced `col_sum()` that applies a summary function to every numeric column in a data frame.
</div>

<div class="answer">

**Note:** this question has a typo. It is referring to `col_summary()`.

I will use `map` to apply the function to all the columns, and `keep` to only select numeric columns.
```{r}
col_sum2 <- function(df, f, ...) {
  map(keep(df, is.numeric), f, ...)
}
```

```{r}
col_sum2(iris, mean)
```

</div>

### Exercise <span class="exercise-number">21.9.3.3</span> {.unnumbered .exercise}

<div class="question">
Create possible base R equivalent of `col_sum()` is:

```{r}
col_sum3 <- function(df, f) {
  is_num <- sapply(df, is.numeric)
  df_num <- df[, is_num]
  sapply(df_num, f)
}
```

But it has a number of bugs as illustrated with the following inputs:

```{r, eval=FALSE}
df <- tibble(
  x = 1:3,
  y = 3:1,
  z = c("a", "b", "c")
)

# OK
col_sum3(df, mean)
# Has problems: don't always return numeric vector
col_sum3(df[1:2], mean)
col_sum3(df[1], mean)
col_sum3(df[0], mean)
```

What causes these bugs?
</div>

<div class="answer">
The problem is that `sapply` does not always return numeric vectors.
If no columns are selected, instead of returning an empty numeric vector, it returns an empty list.
This causes an error since we can't use a list with `[`.
```{r}
sapply(df[0], is.numeric)
```
```{r}
sapply(df[1], is.numeric)
```
```{r}
sapply(df[1:2], is.numeric)
```
</div>

<!-- match unopened div --><div>
