# Model Basics


## Prerequisites

```{r, message=FALSE}
library("tidyverse")
library("modelr")
options(na.action = na.warn)
```

The option `na.action` determines how missing values are handled.
It is a function.
`na.warn` sets it so that there is a warning if there are any missing values (by default, R will just silently drop them).

## A simple model


### Exercises

1. One downside of the linear model is that it is sensitive to unusual values because the distance incorporates a squared term. Fit a linear model to the simulated data below, and visualise the results. Rerun a few times to generate different simulated datasets. What do you notice about the model?

```{r}
sim1a <- tibble(
  x = rep(1:10, each = 3),
  y = x * 1.5 + 6 + rt(length(x), df = 2)
)
```

Let's run it once and plot the results:
```{r}
ggplot(sim1a, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

There appear to be a few outliers in this data.
You can re-rerun this a couple times to see if this hold.

We can also do this slightly more systematically. 
We will simulate this several times using `purrr` and plot the line using `geom_smooth`:
```{r}
simt <- function(i) {
  tibble(
    x = rep(1:10, each = 3),
    y = x * 1.5 + 6 + rt(length(x), df = 2),
    .id = i
  )
}

sims <- map_df(1:12, simt)

ggplot(sims, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~ .id, ncol = 4)

```

What if we did the same things with normal distributions? 
```{r}
sim_norm <- function(i) {
  tibble(
    x = rep(1:10, each = 3),
    y = x * 1.5 + 6 + rnorm(length(x)),
    .id = i
  )
}

simdf_norm <- map_df(1:12, sim_norm)

ggplot(simdf_norm, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~ .id, ncol = 4)

```
There are not large outliers, and the slopes are more similar. 

The reason for this is that the Student's $t$-distribution, from which we sample with `rt` has fatter tails than the normal distribution (`rnorm`), which means is assigns larger probability to values further from the center of the distribution.
```{r}
tibble(
  x = seq(-5, 5, length.out = 100),
  normal = dnorm(x),
  student_t = dt(x, df = 2) 
) %>% 
  gather(distribution, density, -x) %>%
  ggplot(aes(x = x, y = density, colour = distribution)) +
  geom_line()
```

For a normal distribution with mean zero and standard deviation one, the probability of being greater than 2 is,
```{r}
pnorm(2, lower.tail = FALSE)
```
For a Student's $t$ distribution with degrees of freedom = 2, it is more than 3 times higher,
```{r}
pt(2, df = 2, lower.tail = FALSE)
```


2. One way to make linear models more robust is to use a different distance measure. For example, instead of root-mean-squared distance, you could use mean-absolute distance:

```{r}
measure_distance <- function(mod, data) {
  diff <- data$y - make_prediction(mod, data)
  mean(abs(diff))
}
```

For the above function to work, we need to define a function `make_prediction` that
takes a numeric vector of length two (the intercept and slope) and returns the prections,
```{r}
make_prediction <- function(mod, data) {
  mod[1] + mod[2] * data$x
}
```

Using the `sim1a` data, the best parameters of the least absolute deviation are:
```{r}
best <- optim(c(0, 0), measure_distance, data = sim1a)
best$par
```
Using the `sim1a` data, while the parameters the minimize the least squares objective function are:
```{r}
measure_distance_ls <- function(mod, data) {
  diff <- data$y - (mod[1] + mod[2] * data$x)
  sqrt(mean(diff ^ 2))
}
best <- optim(c(0, 0), measure_distance_ls, data = sim1a)
best$par
```

In practice, you would not use a `optim` to fit this model, you would you an existing implementation.
See the `MASS` package's `rlm` and `lqs` functions for more information and functions to fit robust and resistant linear models.


3. One challenge with performing numerical optimization is that it’s only guaranteed to find a local optimum. What’s the problem with optimizing a three parameter model like this?

```{r eval=FALSE}
model3 <- function(a, data) {
  a[1] + data$x * a[2] + a[3]
}
```

The problem is that you for any values `a[1] = a1` and `a[3] = a3`, any other values of `a[1]` and `a[3]` where `a[1] + a[3] == (a1 + a3)` will have the same fit.

For example, we can find starting points 
```{r}
measure_distance_3 <- function(a, data) {
  diff <- data$y - model3(a, data)
  sqrt(mean(diff ^ 2))
}
```
Depending on our starting points, we can find different optimal values:
```{r}
best3a <- optim(c(0, 0, 0), measure_distance_3, data = sim1)
best3a$par
```
```{r}
best3b <- optim(c(0, 0, 1), measure_distance_3, data = sim1)
best3b$par
```
```{r}
best3c <- optim(c(0, 0, 5), measure_distance_3, data = sim1)
best3c$par
```
In fact there are an infinite number of optimal values for this model.

<!-- How to discuss this better ?

Problem is that due to finite iterations, numerically these converge:

> sum(best3a$par[c(1, 3)])
[1] 4.220074
> sum(best3b$par[c(1, 3)])
[1] 4.220404
> sum(best3c$par[c(1, 3)])
[1] 4.22117

-->


## Visualizing Models

### Exercises

1. Instead of using `lm()` to fit a straight line, you can use `loess()` to fit a smooth curve. Repeat the process of model fitting, grid generation, predictions, and visualization on `sim1` using `loess()` instead of `lm()`. How does the result compare to `geom_smooth()`?

I'll use `add_predictions` and `add_residuals` to add the predictions and residuals from a loess regression to the `sim1` data. 
```{r}
sim1_loess <- loess(y ~ x, data = sim1)
grid_loess <- sim1 %>%
  add_predictions(sim1_loess)

sim1 <- sim1 %>%
  add_residuals(sim1_loess, var = "resid_loess") %>%
  add_predictions(sim1_loess, var = "pred_loess")
  
```

This plots the loess predictions. The loess produces a non-linear, but smooth line through the data.
```{r}
plot_sim1_loess <- 
  ggplot(sim1, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(x = x, y = pred), data = grid_loess, colour = "red")
plot_sim1_loess
```

The predictions of loess are the same as the default method for `geom_smooth` because `geom_smooth()` uses `loess()` by default; the message even tells us that.
```{r}
plot_sim1_loess +
  geom_smooth(colour = "blue", se = FALSE, alpha = 0.20)
```

We can plot the residuals (red), and compare them to the residuals from `lm` (black). 
In general, the loess model has smaller residuals within the sample (out of sample is a different issue, and we haven't considered the uncertainty of these estimates).

```{r}
ggplot(sim1, aes(x = x)) +
  geom_ref_line(h = 0) +
  geom_point(aes(y = resid)) +
  geom_point(aes(y = resid_loess), colour = "red")

```

2. `add_predictions()` is paired with `gather_predictions()` and `spread_predictions()`. How do these three functions differ?

The functions `gather_predictions` and `spread_predictions` allow for adding predictions from multiple models at once.

Taking the `sim1_mod` example,
```{r}
sim1_mod <- lm(y ~ x, data = sim1)
grid <- sim1 %>% 
  data_grid(x) 
```

The function `add_predictions` adds only a single model at a time.
To add two models:
```{r}
grid %>%
  add_predictions(sim1_mod, var = "pred_lm") %>%
  add_predictions(sim1_loess, var = "pred_loess")
```
The function `gather_predictions` adds predictions from multiple models by
stacking the results and adding a column with the model name,
```{r}
grid %>%
  gather_predictions(sim1_mod, sim1_loess) 
```
The function `spread_predictions` adds predictions from multiple models by
adding multiple columns (postfixed with the model name) with predictions from each model.
```{r}
grid %>%
  spread_predictions(sim1_mod, sim1_loess)
```
The function `spread_predictions` is similar to the example which runs `add_predictions` for each model, and is equivalent to running `spread` after
running `gather_predictions`:
```{r}
grid %>%
  gather_predictions(sim1_mod, sim1_loess) %>%
  spread(model, pred)
```



3. What does `geom_ref_line()` do? What package does it come from? Why is displaying a reference line in plots showing residuals useful and important?

The geom `geom_ref_line()` adds as reference line to a plot.
It is equivalent to running `geom_hline` or `geom_vline` with default settings that are useful for visualizing models.

Even though it alters a **ggplot2** plot, it is in the **modelr** package.
Putting a reference line at zero for residuals is important because good models (generally) should have residuals centered at zero, with approximately the same variance (or distribution) over the support of x, and no correlation.
A zero reference line makes it easier to judge these characteristics visually.

4. Why might you want to look at a frequency polygon of absolute residuals? What are the pros and cons compared to looking at the raw residuals?

The frequency polygon makes it easier to judge whether the variance and/or absolute size of the residuals varies with respect to x.
This pattern of residuals is called heteroskedasticity, and results in incorrect (biased) standard errors in inference.
In prediction, this provides insight into where the model is working well and where it is not.
What is lost, is that since the absolute values are shown, whether the model is over-predicting or underpredicting, or on average correctly predicting in different regions of x cannot be determined.

## Formulas and Model Families


### Categorical Variables


### Exercises

1. What happens if you repeat the analysis of `sim2` using a model without an intercept. What happens to the model equation? What happens to the predictions?

To run a model without an intercept, add `- 1` or `+ 0` to the right-hand-side o f the formula:
```{r}
mod2a <- lm(y ~ x - 1, data = sim2)
```
```{r}
mod2 <- lm(y ~ x, data = sim2)
```

The predictions are exactly the same in the models with and without an intercept:
```{r}
grid <- sim2 %>% 
  data_grid(x) %>% 
  add_predictions(mod2, mod2a)
grid
```



2. Use `model_matrix()` to explore the equations generated for the models I fit to `sim3` and `sim4`. Why is `*` a good shorthand for interaction?





3. Using the basic principles, convert the formulas in the following two models into functions. (Hint: start by converting the categorical variable into 0-1 variables.)

```{r}
mod1 <- lm(y ~ x1 + x2, data = sim3)
mod2 <- lm(y ~ x1 * x2, data = sim3)
```

**TODO**

4. For `sim4`, which of `mod1` and `mod2` is better? I think `mod2` does a slightly better job at removing patterns, but it’s pretty subtle. Can you come up with a plot to support my claim?

**TODO**

## Missing values

## Other model families

No exercises

