# Model Building

## Introduction

```{r}
library(tidyverse)
library(modelr)
options(na.action = na.warn)
library("broom")

library(nycflights13)
library(lubridate)
```

**Notes**

- talks about `MASS::rlm()`
- splines


## Why are low quality diamonds more expensive?



```{r}
diamonds2 <- diamonds %>% 
  filter(carat <= 2.5) %>% 
  mutate(lprice = log2(price), lcarat = log2(carat))

mod_diamond2 <- lm(lprice ~ lcarat + color + cut + clarity, data = diamonds2)
```



### Exercises

1. In the plot of `lcarat` vs. `lprice`, there are some bright vertical strips. What do they represent?

The distribution of diamonds has more diamonds at round or otherwise human friendly numbers (fractions).

2. If `log(price) = a_0 + a_1 * log(carat)`, what does that say about the relationship between `price` and `carat`?

An 1% increase in carat is associated with an $a_1$% increase in price.

3. Extract the diamonds that have very high and very low residuals. Is there anything unusual about these diamonds? Are the particularly bad or good, or do you think these are pricing errors?

This was already covered in the text. I don't see anything either.

4. Does the final model, `mod_diamonds2`, do a good job of predicting diamond prices? Would you trust it to tell you how much to spend if you were buying a diamond?


```{r}
diamonds2 %>% 
  add_predictions(mod_diamond2) %>%
  add_residuals(mod_diamond2) %>%
  summarise(sq_err = sqrt(mean(resid^2)),
            abs_err = mean(abs(resid)),
            p975_err = quantile(resid, 0.975),
            p025_err = quantile(resid, 0.025))
            

```

The average squared and absolute erorrs are $2^0.19 = 1.14$ and $2^0.10$ so on average, the error is $\pm 10--15$%. 
And the 95% range of residuals is about $2^0.37 = 1.3$ so within $\pm 30$%.
This doesn't seem terrible to me.

### What affects the number of daily flights?

```{r}
library("nycflights13")
daily <- flights %>% 
  mutate(date = make_date(year, month, day)) %>% 
  group_by(date) %>% 
  summarise(n = n())
daily
```

```{r}
daily <- daily %>% 
  mutate(wday = wday(date, label = TRUE))
ggplot(daily, aes(wday, n)) + 
  geom_boxplot()
```

```{r}
mod <- lm(n ~ wday, data = daily)

grid <- daily %>% 
  data_grid(wday) %>% 
  add_predictions(mod, "n")

ggplot(daily, aes(wday, n)) + 
  geom_boxplot() +
  geom_point(data = grid, colour = "red", size = 4)
```

```{r}
daily <- daily %>% 
  add_residuals(mod)
daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

```{r}
ggplot(daily, aes(date, resid, colour = wday)) + 
  geom_ref_line(h = 0) + 
  geom_line()
```

```{r}
daily %>% 
  filter(resid < -100)
```

```{r}
daily %>% 
  ggplot(aes(date, resid)) + 
  geom_ref_line(h = 0) + 
  geom_line(colour = "grey50") + 
  geom_smooth(se = FALSE, span = 0.20)
```

```{r}
term <- function(date) {
  cut(date, 
    breaks = ymd(20130101, 20130605, 20130825, 20140101),
    labels = c("spring", "summer", "fall") 
  )
}

daily <- daily %>% 
  mutate(term = term(date)) 

daily %>% 
  filter(wday == "Sat") %>% 
  ggplot(aes(date, n, colour = term)) +
  geom_point(alpha = 1/3) + 
  geom_line() +
  scale_x_date(NULL, date_breaks = "1 month", date_labels = "%b")
```

```{r}
daily %>% 
  ggplot(aes(wday, n, colour = term)) +
    geom_boxplot()
```

```{r}
mod1 <- lm(n ~ wday, data = daily)
mod2 <- lm(n ~ wday * term, data = daily)

daily %>% 
  gather_residuals(without_term = mod1, with_term = mod2) %>% 
  ggplot(aes(date, resid, colour = model)) +
    geom_line(alpha = 0.75)
```

```{r}
grid <- daily %>% 
  data_grid(wday, term) %>% 
  add_predictions(mod2, "n")

ggplot(daily, aes(wday, n)) +
  geom_boxplot() + 
  geom_point(data = grid, colour = "red") + 
  facet_wrap(~ term)
```

```{r}
mod3 <- MASS::rlm(n ~ wday * term, data = daily)

daily %>% 
  add_residuals(mod3, "resid") %>% 
  ggplot(aes(date, resid)) + 
  geom_hline(yintercept = 0, size = 2, colour = "white") + 
  geom_line()
```

```{r}
compute_vars <- function(data) {
  data %>% 
    mutate(
      term = term(date), 
      wday = wday(date, label = TRUE)
    )
}
```

- Doesn't discuss why and how chooses spline of 5 knots.
- Introduces natural splines

```{r}
library(splines)
mod <- MASS::rlm(n ~ wday * ns(date, 5), data = daily)

daily %>% 
  data_grid(wday, date = seq_range(date, n = 13)) %>% 
  add_predictions(mod) %>% 
  ggplot(aes(date, pred, colour = wday)) + 
    geom_line() +
    geom_point()
```

```{r}
library("nycflights13")
daily <- flights %>% 
  mutate(date = make_date(year, month, day)) %>% 
  group_by(date) %>% 
  summarise(n = n())
daily

daily <- daily %>% 
  mutate(wday = wday(date, label = TRUE))

term <- function(date) {
  cut(date, 
    breaks = ymd(20130101, 20130605, 20130825, 20140101),
    labels = c("spring", "summer", "fall") 
  )
}

daily <- daily %>% 
  mutate(term = term(date)) 

mod <- lm(n ~ wday, data = daily)

daily <- daily %>% 
  add_residuals(mod)

mod1 <- lm(n ~ wday, data = daily)
mod2 <- lm(n ~ wday * term, data = daily)

```

### Exercises

**Ex 1.** Use your Google sleuthing skills to brainstorm why there were fewer than expected flights on Jan 20, May 26, and Sep 1. (Hint: they all have the same explanation.) How would these days generalise to another year?

These are the Sundays before Monday holidays Martin Luther King Day, Memorial Day, and Labor Day.

**Ex 2.** 

```{r}
daily %>%
  top_n(3, resid)
```


**Ex 3.** Create a new variable that splits the wday variable into terms, but only for Saturdays, i.e. it should have `Thurs`, `Fri`, but `Sat-summer`, `Sat-spring`, `Sat-fall` How does this model compare with the model with every combination of wday and term?

I'll use the function `case_when` to do this, though there are other ways which it could be solved.
```{r}
daily <- daily %>%
  mutate(wday2 = 
         case_when(.$wday == "Sat" & .$term == "summer" ~ "Sat-summer",
         .$wday == "Sat" & .$ term == "fall" ~ "Sat-fall",
         .$wday == "Sat" & .$term == "spring" ~ "Sat-spring",
         TRUE ~ as.character(.$wday)))
```



```{r}
mod4 <- lm(n ~ wday2, data = daily)

daily %>% 
  gather_residuals(sat_term = mod4, all_interact = mod2) %>% 
  ggplot(aes(date, resid, colour = model)) +
    geom_line(alpha = 0.75)
```

I think the overlapping plot is hard to understand.
If we are interested in the differences, it is better to plot the differences directly.
In this code I use `spread_residuals` to add one *column* per model, rather than `gather_residuals` which creates a new row for each model.
```{r}
daily %>% 
  spread_residuals(sat_term = mod4, all_interact = mod2) %>%
  mutate(resid_diff = sat_term - all_interact) %>%
  ggplot(aes(date, resid_diff)) +
    geom_line(alpha = 0.75)
```

The model with terms x Saturday has higher residuals in the fall, and lower residuals in the spring than the model with all interactions. 

Using overall model comparison terms, `mod4` has a lower $R^2$ and regression standard error, $\hat\sigma$, despite using fewer variables.
More importantly for prediction purposes, it has a higher AIC - which is an estimate of the out of sample error.
```{r}
glance(mod4) %>% select(r.squared, sigma, AIC, df)
```
```{r}
glance(mod2) %>% select(r.squared, sigma, AIC, df)
```

**Ex 4.** TODO 


**Ex. 5**  TODO


**Ex 5.** What happens if you fit a day of week effect that varies by month (i.e. n ~ wday * month)? Why is this not very helpful?

There are only 4-5 observations per parameter since only there are only 4-5 weekdays in a given month.


**Ex 6.** It will estimate a smooth seasonal trend (`ns(date, 5)`) with a day of the week cylicality, (`wday`).
It probably will not be effective since 

**Ex 7.** We hypothesised that people leaving on Sundays are more likely to be business travellers who need to be somewhere on Monday. Explore that hypothesis by seeing how it breaks down based on distance and time: if it’s true, you’d expect to see more Sunday evening flights to places that are far away.

Looking at only day of the week, we see that Sunday flights are on average longer than the rest of the day of the week flights, but not as long as Saturday flights (perhaps vacation flights?).
```{r}
flights %>% 
  mutate(date = make_date(year, month, day),
         wday = wday(date, label = TRUE)) %>%
  group_by(wday) %>%
  summarise(dist_mean =  mean(distance),
            dist_median = median(distance)) %>%
  ggplot(aes(y = dist_mean, x = wday)) +
  geom_point()

```

However, breaking it down by hour, I don't see much evidence at first.
Conditional on hour, the distance of Sunday flights seems similar to that of other days (excluding Saturday):

```{r}
flights %>% 
  mutate(date = make_date(year, month, day),
         wday = wday(date, label = TRUE)) %>%
  group_by(wday, hour) %>%
  summarise(dist_mean =  mean(distance),
            dist_median = median(distance)) %>%
  ggplot(aes(y = dist_mean, x = hour, colour = wday)) +
  geom_point() + 
  geom_line()
```

Can someone think of a better way to check this?

**Ex 8.** 

**Ex 8. It’s a little frustrating that Sunday and Saturday are on separate ends of the plot. Write a small function to set the levels of the factor so that the week starts on Monday.

```{r}


```

