# Relational Data

## Introduction

```{r message=FALSE}
library("tidyverse")
library("nycflights13")
```

The package datamodelr is used to draw database schema:
```{r}
library("datamodelr")
```

### Exercise 1 {.exercise}

`r BeginQuestion()`
Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?
`r EndQuestion()`

`r BeginAnswer()`

-   `flights` table: `origin` and `dest`
-   `airports` table: longitude and latitude variables
-   join `flights` with `airports` twice. The first join adds the location of the origin airport (`origin`). The second join adds the location of destination airport (`dest`).

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?
`r EndQuestion()`

`r BeginAnswer()`

The variable `origin` in `weather` is matched with `faa` in `airports`.

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
weather only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with `flights`?
`r EndQuestion()`

`r BeginAnswer()`

`year`, `month`, `day`, `hour`, `origin` in `weather` would be matched to `year`, `month`, `day`, `hour`, `dest` in `flight` (though it should use the arrival date-time values for `dest` if possible).

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?
`r EndQuestion()`

`r BeginAnswer()`

I would add a table of special dates.
The primary key would be date.
It would match to the `year`, `month`, `day` columns of `flights.

`r EndAnswer()`

## Keys

### Exercise 1 {.exercise}

`r BeginQuestion()`
Add a surrogate key to flights.
`r EndQuestion()`

`r BeginAnswer()`

I add the column `flight_id` as a surrogate key.
I sort the data prior to making the key, even though it is not strictly necessary, so the order of the rows has some meaning.
```{r}
flights %>%
  arrange(year, month, day, sched_dep_time, carrier, flight) %>%
  mutate(flight_id = row_number()) %>%
  glimpse()
```

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Identify the keys in the following datasets

1.  `Lahman::Batting`
1.  `babynames::babynames`
1.  `nasaweather::atmos`
1.  `fueleconomy::vehicles`
1.  `ggplot2::diamonds`

(You might need to install some packages and read some documentation.)

`r EndQuestion()`

`r BeginAnswer()`

The primary key for `Lahman::Batting` is `playerID`, `yearID`, `stint`. It is not simply `playerID`, `yearID` because players can have different stints in different leagues within the same year.
```{r}
Lahman::Batting %>%
  group_by(playerID, yearID, stint) %>%
  filter(n() > 1) %>%
  nrow()
```

The primary key for `babynames::babynames` is `year`, `sex`, `name`. It is not simply `year`, `name` since names can appear for both sexes with different counts.
```{r}
babynames::babynames %>%
  group_by(year, sex, name) %>%
  filter(n() > 1) %>%
  nrow()
```

The primary key for `nasaweather::atmos` is the location and time of the measurement: `lat`, `long`, `year`, `month`.
```{r}
nasaweather::atmos %>%
  group_by(lat, long, year, month) %>%
  filter(n() > 1) %>%
  nrow()
```

The column `id` (unique EPA identifier) is the primary key for `fueleconomy::vehicles`:
```{r}
fueleconomy::vehicles %>%
  group_by(id) %>%
  filter(n() > 1) %>%
  nrow()
```

There is no primary key for `ggplot2::diamonds`. Using all variables in the data frame, the number of distinct rows is less than the total number of rows, meaning no combination of variables uniquely identifies the observations.
```{r}
ggplot2::diamonds %>%
  distinct() %>%
  nrow()
nrow(ggplot2::diamonds)
```

`r EndAnswer()`

### Exercise 3 {.exercise}

Draw a diagram illustrating the connections between the `Batting`, `Master`, and `Salaries` tables in the **Lahman** package. Draw another diagram that shows the relationship between `Master`, `Managers`, `AwardsManagers`.

Most flowchart or diagramming software can be used used to create database schema diagrams.
For example, the diagrams in *R for Data Science* were created with [Gliffy](https://www.gliffy.com/).

You can use anything to create these diagrams, but I'll use the R package [datamodelr](https://github.com/bergant/datamodelr) to programmatically create data models from R.

For the `Batting`, `Master`, and `Salaries` tables:

-   `Master`

    -   Primary keys: `playerID`

-   `Batting`

    -   Primary keys: `yearID`, `yearID`, `stint`

    -   Foreign Keys:

        -   `playerID` = `Master$playerID` (many-to-1)

-   `Salaries`

    -   Primary keys: `yearID`, `teamID`, `playerID`

    -   Foreign Keys

        -   `playerID` = `Master$playerID` (many-to-1)

```{r}
dm1 <- dm_from_data_frames(list(Batting = Lahman::Batting,
                                Master = Lahman::Master,
                                Salaries = Lahman::Salaries)) %>%
  dm_set_key("Batting", c("playerID", "yearID", "stint")) %>%
  dm_set_key("Master", "playerID") %>%
  dm_set_key("Salaries", c("yearID", "teamID", "playerID")) %>%
  dm_add_references(
    Batting$playerID == Master$playerID,
    Salaries$playerID == Master$playerID
  )

dm_create_graph(dm1, rankdir = "LR", columnArrows = TRUE)
```

For the `Master`, `Manager`, and `AwardsManagers` tables:

-   `Master`

    -   Primary keys: `playerID`

-   `Managers`

    -   Primary keys: `yearID`, `teamID`, `inseason`

    -   Foreign Keys:

        -   `playerID` = `Master$playerID` (many-to-1)

-   `AwardsManagers`:

    -   `playerID` = `Master$playerID` (many-to-1)

```{r}
dm2 <- dm_from_data_frames(list(Master = Lahman::Master,
                                Managers = Lahman::Managers,
                                AwardsManagers = Lahman::AwardsManagers)) %>%
  dm_set_key("Master", "playerID") %>%
  dm_set_key("Managers", c("yearID", "teamID", "inseason")) %>%
  dm_set_key("AwardsManagers", c("playerID", "awardID", "yearID")) %>%
  dm_add_references(
    Managers$playerID == Master$playerID,
    AwardsManagers$playerID == Master$playerID
  )

dm_create_graph(dm2, rankdir = "LR", columnArrows = TRUE)
```

In the previous diagrams, I do not consider `teamID` and `lgID` as foreign keys even though they appear in multiple tables (and have the same meaning) because they are not primary keys in the tables considered in this exercise.
The `teamID` variable references `Teams$teamID`, and `lgID` does not have its own table.

How would you characterize the relationship between the `Batting`, `Pitching`, and `Fielding` tables?

The `Batting`, `Pitching`, and `Fielding` tables all have a primary key consisting of the `playerID`, `yearID`, and `stint` variables.
They all have a 1-1 relationship to each other.

## Mutating Joins

```{r}
flights2 <- flights %>%
  select(year:day, hour, origin, dest, tailnum, carrier)
```

### Exercise 1 {.exercise}

`r BeginQuestion()`
Compute the average delay by destination, then join on the `airports` data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:
`r EndQuestion()`

`r BeginAnswer()`

```{r}
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

(Don’t worry if you don’t understand what `semi_join()` does — you’ll learn about it next.)

```{r}
avg_dest_delays <-
  flights %>%
  group_by(dest) %>%
  # arrival delay NA's are cancelled flights
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c(dest = "faa"))

avg_dest_delays %>%
  ggplot(aes(lon, lat, colour = delay)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

You might want to use the size or color of the points to display the average delay for each airport.

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Add the location of the origin and destination (i.e. the `lat` and `lon`) to `flights`.
`r EndQuestion()`

`r BeginAnswer()`

You can perform one join after another. If duplicate variables are found, by default, dplyr will distinguish the two by adding `.x`, and `.y` to the ends of the variable names to solve naming conflicts. 
```{r}
airport_locations <- airports %>%
  select(faa, lat, lon)
  
flights %>%
    select(year:day, hour, origin, dest) %>%
  left_join(
    airport_locations,
    by = c("origin" = "faa")
  ) %>%
  left_join(
    airport_locations,
    by = c("dest" = "faa")
  )
```
This default can be over-ridden using the `suffix` argument.
```{r}
airport_locations <- airports %>%
  select(faa, lat, lon)
  
flights %>%
    select(year:day, hour, origin, dest) %>%
  left_join(
    airport_locations,
    by = c("origin" = "faa")
  ) %>%
  left_join(
    airport_locations,
    by = c("dest" = "faa"),
    suffix = c("_origin", "_dest")
    #existing lat and lon variables in tibble gain the _origin suffix
    #new lat and lon variables are given _dest suffix
  )
```
It's always good practice to have clear variable names.

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
Is there a relationship between the age of a plane and its delays?
`r EndQuestion()`

`r BeginAnswer()`

Surprisingly not. If anything (departure) delay seems to decrease slightly with the age of the plane.
This could be due to choices about how airlines allocate planes to airports.
```{r}
plane_ages <-
  planes %>%
  mutate(age = 2013 - year) %>%
  select(tailnum, age)

flights %>%
  inner_join(plane_ages, by = "tailnum") %>%
  group_by(age) %>%
  filter(!is.na(dep_delay)) %>%
  summarise(delay = mean(dep_delay)) %>%
  ggplot(aes(x = age, y = delay)) +
  geom_point() +
  geom_line()

```

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
What weather conditions make it more likely to see a delay?
`r EndQuestion()`

`r BeginAnswer()`

Almost any amount or precipitation is associated with a delay, though not as strong a trend after 0.02 in as one would expect
```{r}
flight_weather <-
  flights %>%
  inner_join(weather, by = c("origin" = "origin",
                            "year" = "year",
                            "month" = "month",
                            "day" = "day",
                            "hour" = "hour"))

flight_weather %>%
  group_by(precip) %>%
  summarise(delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = precip, y = delay)) +
    geom_line() + geom_point()
```

`r EndAnswer()`

### Exercise 5 {.exercise}

`r BeginQuestion()`
What happened on June 13 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.
`r EndQuestion()`

`r BeginAnswer()`

There was a large series of storms (derechos) in the southeastern US (see [June 12-13, 2013 derecho series](https://en.wikipedia.org/wiki/June_12%E2%80%9313,_2013_derecho_series))

The largest delays are in Tennessee (Nashville), the Southeast, and the Midwest, which were the locations of the derechos:
```{r}
library(viridis)
flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c("dest" = "faa")) %>%
  ggplot(aes(y = lat, x = lon, size = delay, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap() +
  scale_colour_viridis()

```

`r EndAnswer()`

## Filtering Joins

### Exercise 1 {.exercise}

`r BeginQuestion()`
What does it mean for a flight to have a missing `tailnum`? What do the tail numbers that don’t have a matching record in planes have in common? (Hint: one variable explains ~90% of the problems.)
`r EndQuestion()`

`r BeginAnswer()`

American Airlines (AA) and Envoy Airlines (MQ) don't report tail numbers.
```{r}
flights %>%
  anti_join(planes, by = "tailnum") %>%
  count(carrier, sort = TRUE)

```

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Filter flights to only show flights with planes that have flown at least 100 flights.
`r EndQuestion()`

`r BeginAnswer()`

```{r}
planes_gt100 <-
  filter(flights) %>%
  group_by(tailnum) %>%
  count() %>%
  filter(n > 100)

flights %>%
  semi_join(planes_gt100, by = "tailnum")

```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
Combine `fueleconomy::vehicles` and `fueleconomy::common` to find only the records for the most common models.
`r EndQuestion()`

`r BeginAnswer()`

The table `fueleconomy::common` identifies vehicles by `make` and `model`:
```{r}
glimpse(fueleconomy::vehicles)
glimpse(fueleconomy::common)
```

```{r}
fueleconomy::vehicles %>%
  semi_join(fueleconomy::common, by = c("make", "model"))
```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the weather data. Can you see any patterns?
`r EndQuestion()`

`r BeginAnswer()`

```{r}
flights %>%
  group_by(year, month, day) %>%
  summarise(total_24 = sum(dep_delay, na.rm = TRUE)+ sum(arr_delay, na.rm = TRUE)) %>%
  mutate(total_48 = total_24 + lag(total_24)) %>%
  arrange(desc(total_48))
```

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
What does `anti_join(flights, airports, by = c("dest" = "faa"))` tell you? What does `anti_join(airports, flights, by = c("faa" = "dest"))` tell you?
`r EndQuestion()`

`r BeginAnswer()`

`anti_join(flights, airports, by = c("dest" = "faa"))` are flights that go to an airport that is not in FAA list of destinations, likely foreign airports.

`anti_join(airports, flights, by = c("faa" = "dest"))` are US airports that don't have a flight in the data, meaning that there were no flights to that airport **from** New York in 2013.

`r EndAnswer()`

### Exercise 5 {.exercise}

`r BeginQuestion()`
You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.
`r EndQuestion()`

`r BeginAnswer()`

There isn't such a relationship over the lifetime of an airplane since planes can be sold or leased and airlines can merge.
However, even though that's a possibility, it doesn't necessarily mean that plane associated with more than one  appear in this data.
Let's check:
```{r mulit_carrier_planes}
multi_carrier_planes <-
  flights %>%
  filter(!is.na(tailnum)) %>%
  count(tailnum, carrier) %>%
  count(tailnum) %>%
  filter(nn > 1)
multi_carrier_planes
```
There are `r nrow(multi_carrier_planes)` airplanes in this dataset that have had more than one carrier.

To see which carriers these planes have been associated, filter the `flights` by `tailnum` in `multi_carrier_planes`, and extract the unique combinations of `tailnum` and `carrier`.
```{r multi_carrier_planes_with_carrier}
multi_carrier_planes <-
  flights %>%
  semi_join(multi_carrier_planes, by = "tailnum") %>%
  select(tailnum, carrier) %>%
  distinct() %>%
  arrange(tailnum)
multi_carrier_planes
```

The names of airlines are easier to understand than the two-letter carrier codes.
Join the multi-airline table with the associated airline in `airlines` using the `carrier` column.
The spread the data so it has columns `carrier_1`, `carrier_2`, and so on.
This is not tidy, but it is more easier to display.
```{r carrier_transfer_tbl}
carrier_transfer_tbl <-
  multi_carrier_planes %>%
  group_by(tailnum) %>%
  mutate(
    carrier_num = seq_along(tailnum),
    carrier_num = paste0("carrier_", carrier_num)
  ) %>%
  left_join(airlines, by = "carrier") %>%
  select(-carrier) %>%
  spread(carrier_num, name)
carrier_transfer_tbl
```

`r EndAnswer()`

## Join problems

No exercises

## Set operations

No exercises
