# Relational data

## Introduction

```{r message=FALSE}
library("tidyverse")
library("nycflights13")
```

The package datamodelr is used to draw database schema:
```{r}
library("datamodelr")
```

## nycflights13

### Exercise <span class="exercise-number">13.2.1.1</span> {.unnumbered .exercise}

<div class="question">
Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?
</div>

<div class="answer">

-   `flights` table: `origin` and `dest`
-   `airports` table: longitude and latitude variables
-   join `flights` with `airports` twice. The first join adds the location of the origin airport (`origin`). The second join adds the location of destination airport (`dest`).

</div>

### Exercise <span class="exercise-number">13.2.1.2</span> {.unnumbered .exercise}

<div class="question">
I forgot to draw the relationship between weather and airports. What is the relationship and how should it appear in the diagram?
</div>

<div class="answer">

The column `faa$airports` is the foreign key of `weather$origin`.

</div>

### Exercise <span class="exercise-number">13.2.1.3</span> {.unnumbered .exercise}

<div class="question">
Weather only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with `flights`?
</div>

<div class="answer">

If the weather was included for all airports in the US, then it would provide the weather for the destination of each flight.
The `weather` data frame columns (`year`, `month`, `day`, `hour`, `origin`) are a foreign key for the `flights` data frame columns (`year`, `month`, `day`, `hour`, `dest`).
This would provide information about the weather at the destination airport at the time of the flight take off, unless the arrival date-time were calculated.

So why was this not a relationship prior to adding additional rows to the `weather` table?
In a foreign key relationship, the collection columns of columns in the child table 
must refer to a unique collection of columns in the parent table. 
When the `weather` table only contained New York airports,
there were many values of  (`year`, `month`, `day`, `hour`, `dest`) in `flights` that
did not appear in the `weather` table. Therefore, it was not a foreign key. It was only after
all combinations of year, month, day, hour, and airports that are defined in `flights`
were added to the `weather` table that there existed this relation between these tables.

</div>

### Exercise <span class="exercise-number">13.2.1.4</span> {.unnumbered .exercise}

<div class="question">
We know that some days of the year are “special”, and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?
</div>

<div class="answer">

I would add a table of special dates.
Its primary key would be  `date`.
This would match to the `year`, `month`,
and `day` columns of `flights`.

The table would resemble the following:
```{r}
special_days <- tribble(
  ~year, ~month, ~day, ~holiday,
  2013, 01, 01, "New Years Day",
  2013, 07, 04, "Independence Day",
  2013, 11, 29, "Thanksgiving Day",
  2013, 12, 25, "Christmas Day"
)
```

</div>

## Keys

### Exercise <span class="exercise-number">13.3.1.1</span> {.unnumbered .exercise}

<div class="question">
Add a surrogate key to flights.
</div>

<div class="answer">

I add the column `flight_id` as a surrogate key.
I sort the data prior to making the key, even though it is not strictly necessary, so the order of the rows has some meaning.
```{r}
flights %>%
  arrange(year, month, day, sched_dep_time, carrier, flight) %>%
  mutate(flight_id = row_number()) %>%
  glimpse()
```

</div>

### Exercise <span class="exercise-number">13.3.1.2</span> {.unnumbered .exercise}

<div class="question">
Identify the keys in the following datasets

1.  `Lahman::Batting`
1.  `babynames::babynames`
1.  `nasaweather::atmos`
1.  `fueleconomy::vehicles`
1.  `ggplot2::diamonds`

(You might need to install some packages and read some documentation.)

</div>

<div class="answer">

The answer to each part follows.

1.  The primary key for `Lahman::Batting` is `playerID`, `yearID`, `stint`.
    It is not simply `playerID`, `yearID` because players can have different stints in different leagues within the same year.

    ```{r}
    Lahman::Batting %>%
      count(playerID, yearID, stint) %>%
      filter(n > 1) %>%
      nrow()
    ```

1.  The primary key for `babynames::babynames` is `year`, `sex`, `name`.
    It is not simply `year`, `name` since names can appear for both sexes with different counts.

    ```{r}
    babynames::babynames %>%
      count(year, sex, name) %>%
      filter(nn > 1) %>%
      nrow()
    ```
    You may notice that in this example, we filter on `nn` instead of `n`.
    In this example, the name of the column generated by `count()` is named `nn` 
    because the `babynames` data frame already contains a column named `n`.
    
    This behavior requires knowing the names of the columns which `count()` 
    is being called on. The following code uses `group_by()` and `n()` to 
    produces the same result.
    ```{r}
    babynames::babynames %>%
      group_by(year, sex, name) %>%
      filter(n() > 1) %>%
      nrow()
    ```    

1.  The primary key for `nasaweather::atmos` is the location and time of the measurement: `lat`, `long`, `year`, `month`.

    ```{r}
    nasaweather::atmos %>%
      count(lat, long, year, month) %>%
      filter(n > 1) %>%
      nrow()
    ```

1.  The column `id` (unique EPA identifier) is the primary key for `fueleconomy::vehicles`:

    ```{r}
    fueleconomy::vehicles %>%
      count(id) %>%
      filter(n > 1) %>%
      nrow()
    ```

1.  There is no primary key for `ggplot2::diamonds`.
    The number of distinct rows in the dataset is less than the total number of rows, which implies that there is no combination of variables uniquely identifies the observations.

    ```{r}
    ggplot2::diamonds %>%
      distinct() %>%
      nrow()
    nrow(ggplot2::diamonds)
    ```

</div>

### Exercise <span class="exercise-number">13.3.1.3</span> {.unnumbered .exercise}

Draw a diagram illustrating the connections between the `Batting`, `Master`, and `Salaries` tables in the **Lahman** package. Draw another diagram that shows the relationship between `Master`, `Managers`, `AwardsManagers`.

Most flowchart or diagramming software can be used used to create database schema diagrams.
For example, the diagrams in *R for Data Science* were created with [Gliffy](https://www.gliffy.com/).

You can use anything to create these diagrams, but I'll use the R package [datamodelr](https://github.com/bergant/datamodelr) to programmatically create data models from R.

For the `Batting`, `Master`, and `Salaries` tables:

-   `Master`

    -   Primary keys: `playerID`

-   `Batting`

    -   Primary keys: `playerID`, `yearID`, `stint`

    -   Foreign Keys:

        -   `playerID` = `Master$playerID` (many-to-1)

-   `Salaries`

    -   Primary keys: `yearID`, `teamID`, `playerID`

    -   Foreign Keys

        -   `playerID` = `Master$playerID` (many-to-1)

```{r}
dm1 <- dm_from_data_frames(list(Batting = Lahman::Batting,
                                Master = Lahman::Master,
                                Salaries = Lahman::Salaries)) %>%
  dm_set_key("Batting", c("playerID", "yearID", "stint")) %>%
  dm_set_key("Master", "playerID") %>%
  dm_set_key("Salaries", c("yearID", "teamID", "playerID")) %>%
  dm_add_references(
    Batting$playerID == Master$playerID,
    Salaries$playerID == Master$playerID
  )

dm_create_graph(dm1, rankdir = "LR", columnArrows = TRUE)
```

For the `Master`, `Manager`, and `AwardsManagers` tables:

-   `Master`

    -   Primary keys: `playerID`

-   `Managers`

    -   Primary keys: `yearID`, `teamID`, `inseason`

    -   Foreign Keys:

        -   `playerID` = `Master$playerID` (many-to-1)

-   `AwardsManagers`:

    -   Primary keys: `playerID`, `awardID`, `yearID`
    -   `playerID` = `Master$playerID` (many-to-1)

```{r}
dm2 <- dm_from_data_frames(list(Master = Lahman::Master,
                                Managers = Lahman::Managers,
                                AwardsManagers = Lahman::AwardsManagers)) %>%
  dm_set_key("Master", "playerID") %>%
  dm_set_key("Managers", c("yearID", "teamID", "inseason")) %>%
  dm_set_key("AwardsManagers", c("playerID", "awardID", "yearID")) %>%
  dm_add_references(
    Managers$playerID == Master$playerID,
    AwardsManagers$playerID == Master$playerID
  )

dm_create_graph(dm2, rankdir = "LR", columnArrows = TRUE)
```

In the previous diagrams, I do not consider `teamID` and `lgID` as foreign keys even though they appear in multiple tables (and have the same meaning) because they are not primary keys in the tables considered in this exercise.
The `teamID` variable references `Teams$teamID`, and `lgID` does not have its own table.

How would you characterize the relationship between the `Batting`, `Pitching`, and `Fielding` tables?

The `Batting`, `Pitching`, and `Fielding` tables all have a primary key consisting of the `playerID`, `yearID`, and `stint` variables.
They all have a 1-1 relationship to each other.

## Mutating Joins

```{r}
flights2 <- flights %>%
  select(year:day, hour, origin, dest, tailnum, carrier)
```

### Exercise <span class="exercise-number">13.4.6.1</span> {.unnumbered .exercise}

<div class="question">
Compute the average delay by destination, then join on the `airports` data frame so you can show the spatial distribution of delays. Here’s an easy way to draw a map of the United States:
</div>

```{r}
airports %>%
  semi_join(flights, c("faa" = "dest")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

(Don’t worry if you don’t understand what `semi_join()` does — you’ll learn about it next.)

You might want to use the size or color of the points to display the average delay for each airport.

<div class="answer">

```{r}
avg_dest_delays <-
  flights %>%
  group_by(dest) %>%
  # arrival delay NA's are cancelled flights
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c(dest = "faa"))

avg_dest_delays %>%
  ggplot(aes(lon, lat, colour = delay)) +
    borders("state") +
    geom_point() +
    coord_quickmap()
```

</div>

### Exercise <span class="exercise-number">13.4.6.2</span> {.unnumbered .exercise}

<div class="question">
Add the location of the origin and destination (i.e. the `lat` and `lon`) to `flights`.
</div>

<div class="answer">

You can perform one join after another. If duplicate variables are found, by default, dplyr will distinguish the two by adding `.x`, and `.y` to the ends of the variable names to solve naming conflicts.
```{r}
airport_locations <- airports %>%
  select(faa, lat, lon)

flights %>%
    select(year:day, hour, origin, dest) %>%
  left_join(
    airport_locations,
    by = c("origin" = "faa")
  ) %>%
  left_join(
    airport_locations,
    by = c("dest" = "faa")
  )
```
This default can be over-ridden using the `suffix` argument.
```{r}
airport_locations <- airports %>%
  select(faa, lat, lon)

flights %>%
    select(year:day, hour, origin, dest) %>%
  left_join(
    airport_locations,
    by = c("origin" = "faa")
  ) %>%
  left_join(
    airport_locations,
    by = c("dest" = "faa"),
    suffix = c("_origin", "_dest")
    # existing lat and lon variables in tibble gain the _origin suffix
    # new lat and lon variables are given _dest suffix
  )
```
It's always good practice to have clear variable names.

</div>

### Exercise <span class="exercise-number">13.4.6.3</span> {.unnumbered .exercise}

<div class="question">
Is there a relationship between the age of a plane and its delays?
</div>

<div class="answer">

The question does not specify whether the relationship is with departure delay
or arrival delay. I will look at both.

To compare the age of the plane to flights delay, I merge `flights` with
the `planes`, which contains a variable `plane_year`, with the year in which the 
plane was built. To look at the relationship between plane age and departure
delay, I will calculate the average arrival and departure delay for each age 
of a flight.
```{r}
plane_cohorts <- inner_join(flights,
           select(planes, tailnum, plane_year = year), 
           by = "tailnum") %>%
  mutate(age = year - plane_year) %>%
  filter(!is.na(age)) %>%
  group_by(age) %>%
  summarise(dep_delay_mean = mean(dep_delay, na.rm = TRUE),
            dep_delay_sd = sd(dep_delay, na.rm = TRUE),
            arr_delay_mean = mean(arr_delay, na.rm = TRUE),
            arr_delay_sd = sd(arr_delay, na.rm = TRUE),
            n = n())
```

I will look for a relationship between departure delay and age by plotting 
age against the average departure delay. The average departure delay is increasing
for planes with ages up until 10 years. After that the departure delay decreases 
or levels off. The decrease in departure delay could be because older planes
with many mechanical issues are removed from service or because air lines schedule
these planes with enough time so that mechanical issues do not delay them.
There are few planes older than 30 years, so the relationship is mostly noise
after that.

```{r}
ggplot(plane_cohorts, aes(x = age, y = dep_delay_mean)) +
  geom_ribbon(mapping = aes(ymin = dep_delay_mean - 
                              2 * dep_delay_sd / sqrt(n), 
                            ymax = dep_delay_mean + 
                              2 * dep_delay_sd / sqrt(n)),
              alpha = 0.3) +
  geom_point() +
  scale_x_continuous("Age of plane (years)", breaks = seq(0, 50, by = 10)) +
  scale_y_continuous("Mean Departure Delay (minutes)")

```

There is a similar relationship in arrival delays. Delays increase with the age 
of the plane until ten years, then it declines and flattens out.
```{r}
ggplot(plane_cohorts, aes(x = age, y = arr_delay_mean)) +
  geom_ribbon(mapping = aes(ymin = arr_delay_mean - 
                              2 * arr_delay_sd / sqrt(n), 
                            ymax = arr_delay_mean + 
                              2 * arr_delay_sd / sqrt(n)),
              alpha = 0.3) +
  geom_point() +
  scale_x_continuous("Age of Plane (years)", breaks = seq(0, 50, by = 10)) +
  scale_y_continuous("Mean Departure Delay (minutes)")
```

</div>

### Exercise <span class="exercise-number">13.4.6.4</span> {.unnumbered .exercise}

<div class="question">
What weather conditions make it more likely to see a delay?
</div>

<div class="answer">

Almost any amount or precipitation is associated with a delay.
However, there is not a strong a trend above 0.02 in. of precipitation.
```{r}
flight_weather <-
  flights %>%
  inner_join(weather, by = c("origin" = "origin",
                            "year" = "year",
                            "month" = "month",
                            "day" = "day",
                            "hour" = "hour"))

flight_weather %>%
  group_by(precip) %>%
  summarise(delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = precip, y = delay)) +
    geom_line() + geom_point()
```

</div>

### Exercise <span class="exercise-number">13.4.6.5</span> {.unnumbered .exercise}

<div class="question">
What happened on June 13, 2013? Display the spatial pattern of delays, and then use Google to cross-reference with the weather.
</div>

<div class="answer">

There was a large series of storms (derechos) in the southeastern US (see [June 12-13, 2013 derecho series](https://en.wikipedia.org/wiki/June_12%E2%80%9313,_2013_derecho_series))

The largest delays are in Tennessee (Nashville), the Southeast, and the Midwest, which were the locations of the derechos:
```{r}
library(viridis)
flights %>%
  filter(year == 2013, month == 6, day == 13) %>%
  group_by(dest) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE)) %>%
  inner_join(airports, by = c("dest" = "faa")) %>%
  ggplot(aes(y = lat, x = lon, size = delay, colour = delay)) +
  borders("state") +
  geom_point() +
  coord_quickmap() +
  scale_colour_viridis()

```

</div>

## Filtering Joins

### Exercise <span class="exercise-number">13.5.1.1</span> {.unnumbered .exercise}

<div class="question">
What does it mean for a flight to have a missing `tailnum`? What do the tail numbers that don’t have a matching record in planes have in common? (Hint: one variable explains ~90% of the problems.)
</div>

<div class="answer">

American Airlines (AA) and Envoy Airlines (MQ) don't report tail numbers.
```{r}
flights %>%
  anti_join(planes, by = "tailnum") %>%
  count(carrier, sort = TRUE)

```

</div>

### Exercise <span class="exercise-number">13.5.1.2</span> {.unnumbered .exercise}

<div class="question">
Filter flights to only show flights with planes that have flown at least 100 flights.
</div>

<div class="answer">

```{r}
planes_gt100 <-
  filter(flights) %>%
  group_by(tailnum) %>%
  count() %>%
  filter(n > 100)

flights %>%
  semi_join(planes_gt100, by = "tailnum")

```

</div>

### Exercise <span class="exercise-number">13.5.1.3</span> {.unnumbered .exercise}

<div class="question">
Combine `fueleconomy::vehicles` and `fueleconomy::common` to find only the records for the most common models.
</div>

<div class="answer">

The columns `make` and `model` are the primary key for `fueleconomy::common`.
```{r}
fueleconomy::vehicles %>%
  semi_join(fueleconomy::common, by = c("make", "model"))
```

Why is `model` alone not a primary key? It is possible for two car brands
(`make`) to produce a car with the same name (`make`). In both the `vehicles`
and `common` data we can find examples.
```{r}
fueleconomy::vehicles %>% 
  distinct(model, make) %>% 
  group_by(model) %>%
  filter(n() > 1) %>%
  arrange(model) %>%
  head()
```
```{r}
fueleconomy::common %>% 
  distinct(model, make) %>% 
  group_by(model) %>%
  filter(n() > 1) %>%
  arrange(model) %>%
  head()
```
If we were to merge these data `model` alone, there would be incorrect matches.
For example, see "Limousine".

</div>

### Exercise <span class="exercise-number">13.5.1.4</span> {.unnumbered .exercise}

<div class="question">
Find the 48 hours (over the course of the whole year) that have the worst delays.
Cross-reference it with the weather data. Can you see any patterns?
</div>

<div class="answer">

I will start with clarifying the scope of the question. 
The question asks us to find the 48 hours over the course of the entire year.
I interpret that as meaning 48 not-necessarily contiguous hours, as opposed a single contiguous period of 48-hours.
The question does not specify what is meant by "worst delays".
Since the `weather` data only contains data for the New York City airports, I will use departure delays, which would seem to be more sensitive to New York City weather conditions than arrival delays.

For hour, use the scheduled departure time rather than the actual departure time.
If planes are delayed due to weather conditions, the weather conditions during
the scheduled time are more important than the actual departure time, at which
point, the weather should be better.

```{r}
dep_delays_hour <- flights %>%
  mutate(hour = sched_dep_time %/% 100) %>%
  group_by(year, month, day, hour) %>%
  summarise(dep_delay = mean(dep_delay, na.rm = TRUE))

```

```{r}
dep_delays_hour %>%
  arrange(desc(dep_delay)) %>%
  print(n = 48)
```

The primary key of the `weather` dataset is `year`, `month`, `day`, `hour`, 
`origin`. 
However, the primary key of `dep_delays_hour` is (`year`, `month`, `day`, `hour`), since it combines flights from all airports.
I will create a data fram `weather_hourly` with the averages of temperature, wind speed, and precipitation measurements from the airports.
```{r}
weather_hourly <- weather %>%
  # daily values for each airport
  group_by(year, month, day, hour) %>%
  summarise(temp = mean(temp),
            wind_speed = mean(wind_speed),
            precip = mean(precip))

```

Merge the hourly weather data with hourly departure delays.
```{r}
dep_delays_hour <- dep_delays_hour %>%
  left_join(weather_hourly, by = c("year", "month", "day", "hour"))

```

```{r}
dep_delays_hour %>%
  arrange(desc(dep_delay)) %>%
  print(n = 48)
```

```{r}
ggplot(dep_delays_hour, aes(x = wind_speed, y = dep_delay)) +
  geom_point() +
  geom_smooth()
```

```{r}
ggplot(dep_delays_hour, aes(x = precip, y = dep_delay)) +
  geom_point() +
  geom_smooth()
```

```{r}
ggplot(dep_delays_hour, aes(x = wind_speed, y = dep_delay)) +
  geom_point() +
  geom_smooth()
```

</div>

### Exercise <span class="exercise-number">13.5.1.5</span> {.unnumbered .exercise}

<div class="question">
What does `anti_join(flights, airports, by = c("dest" = "faa"))` tell you? What does `anti_join(airports, flights, by = c("faa" = "dest"))` tell you?
</div>

<div class="answer">

`anti_join(flights, airports, by = c("dest" = "faa"))` are flights that go to an airport that is not in FAA list of destinations, likely foreign airports.

`anti_join(airports, flights, by = c("faa" = "dest"))` are US airports that don't have a flight in the data, meaning that there were no flights to that airport **from** New York in 2013.

</div>

### Exercise <span class="exercise-number">13.5.1.6</span> {.unnumbered .exercise}

<div class="question">
You might expect that there’s an implicit relationship between plane and airline, because each plane is flown by a single airline. Confirm or reject this hypothesis using the tools you’ve learned above.
</div>

<div class="answer">

At each point in time, each plane is flown by a single airline.
However, a plane can be sold and fly for multiple airlines.
Logically, it is possible that a plane can fly for multiple airlines over the course of its lifetime.
But, it is not necessarily the case that a plane will fly for more than one airline in this  data, especially since it comprises only a year of data.
So let's check to see if there are any planes in the data flew for multiple airlines.

First, find all 
```{r mulit_carrier_planes}
planes_carriers <-
  flights %>%
  filter(!is.na(tailnum)) %>%
  distinct(tailnum, carrier)

```
There are `r nrow(multi_carrier_planes)` planes in this dataset that have flown for more than one airline.

To find the airlines for these multi-airline planes:
```{r multi_carrier_planes_with_carrier}
multi_carrier_planes <- flights %>%
  semi_join(multi_carrier_planes, by = "tailnum") %>%
  select(tailnum, carrier) %>%
  distinct() %>%
  arrange(tailnum)
multi_carrier_planes
```

The names of airlines are easier to understand than the two-letter carrier codes.
The `airlines` data frame contains the names of the airlines.

```{r carrier_transfer_tbl}
carrier_transfer_tbl <- multi_carrier_planes %>%
  left_join(airlines, by = "carrier") %>%
  distinct() %>%
  arrange(tailnum, carrier)

carrier_transfer_tbl
```

</div>

## Join problems

No exercises

## Set operations

No exercises
