# Tibbles

## Prerquisites

```{r message=FALSE}
library("tidyverse")
```

Functions and packages covered in this chapter:

- package **tibble**
- `as_tibble`, `tibble`

## Creating Tibbles

Why might you want to create non-syntactic variable names? 
Since variable names are often used as in plots (e.g. axis-titles) or headers in tables, where having spaces or other characters that are invalid R variable names is useful.
Those functions will have ways to use text other than the column.

## Tibbles vs. data.frame

Discuss the definition of a data frame.

What is the traditional R `data.frame`?

In general, discuss how this "dialect" of R relates to base R and other R that they will see.

Also, need to discuss types of variables.

If `nycflights::flights` were printed in the console it would be much worse. Just try it, I dare you.
```{r eval=FALSE}
as.data.frame(nycflights13::flights) 
```


## Subsetting

**Note** Warnings about partial matching! What is it and why is it dangerous.



## Interacting with older code

**Note** Not all older functions work with tibbles (an example includes giAmelia); usually because they rely on quirks in `data.frame` behavior that `tibbles` "fix". Use `as.data.frame()` to turn a tibble back into a `data.frame`. 
This is usually because of `[` and the way it inconsistenly returns a vector or a data frame.
With tibbles `[` always returns a data frame

## Exercises

1. How can you tell if an object is a tibble? (Hint: try printing `mtcars`, which is a regular data frame).

```{r}
mtcars
```

```{r}
class(mtcars)
```

```{r}
class(as_tibble(mtcars))
```

Tibbles will only print out a limited number of rows and show the class on top of each column. Addtionally, tibbles have class `"tbl_df"` and `"tbl_"` in addition to `"data.frame"`.

2. Compare and contrast the following operations on a `data.frame` and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?

```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

```{r}
tbl <- as_tibble(df)
tbl$x
tbl[, "xyz"]
tbl[, c("abc", "xyz")]
```

Using `$` a data.frame will partially complete the column. So even though we wrote `df$x` it returned `df$xyz`. This saves a few keystrokes, but can result in accidentally using a different variable than you thought you were using.

With data.frames, with `[` the type of object that is returned differs on the number of columns. If it is one column, it won't return a data.frame, but instead will return a vector. With more than one column, then it will return a data.frame. This is fine if you know what you are passing in, but suppose you did `df[ , vars]` where `vars` was a variable. Then you what that code does depends on `length(vars)` and you'd have to write code to account for those situations or risk bugs.

3. If you have the name of a variable stored in an object, e.g. `var <- "mpg"`, how can you extract the reference variable from a tibble?

You can use the double bracket, like `df[[var]]`. You cannot use the dollar sign,
becuase `df$var` would look for a column named `var`. 

4. Practice referring to non-syntactic names in the following data frame by:

    1. Extracting the variable called 1.
    2. Plotting a scatterplot of 1 vs 2.
    3. Creating a new column called 3 which is 2 divided by 1.
    4. Renaming the columns to one, two and three.


```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```

Extract the variable called 1:
```{r}
annoying[["1"]]
```
or
```{r}
annoying$`1`
```

A scatterplot of `1` vs. `2`:
```{r}
ggplot(annoying, aes(x = `1`, y = `2`)) +
  geom_point()
```

A new column `3` with is `2` divided by `1`:
```{r}
annoying[["3"]] <- annoying$`2` / annoying$`1`
```
or
```{r}
annoying[["3"]] <- annoying[["2"]] / annoying[["1"]]
```


Renaming the columns to `one`, `two`, and `three`:
```{r}
annoying <- rename(annoying, one = `1`, two = `2`, three = `3`)
glimpse(annoying)
```

5. What does `tibble::enframe()` do? When might you use it?

It converts named vectors to a data frame with names and values
```{r eval=FALSE}
?tibble::enframe
```

```{r}
enframe(c(a = 1, b = 2, c = 3))
```


6. What option controls how many additional column names are printed at the footer of a tibble?

The print function for tibbles is in `print.tbl_df`:
```{r eval=FALSE}
?print.tbl_df
```
The option `n_extra` determines the number of extra columns to print information for.

