# Tidy data {#tidy-data .r4ds-section}

## Introduction {#introduction-6 .r4ds-section}

```{r setup,message=FALSE,cache=FALSE}
library("tidyverse")
```

## Tidy data {#tidy-data-1 .r4ds-section}

### Exercise 12.2.1 {.unnumbered .exercise data-number="12.2.1"}

<div class="question">
Using prose, describe how the variables and observations are organized in each of the sample tables.
</div>

<div class="answer">

In table `table1`, each row represents a (country, year) combination.
The columns `cases` and `population` contain the values for those variables.
```{r}
table1
```

In `table2`, each row represents a (country, year, variable) combination.
The column `count` contains the values of variables `cases` and `population` in separate rows.
```{r}
table2
```

In `table3`, each row represents a (country, year) combination.
The column `rate` provides the values of both `cases` and `population` in a string formatted like `cases / population`.
```{r}
table3
```

Table 4 is split into two tables, one table for each variable.
The table `table4a` contains the values of cases and `table4b` contains the values of population.
Within each table, each row represents a country, each column represents a year, and the cells are the value of the table's variable for that country and year.
```{r}
table4a
```
```{r}
table4b
```

</div>

### Exercise 12.2.2 {.unnumbered .exercise data-number="12.2.2"}

<div class="question">

Compute the `rate` for `table2`, and `table4a` + `table4b`. 
You will need to perform four operations:

1.  Extract the number of TB cases per country per year.
1.  Extract the matching population per country per year.
1.  Divide cases by population, and multiply by 10000.
1.  Store back in the appropriate place.

Which representation is easiest to work with? 
Which is hardest? 
Why?

</div>

<div class="answer">

To calculate cases per person, we need to divide cases by population for each country and year.
This is easiest if the cases and population variables are two columns in a data frame in which rows represent (country, year) combinations.

Table 2: First, create separate tables for cases and population and ensure that they are sorted in the same order.
```{r}
t2_cases <- filter(table2, type == "cases") %>%
  rename(cases = count) %>%
  arrange(country, year)
t2_population <- filter(table2, type == "population") %>%
  rename(population = count) %>%
  arrange(country, year)
```
Then create a new data frame with the population and cases columns, and calculate the cases per capita in a new column.
```{r}
t2_cases_per_cap <- tibble(
  year = t2_cases$year,
  country = t2_cases$country,
  cases = t2_cases$cases,
  population = t2_population$population
) %>%
  mutate(cases_per_cap = (cases / population) * 10000) %>%
  select(country, year, cases_per_cap)
```

To store this new variable in the appropriate location, we will add new rows to `table2`.
```{r}
t2_cases_per_cap <- t2_cases_per_cap %>%
  mutate(type = "cases_per_cap") %>%
  rename(count = cases_per_cap)
```
```{r}
bind_rows(table2, t2_cases_per_cap) %>%
  arrange(country, year, type, count)
```
Note that after adding the `cases_per_cap` rows, the type of `count` is coerced to `numeric` (double) because `cases_per_cap` is not an integer.

For `table4a` and `table4b`, create a new table for cases per capita, which we'll name `table4c`, with country rows and year columns.
```{r}
table4c <-
  tibble(
    country = table4a$country,
    `1999` = table4a[["1999"]] / table4b[["1999"]] * 10000,
    `2000` = table4a[["2000"]] / table4b[["2000"]] * 10000
  )
table4c
```

Neither table is particularly easy to work with.
Since `table2` has separate rows for cases and population we needed to generate a table with columns for cases and population where we could
calculate cases per capita.
`table4a` and `table4b` split the cases and population variables into different tables which
made it easy to divide cases by population.
However, we had to repeat this calculation for each row.

The ideal format of a data frame to answer this question is one with columns `country`, `year`, `cases`, and `population`.
Then problem could be answered with a single `mutate()` call.

</div>

### Exercise 12.2.3 {.unnumbered .exercise data-number="12.2.3"}

<div class="question">

Recreate the plot showing change in cases over time using `table2` instead of `table1`. 
What do you need to do first?

</div>

<div class="answer">

Before creating the plot with change in cases over time, we need to filter `table` to only include rows representing cases of TB.
```{r}
table2 %>%
  filter(type == "cases") %>%
  ggplot(aes(year, count)) +
  geom_line(aes(group = country), colour = "grey50") +
  geom_point(aes(colour = country)) +
  scale_x_continuous(breaks = unique(table2$year)) +
  ylab("cases")
```

</div>

## Pivoting {#pivoting .r4ds-section}

This code is reproduced from the chapter because it is needed by the exercises.
```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```

### Exercise 12.3.1 {.unnumbered .exercise data-number="12.3.1"}

<div class="question">

Why are `pivot_longer()` and `pivot_wider()` not perfectly symmetrical?
Carefully consider the following example:

Carefully consider the following example:

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```

(Hint: look at the variable types and think about column names.)

`pivot_longer()` has a `names_ptype` argument, e.g. `names_ptype = list(year = double())`. What does it do?

</div>

<div class="answer">

The functions `pivot_longer()` and `pivot_wider()` are not perfectly symmetrical because column type information is lost when a data frame is converted from wide to long.
The function `pivot_longer()` stacks multiple columns which may have had multiple data types into a single column with a single data type.
This transformation throws away the individual data types of the original columns.
The function `pivot_wider()` creates column names from values in column.
These column names will always be treated as `character` values by `pivot_longer()` so if the original variable used to create the column names did not have a `character` data type, then the round-trip will not reproduce the same dataset.

In the provided example, columns have the following data types:
```{r}
glimpse(stocks)
```

The `pivot_wider()` expression pivots the table to create a data frame with years as column names, and the values in `return` as the column values.
```{r}
stocks %>%
  pivot_wider(names_from = year, values_from = return) 
```

The `pivot_longer()` expression unpivots the table, returning it to a tidy data frame with columns for `half`, `year`, and `return`.
```{r}
stocks %>%
  pivot_wider(names_from = year, values_from = return)%>%
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
```
There is one difference, in the new data frame, `year` has a data type of `character` rather than `numeric`. 
The `names_to` column created from column names by `pivot_longer()` will be character by default, which is usually a safe assumption, since syntactically valid-column names can only be character values.

The original data types of column which `pivot_wider()` used to create the column names was not stored, so `pivot_longer()` has no idea that the column names in this case should be numeric values.
<!--
The `names_ptype` argument of `pivot_longer()` provide that information. 
Using `names_ptype = list(year = double())` informs `pivot_longer()` that the column created from column names should be a numeric vector.
Adding this argument to `pivot_longer()` in the example above will make the output of the experession the same as the original `stocks` data frame.
--> 
In the current version of `tidyr`, the `names_ptype` argument does not convert the `year` column to a numeric vector, and it will raise an error.
```{r error=TRUE}
stocks %>%
  pivot_wider(names_from = year, values_from = return)%>%
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return",
               names_ptype = list(year = double()))
```

Instead, use the `names_transform` argument to `pivot_longer()`, which provides a function to coerce the column to a different data type.
```{r}
stocks %>%
  pivot_wider(names_from = year, values_from = return)%>%
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return",
               names_transform = list(year = as.numeric))

```

</div>

### Exercise 12.3.2 {.unnumbered .exercise data-number="12.3.2"}

<div class="question">

Why does this code fail?

```{r error=TRUE}
table4a %>% 
  pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")
```
</div>

<div class="answer">

The code fails because the column names `1999` and `2000` are not non-syntactic variable names.[^non-syntactic]
When selecting variables from a data frame, tidyverse functions will interpret numbers, like `1999` and `2000`, as column numbers.
In this case, `pivot_longer()` tries to select the 1999th and 2000th column of the data frame.
To select the columns `1999` and `2000`, the names must be surrounded in backticks (`\``) or provided as strings.

```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
```
```{r}
table4a %>% 
  pivot_longer(c("1999", "2000"), names_to = "year", values_to = "cases")
```

</div>

### Exercise 12.3.3 {.unnumbered .exercise data-number="12.3.3"}

<div class="question">

What would happen if you widen this table? 
Why? 
How could you add a new column to uniquely identify each value?

```{r}
people <- tribble(
  ~name, ~key, ~value,
  #-----------------|--------|------
  "Phillip Woods",  "age", 45,
  "Phillip Woods", "height", 186,
  "Phillip Woods", "age", 50,
  "Jessica Cordero", "age", 37,
  "Jessica Cordero", "height", 156
)
glimpse(people)
```

</div>

<div class="answer">

Widening this data frame using `pivot_wider()` produces columns that are lists of numeric vectors because the `name` and `key` columns do not uniquely identify rows.
In particular, there are two rows with values for the age of "Phillip Woods".
```{r}
pivot_wider(people, names_from="name", values_from = "value")
```

We could solve the problem by adding a row with a distinct observation count for each combination of name and key.
```{r}
people2 <- people %>%
  group_by(name, key) %>%
  mutate(obs = row_number())
people2
```
We can make `people2` wider because the combination of `name` and `obs` will uniquely identify the rows in the wide data frame.
```{r}
pivot_wider(people2, names_from="name", values_from = "value")
```

Another way to solve this problem is by keeping only distinct rows of the name and key values, and dropping duplicate rows.
```{r}
people %>%
  distinct(name, key, .keep_all = TRUE) %>%
  pivot_wider(names_from="name", values_from = "value")
```
However, before doing this understand why there are duplicates in the data.
The duplicate values may not be just a nuisance, but may indicate deeper problems with the data.

</div>

### Exercise 12.3.4 {.unnumbered .exercise data-number="12.3.4"}

<div class="question">

Tidy the simple tibble below. Do you need to make it wider or longer? 
What are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes", NA, 10,
  "no", 20, 12
)
```

</div>

<div class="answer">

To tidy the `preg` table use `pivot_longer()` to create a long table.
The variables in this data are:

-   `sex` ("female", "male")
-   `pregnant` ("yes", "no")
-   `count`, which is a non-negative integer representing the number of observations.

The observations in this data are unique combinations of sex and pregnancy status.

```{r}
preg_tidy <- preg %>%
  pivot_longer(c(male, female), names_to = "sex", values_to = "count")
preg_tidy
```

Remove the (male, pregnant) row with a missing value to simplify the tidied data frame.

```{r}
preg_tidy2 <- preg %>%
  pivot_longer(c(male, female), names_to = "sex", values_to = "count", values_drop_na = TRUE)
preg_tidy2
```

This an example of turning an explicit missing value into an implicit missing value, which is discussed in the upcoming section, [Missing Values](https://r4ds.had.co.nz/tidy-data.html#missing-values-3) section.
The missing (male, pregnant) row represents an implicit missing value because the value of `count` can be inferred from its absence.
In the tidy data, we can represent rows with missing values of `count` either explicitly with an `NA` (as in `preg_tidy`) or implicitly by the absence of a row (as in `preg_tidy2`). 
But in the wide data, the missing values can only be represented explicitly.

Though we have already done enough to make the data tidy, there are some other transformations that can clean the data further.
If a variable takes two values, like `pregnant` and `sex`, it is often preferable to store them as logical vectors.
```{r}
preg_tidy3 <- preg_tidy2 %>%
  mutate(
    female = sex == "female",
    pregnant = pregnant == "yes"
  ) %>%
  select(female, pregnant, count)
preg_tidy3
```
In the previous data frame, I named the logical variable representing the sex `female`, not `sex`.
This makes the meaning of the variable self-documenting.
If the variable were named `sex` with values `TRUE` and `FALSE`, without reading the documentation, we wouldn't know whether `TRUE` means male or female.

Apart from some minor memory savings, representing these variables as logical vectors results in more clear and concise code.
Compare the `filter()` calls to select non-pregnant females from `preg_tidy2` and `preg_tidy`.
```{r}
filter(preg_tidy2, sex == "female", pregnant == "no")
filter(preg_tidy3, female, !pregnant)
```

</div>

## Separating and uniting {#separating-and-uniting .r4ds-section}

### Exercise 12.4.1 {.unnumbered .exercise data-number="12.4.1"}

<div class="question">

What do the extra and fill arguments do in `separate()`?
Experiment with the various options for the following two toy datasets.

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))
```

</div>

<div class="answer">

The `extra` argument tells `separate()` what to do if there are too many pieces, and the `fill` argument tells it what to do if there aren't enough.
By default, `separate()` drops extra values with a warning.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))
```

Adding the argument, `extra = "drop"`, produces the same result as above but without the warning.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "drop")
```

Setting `extra = "merge"`, then the extra values are not split, so `"f,g"` appears in column three.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "merge")
```

In this example, one of the values, `"d,e"`, has too few elements.
The default for `fill` is similar to those in `separate()`; 
it fills columns with missing values but emits a warning. 
In this example, the 2nd row of column `three` is `NA`.

```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))
```

Alternative options for the `fill` are `"right"`, to fill with missing values from the right, but without a warning
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"), fill = "right")
```
The option `fill = "left"` also fills with missing values without emitting a warning, but this time from the left side.
Now, the 2nd row of column `one` will be missing, and the other values in that row are shifted right.
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"), fill = "left")
```

</div>

### Exercise 12.4.2 {.unnumbered .exercise data-number="12.4.2"}

<div class="question">

Both `unite()` and `separate()` have a `remove` argument. 
What does it do?
Why would you set it to `FALSE`?

</div>

<div class="answer">

The `remove` argument discards input columns in the result data frame. You would set it to `FALSE` if you want to create a new variable, but keep the old one.

</div>

### Exercise 12.4.3 {.unnumbered .exercise data-number="12.4.3"}

<div class="question">
Compare and contrast `separate()` and `extract()`, Why are there three variations of separation (by position, by separator, and with groups), but only one unite?
</div>

<div class="answer">

The function `separate()`, splits a column into multiple columns by separator, if the `sep` argument is a character vector, or by character positions, if `sep` is numeric.
```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  separate(x, c("variable", "into"), sep = "_")

# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  separate(x, c("variable", "into"), sep = c(1))
```

The function `extract()` uses a regular expression to specify groups in character vector and split that single character vector into multiple columns.
This is more flexible than `separate()` because it does not require a common
separator or specific column positions.

```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])_([0-9])")

# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])([0-9])")

# example that separate could not parse
tibble(x = c("X1", "X20", "AA11", "AA2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z]+)([0-9]+)")
```

Both `separate()` and `extract()` convert a single column to many columns.
However, `unite()` converts many columns to one, with a choice of a separator to include between column values.

```{r}
tibble(variable = c("X", "X", "Y", "Y"), id = c(1, 2, 1, 2)) %>%
  unite(x, variable, id, sep = "_")
```

In other words, with `extract()` and `separate()` only one column can be chosen,
but there are many choices how to split that single column into different columns.
With `unite()`, there are many choices as to which columns to include, but only one 
choice as to how to combine their contents into a single vector.

</div>

## Missing values {#missing-values-3 .r4ds-section}

### Exercise 12.5.1 {.unnumbered .exercise data-number="12.5.1"}

<div class="question">
Compare and contrast the `fill` arguments to `pivot_wider()` and `complete()`.
</div>

<div class="answer">

The `values_fill` argument in `pivot_wider()` and the `fill` argument to `complete()` both set values to replace `NA`. 
Both arguments accept named lists to set values for each column.
Additionally, the `values_fill` argument of `pivot_wider()` accepts a single value.
In `complete()`, the fill argument also sets a value to replace `NA`s but it is a named list, allowing for different values for different variables.
Also, both cases replace both implicit and explicit missing values.

For example, this will fill in the missing values of the long data frame with `0` by the `values_fill` argument of `pivot_wider()`:
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return,
              values_fill = 0)
```
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks %>% 
  pivot_wider(names_from = year, values_from = return,
              values_fill = 0)
```

For example, this will fill in the missing values of the long data frame with `0` by `complete()`:
```{r}
stocks %>% 
 complete(year, qtr, fill=list(return=0))
```

</div>

### Exercise 12.5.2 {.unnumbered .exercise data-number="12.5.2"}

<div class="question">

What does the `direction` argument to `fill()` do?

</div>

<div class="answer">

With `fill`, the `direction` determines whether `NA` values should be replaced by the previous non-missing value (`"down"`) or the next non-missing value (`"up"`).

</div>

## Case Study {#case-study .r4ds-section}

This code is repeated from the chapter because it is needed by the exercises.

```{r}
who1 <- who %>% 
  pivot_longer(
    cols = new_sp_m014:newrel_f65, 
    names_to = "key", 
    values_to = "cases", 
    values_drop_na = TRUE
  )
who1
```

```{r}
who2 <- who1 %>% 
  mutate(names_from = stringr::str_replace(key, "newrel", "new_rel"))
who2
```

```{r}
who3 <- who2 %>% 
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```

```{r}
who3 %>%
  count(new)
```

```{r}
who4 <- who3 %>%
  select(-new, -iso2, -iso3)
```

```{r}
who5 <- who4 %>%
  separate(sexage, c("sex", "age"), sep = 1)
who5
```

### Exercise 12.6.1 {.unnumbered .exercise data-number="12.6.1"}

<div class="question">

In this case study, I set `na.rm = TRUE` just to make it easier to check that we had the correct values. 
Is this reasonable? 
Think about how missing values are represented in this dataset. 
Are there implicit missing values? 
What’s the difference between an `NA` and zero?

</div>

<div class="answer">

The reasonableness of using `na.rm = TRUE` depends on how missing values are represented in this dataset.
The main concern is whether a missing value means that there were no cases of TB or whether it means that the WHO does not have data on the number of TB cases.
Here are some things we should look for to help distinguish between these cases.

-   If there are no 0 values in the data, then missing values may be used to indicate no cases.

-   If there are both explicit and implicit missing values, then it suggests that missing values
    are being used differently. In that case, it is likely that explicit missing values would
    mean no cases, and implicit missing values would mean no data on the number of cases.

First, I'll check for the presence of zeros in the data. 
```{r}
who1 %>%
  filter(cases == 0) %>%
  nrow()
```
There are zeros in the data, so it appears that cases of zero TB are explicitly indicated, and the value of`NA` is used to indicate missing data.

Second, I should check whether all values for a (country, year) are missing or whether it is possible for only some columns to be missing.
```{r}
pivot_longer(who, c(new_sp_m014:newrel_f65), names_to = "key", values_to = "cases") %>%
  group_by(country, year) %>%
  mutate(prop_missing = sum(is.na(cases)) / n()) %>%
  filter(prop_missing > 0, prop_missing < 1)

```
From the results above, it looks like it is possible for a (country, year) row to contain some, but not all, missing values in its columns.

Finally, I will check for implicit missing values. 
Implicit missing values are (`year`, `country`) combinations that do not appear in the data.
```{r}
nrow(who)

who %>% 
  complete(country, year) %>%
  nrow()
```
Since the number of complete cases of (`country`, `year`) is greater than the number of rows in `who`, there are some implicit values.
But that doesn't tell us what those implicit missing values are.
To do this, I will use the `anti_join()` function introduced in the later [Relational Data](https://r4ds.had.co.nz/relational-data.html#filtering-joins) chapter.
```{r}
anti_join(complete(who, country, year), who, by = c("country", "year")) %>% 
  select(country, year) %>% 
  group_by(country) %>% 
  # so I can make better sense of the years
  summarise(min_year = min(year), max_year = max(year))
```
All of these refer to (`country`, `year`) combinations for years prior to the existence of the country.
For example, Timor-Leste achieved independence in 2002, so years prior to that are not included in the data.

To summarize:

-   `0` is used to represent no cases of TB.
-   Explicit missing values (`NA`s) are used to represent missing data for (`country`, `year`) combinations in which the country existed in that year.
-   Implicit missing values are used to represent missing data because a country did not exist in that year.

</div>

### Exercise 12.6.2 {.unnumbered .exercise data-number="12.6.2"}

<div class="question">

What happens if you neglect the `mutate()` step? 
(`mutate(key = str_replace(key, "newrel", "new_rel")`)

</div>

<div class="answer">

The `separate()` function emits the warning "too few values".
If we check the rows for keys beginning with `"newrel_"`, we see that `sexage` is missing,
and `type = m014`.

```{r}
who3a <- who1 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")

filter(who3a, new == "newrel") %>% head()
```

</div>

### Exercise 12.6.3 {.unnumbered .exercise data-number="12.6.3"}

<div class="question">

I claimed that `iso2` and `iso3` were redundant with country. 
Confirm this claim.

</div>

<div class="answer">

If `iso2` and `iso3` are redundant with `country`, then, within each country, 
there should only be one distinct combination of `iso2` and `iso3` values, which is the case.
```{r}
select(who3, country, iso2, iso3) %>%
  distinct() %>%
  group_by(country) %>%
  filter(n() > 1)
```

This makes sense, since `iso2` and `iso3` contain the 2- and 3-letter country abbreviations for the country.
The `iso2` variable contains each country's [ISO 3166 alpha-2](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2), and the `iso3` variable contains each country's [ISO 3166 alpha-3](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) abbreviation.
You may recognize the ISO 3166-2 abbreviations, since they are almost identical to internet [country-code top level domains](https://en.wikipedia.org/wiki/Country_code_top-level_domain), such as `.uk` (United Kingdom), `.ly` (Libya), `.tv` (Tuvalu), and `.io` (British Indian Ocean Territory).

</div>

### Exercise 12.6.4 {.unnumbered .exercise data-number="12.6.4"}

<div class="question">

For each country, year, and sex compute the total number of cases of TB.
Make an informative visualization of the data.

</div>

<div class="answer">

```{r}
who5 %>%
  group_by(country, year, sex) %>%
  filter(year > 1995) %>%
  summarise(cases = sum(cases)) %>%
  unite(country_sex, country, sex, remove = FALSE) %>%
  ggplot(aes(x = year, y = cases, group = country_sex, colour = sex)) +
  geom_line()
```

A small multiples plot faceting by country is difficult given the number of countries.
Focusing on those countries with the largest changes or absolute magnitudes after providing the context above is another option.

</div>

## Non-tidy data {#non-tidy-data .r4ds-section}

`r no_exercises()`

[ex-12.2.2]: It would be better to join these tables using the methods covered in the [Relational Data](https://r4ds.had.co.nz/relational-data.html). 
             We could use `inner_join(t2_cases, t2_population, by = c("country", "year"))`.

[non-syntactic]: See the [Creating Tibbles](https://r4ds.had.co.nz/tibbles.html#tibbles) section.
