# Tidy Data

## Introduction

```{r message=FALSE}
library(tidyverse)
```

## Tidy data

### Exercise <span class="exercise-number">12.2.1</span> {.unnumbered .exercise}

<div class="question">
Using prose, describe how the variables and observations are organized in each of the sample tables.
</div>

<div class="answer">

In table `table1`, each row represents a (country, year) combination.
The columns `cases` and `population` contain the values for those variables.
```{r}
table1
```

In `table2`, each row represents a (country, year, variable) combination.
The column `count` contains the values of variables `cases` and `population` in separate rows.
```{r}
table2
```

In `table3`, each row represents a (country, year) combination.
The column `rate` provides the values of both `cases` and `population` in a string formatted like `cases / population`.
```{r}
table3
```

Table 4 is split into two tables, one table for each variable.
The table `table4a` contains the values of cases and `table4b` contains the values of population.
Within each table, each row represents a country, each column represents a year, and the cells are the value of the table's variable for that country and year.
```{r}
table4a
```
```{r}
table4b
```

</div>

### Exercise <span class="exercise-number">12.2.2</span> {.unnumbered .exercise}

<div class="question">

Compute the `rate` for `table2`, and `table4a` + `table4b`. 
You will need to perform four operations:

1.  Extract the number of TB cases per country per year.
1.  Extract the matching population per country per year.
1.  Divide cases by population, and multiply by 10000.
1.  Store back in the appropriate place.

Which representation is easiest to work with? 
Which is hardest? 
Why?

</div>

<div class="answer">

To calculate cases per person, we need to divide cases by population for each country and year.
This is easiest if the cases and population variables are two columns in a data frame in which rows represent (country, year) combinations.

Table 2: First, create separate tables for cases and population and ensure that they are sorted in the same order.
```{r}
t2_cases <- filter(table2, type == "cases") %>%
  rename(cases = count) %>%
  arrange(country, year)
t2_population <- filter(table2, type == "population") %>%
  rename(population = count) %>%
  arrange(country, year)
```
Then create a new data frame with the population and cases columns, and calculate the cases per capita in a new column.[^ex-12.2.2]
```{r}
t2_cases_per_cap <- tibble(
  year = t2_cases$year,
  country = t2_cases$country,
  cases = t2_cases$cases,
  population = t2_population$population
) %>%
  mutate(cases_per_cap = (cases / population) * 10000) %>%
  select(country, year, cases_per_cap)
```

To store this new variable in the appropriate location, we will add new rows to `table2`.
```{r}
t2_cases_per_cap <- t2_cases_per_cap %>%
  mutate(type = "cases_per_cap") %>%
  rename(count = cases_per_cap)
```
```{r}
bind_rows(table2, t2_cases_per_cap) %>%
  arrange(country, year, type, count)
```
Note that after adding the `cases_per_cap` rows, the type of `count` is coerced to `numeric` (double) because `cases_per_cap` is not an integer.

For `table4a` and `table4b`, create a new table for cases per capita, which we'll name `table4c`, with country rows and year columns.
```{r}
table4c <-
  tibble(
    country = table4a$country,
    `1999` = table4a[["1999"]] / table4b[["1999"]] * 10000,
    `2000` = table4a[["2000"]] / table4b[["2000"]] * 10000
  )
table4c
```

Neither table is particularly easy to work with.
Since `table2` has separate rows for cases and population we needed to generate a table with columns for cases and population where we could
calculate cases per capita.
`table4a` and `table4b` split the cases and population variables into different tables which
made it easy to divide cases by population.
However, we had to repeat this calculation for each row.

The ideal format of a data frame to answer this question is one with columns `country`, `year`, `cases`, and `population`.
Then problem could be answered with a single `mutate()` call.

</div>

### Exercise <span class="exercise-number">12.2.3</span> {.unnumbered .exercise}

<div class="question">

Recreate the plot showing change in cases over time using `table2` instead of `table1`. 
What do you need to do first?

</div>

<div class="answer">

Before creating the plot with change in cases over time, we need to filter `table` to only include rows representing cases of TB.
```{r}
table2 %>%
  filter(type == "cases") %>%
  ggplot(aes(year, count)) +
  geom_line(aes(group = country), colour = "grey50") +
  geom_point(aes(colour = country)) +
  scale_x_continuous(breaks = unique(table2$year)) +
  ylab("cases")
```

</div>

## Spreading and Gathering

This code is reproduced from the chapter because it is needed by the exercises.
```{r}
tidy4a <- table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases")
tidy4b <- table4b %>%
  gather(`1999`, `2000`, key = "year", value = "cases")
```

### Exercise <span class="exercise-number">12.3.1</span> {.unnumbered .exercise}

<div class="question">

Why are `gather()` and `spread()` not perfectly symmetrical?
Carefully consider the following example:

```{r}
stocks <- tibble(
  year = c(2015, 2015, 2016, 2016),
  half = c(1, 2, 1, 2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>%
  spread(year, return) %>%
  gather(`2015`:`2016`, key = "year", value = "return")
```

</div>

<div class="answer">

The functions `spread()` and `gather()` are not perfectly symmetrical because column type information is lost.
When we use `gather()` on a data frame, it discards the original column types.
It has to coerce all the gathered variables into a single vector with a single type.
Later, if we `spread()` that data frame, the `spread()` function does not know the original data types of the variables.

In this example, the column `year` is numeric.
After running `spread()` and `gather()`, the `year` column is a character vector.
After `spread()`, the years are column names.
```{r}
stocks %>%
  spread(year, return)
```
The following use of `gather()` will create a column, `year`, from the column names.
However, since column names are used to create the key column, the resulting vector is a character vector.
Even though we "know" that the all the column names refer to years, which are numbers, the `gather()` function does not know that.
```{r}
stocks %>%
  spread(year, return) %>%
  gather(`2015`:`2016`, key = "year", value = "return")
```

The functions `spread()` and `gather()` are almost symmetrical if we use the `convert` argument.
When `convert = TRUE`, the `gather()` function will attempt to convert vectors to the appropriate type using `type.convert()`.
```{r}
stocks %>%
  spread(key = "year", value = "return") %>%
  gather(`2015`:`2016`, key = "year", value = "return", convert = TRUE)
```
Now, `year` is an integer vector.
However, this conversion is merely guessing the type of the variables and as such, will not always return the original variable types.
That information is lost.

</div>

### Exercise <span class="exercise-number">12.3.2</span> {.unnumbered .exercise}

<div class="question">

Why does this code fail?

```{r error=TRUE}
table4a %>%
  gather(1999, 2000, key = "year", value = "cases")
```
</div>

<div class="answer">

The code fails because the column names `1999` and `2000` are not non-syntactic variable names.[^non-syntactic]
When selecting variables from a data frame, tidyverse functions will interpret numbers, like `1999` and `2000`, as column numbers.
In this case, `gather()` tries to select 1999th and 2000th column of the data frame.
To select the columns `1999` and `2000`, you can either surround their names in backticks (`\``) or provide them as strings.
```{r}
table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases")
```
```{r}
table4a %>%
  gather("1999", "2000", key = "year", value = "cases")
```

</div>

### Exercise <span class="exercise-number">12.3.3</span> {.unnumbered .exercise}

<div class="question">

Why does spreading this tibble fail? 
How could you add a new column to fix the problem?

```{r}
people <- tribble(
  ~name, ~key, ~value,
  #-----------------|--------|------
  "Phillip Woods",  "age", 45,
  "Phillip Woods", "height", 186,
  "Phillip Woods", "age", 50,
  "Jessica Cordero", "age", 37,
  "Jessica Cordero", "height", 156
)
glimpse(people)
```

```{r error=TRUE}
spread(people, key, value)
```

</div>

<div class="answer">

Spreading this data frame fails because the `name` and `key` columns do not uniquely identify rows.
In particular, ther two rows with values for the age of "Phillip Woods".

After adding a column with an observation count 
```{r}
people2 <- tribble(
  ~name, ~key, ~value, ~obs,
  #-----------------|--------|------|------
  "Phillip Woods", "age", 45, 1,
  "Phillip Woods", "height", 186, 1,
  "Phillip Woods", "age", 50, 2,
  "Jessica Cordero", "age", 37, 1,
  "Jessica Cordero", "height", 156, 1
)
spread(people2, key, value)
```

Another way to solve this problem is by keeping only distinct rows of the name and key values, and dropping duplicate rows.
```{r}
people %>%
  distinct(name, key) %>%
  spread(key, value)
```
However, before doing this you would want to understand why there are duplicates in the data to begin with.
This is usually not mereley a nuissance, but indicates deeper problems with the data.

</div>

### Exercise <span class="exercise-number">12.3.4</span> {.unnumbered .exercise}

<div class="question">

Tidy the simple tibble below. 
Do you need to spread or gather it? 
What are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes", NA, 10,
  "no", 20, 12
)
```

</div>

<div class="answer">

To tidy the `preg` preg tibble, we need to use `gather()`. 
The variables in this data are:

-   `sex` ("female", "male")
-   `pregnant` ("yes", "no")
-   `count`, which is a non-negative integer representing the number of observations.

The observations in this data are unique combinations of sex and pregnancy status.

```{r}
preg_tidy <- preg %>%
  gather(male, female, key = "sex", value = "count")
preg_tidy
```

We can simplify the tidied data frame by removing the (male, pregnant) row with the missing value of `NA`.

```{r}
preg_tidy2 <- preg %>%
  gather(male, female, key = "sex", value = "count", na.rm = TRUE)
preg_tidy2
```

This an example of turning an explicit missing value into an implicit missing value, which is discussed in the upcoming [Missing Values](http://r4ds.had.co.nz/tidy-data.html#missing-values-3) section.
The missing (male, pregnant) row represents is an implicit missing value because the missing value of `count` can be inferred from its absence.
In the tidy data, we can represent rows with missing values of `count` either explicitly with an `NA` (as in `preg_tidy`) or implicitly by the absence of a row (as in `preg_tidy2`). 
But in the wide data, the missing values can only be represented explicitly.

Though we have already done enough to make the data tidy, there's some other changes that can be made to clean this data.
If a variable takes two values, like `pregnant` and `sex`, it is often preferable to store them as logical vectors.
```{r}
preg_tidy3 <- preg_tidy2 %>%
  mutate(
    female = sex == "female",
    pregnant = pregnant == "yes"
  ) %>%
  select(female, pregnant, count)
preg_tidy3
```
In the previous data frame, I named the logical variable representing the sex `female`, not `sex`.
This makes the meaning of the variable self-documenting.
If the variable were named `sex` with values `TRUE` and `FALSE`, without reading the documentation, we wouldn't know whether `TRUE` means male or female.

Apart from some minor memory savings, representing these variables as logical vectors results in more clear and concise code.
Compare the `filter()` calls to select non-pregnant females from `preg_tidy2` and `preg_tidy`.
```{r}
filter(preg_tidy2, sex == "female", pregnant == "no")
filter(preg_tidy3, female, !pregnant)
```

</div>

## Separating and Uniting

### Exercise <span class="exercise-number">12.4.1</span> {.unnumbered .exercise}

<div class="question">

What do the extra and fill arguments do in `separate()`?
Experiment with the various options for the following two toy datasets.


```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))
```

</div>

<div class="answer">

The `extra` argument tells `separate()` what to do if there are too many pieces, and the `fill` argument tells it what to do if there aren't enough.
By default, `separate()` drops extra values with a warning.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))
```

Adding the argument, `extra = "drop"`, produces the same result as above but without the warning.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "drop")
```

Setting `extra = "merge"`, then the extra values are not split, so `"f,g"` appears in column three.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "merge")
```

In this example, one of the values, `"d,e"`, has too few elements.
The default for `fill` is similar to those in `separate()`; 
it fills columns with missing values but emits a warning. 
In this example, the 2nd row of column `three` is `NA`.

```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))
```

Alternative options for the `fill` are `"right"`, to fill with missing values from the right, but without a warning
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"), fill = "right")
```
The option `fill = "left"` also fills with missing values without emitting a warning, but this time from the left side.
Now, the 2nd row of column `one` will be missing, and the other values in that row are shifted right.
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"), fill = "left")
```

</div>

### Exercise <span class="exercise-number">12.4.2</span> {.unnumbered .exercise}

<div class="question">

Both `unite()` and `separate()` have a `remove` argument. 
What does it do?
Why would you set it to `FALSE`?

</div>

<div class="answer">

The `remove` argument discards input columns in the result data frame. You would set it to `FALSE` if you want to create a new variable, but keep the old one.

</div>

### Exercise <span class="exercise-number">12.4.3</span> {.unnumbered .exercise}

<div class="question">
Compare and contrast `separate()` and `extract()`, Why are there three variations of separation (by position, by separator, and with groups), but only one unite?
</div>

<div class="answer">

The function `separate()`, splits a column into multiple columns by separator, if the `sep` argument is a character vector, or by character positions, if `sep` is numeric.
```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  separate(x, c("variable", "into"), sep = "_")

# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  separate(x, c("variable", "into"), sep = c(1))
```

The function `extract()` uses a regular expression to specify groups in character vector and split that single character vector into multiple columns.
This is more flexible than `separate()` because it does not require a common
separator or specific column positions.

```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])_([0-9])")

# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])([0-9])")

# example that separate could not parse
tibble(x = c("X1", "X20", "AA11", "AA2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z]+)([0-9]+)")
```

Both `separate()` and `extract()` convert a single column to many columns.
However, `unite()` converts many columns to one, with a choice of a separator to include between column values.

```{r}
tibble(variable = c("X", "X", "Y", "Y"), id = c(1, 2, 1, 2)) %>%
  unite(x, variable, id, sep = "_")
```

In other words, with `extract()` and `separate()` only one column can be chosen,
but there are many choices how to split that single column into different columns.
With `unite()`, there are many choices as to which columns to include, but only one 
choice as to how to combine their contents into a single vector.

</div>

## Missing Values

### Exercise <span class="exercise-number">12.5.1</span> {.unnumbered .exercise}

<div class="question">
Compare and contrast the `fill` arguments to `spread()` and `complete()`.
</div>

<div class="answer">

In `spread()`, the fill argument explicitly sets the value to replace `NA`s.
In `complete()`, the fill argument also sets a value to replace `NA`s but it is named list, allowing for different values for different variables.
Also, both cases replace both implicit and explicit missing values.

</div>

### Exercise <span class="exercise-number">12.5.2</span> {.unnumbered .exercise}

<div class="question">

What does the `direction` argument to `fill()` do?

</div>

<div class="answer">

With `fill`, the `direction` determines whether `NA` values should be replaced by the previous non-missing value (`"down"`) or the next non-missing value (`"up"`).

</div>

## Case Study

This code is repeated from the chapter because it is needed by the exercises.

```{r}
who1 <- who %>%
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm = TRUE)
glimpse(who1)
```

```{r}
who2 <- who1 %>%
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))
```

```{r}
who3 <- who2 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```

```{r}
who3 %>%
  count(new)
```

```{r}
who4 <- who3 %>%
  select(-new, -iso2, -iso3)
```

```{r}
who5 <- who4 %>%
  separate(sexage, c("sex", "age"), sep = 1)
who5
```


### Exercise <span class="exercise-number">12.6.1</span> {.unnumbered .exercise}

<div class="question">

In this case study, I set `na.rm = TRUE` just to make it easier to check that we had the correct values. 
Is this reasonable? 
Think about how missing values are represented in this dataset. 
Are there implicit missing values? 
What’s the difference between an `NA` and zero?

</div>

<div class="answer">

Perhaps? I would need to know more about the data generation process.
There are zero's in the data, which means they may explicitly be indicating no cases.
```{r}
who1 %>%
  filter(cases == 0) %>%
  nrow()
```

So it appears that either a country has all its values in a year as non-missing if the WHO collected data for that country, or all its values are non-missing.
So it is okay to treat explicitly and implicitly missing values the same, and we don't lose any information by dropping them.
```{r}
gather(who, new_sp_m014:newrel_f65, key = "key", value = "cases") %>%
  group_by(country, year) %>%
  mutate(missing = is.na(cases)) %>%
  select(country, year, missing) %>%
  distinct() %>%
  group_by(country, year) %>%
  filter(n() > 1)
```

</div>

### Exercise <span class="exercise-number">12.6.2</span> {.unnumbered .exercise}

<div class="question">

What happens if you neglect the `mutate()` step? 
(`mutate(key = stringr::str_replace(key, "newrel", "new_rel")`)

</div>

<div class="answer">

The `separate()` function emits the warning "too few values".
If we check the rows for keys beginning with `"newrel_"`, we see that `sexage` is missing,
and `type = m014`.

```{r}
who3a <- who1 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")

filter(who3a, new == "newrel") %>% head()
```

</div>

### Exercise <span class="exercise-number">12.6.3</span> {.unnumbered .exercise}

<div class="question">

I claimed that `iso2` and `iso3` were redundant with country. 
Confirm this claim.

</div>

<div class="answer">

If `iso2` and `iso3` are redundant with `country`, then, within each country, 
there should only be one distinct combination of `iso2` and `iso3` values, which is the case.
```{r}
select(who3, country, iso2, iso3) %>%
  distinct() %>%
  group_by(country) %>%
  filter(n() > 1)
```


This makes sense, since `iso2` and `iso3` contain the 2- and 3-letter country abbreviations for the country.
The `iso2` variable contains each country's [ISO 3166 alpha-2](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2), and the `iso3` variable contains the each country's [ISO 3166 alpha-3](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) abbreviation.
You may recognize the ISO 3166-2 abbreviations, since they are almost identical to internet [country-code top level domains](https://en.wikipedia.org/wiki/Country_code_top-level_domain), such as `.uk`.

</div>

### Exercise <span class="exercise-number">12.6.4</span> {.unnumbered .exercise}

<div class="question">

For each country, year, and sex compute the total number of cases of TB.
Make an informative visualization of the data.

</div>

<div class="answer">

```{r}
who5 %>%
  group_by(country, year, sex) %>%
  filter(year > 1995) %>%
  summarise(cases = sum(cases)) %>%
  unite(country_sex, country, sex, remove = FALSE) %>%
  ggplot(aes(x = year, y = cases, group = country_sex, colour = sex)) +
  geom_line()
```

A small multiples plot faceting by country is difficult given the number of countries.
Focusing on those countries with the largest changes or absolute magnitudes after providing the context above is another option.

</div>

## Non-Tidy Data

`r no_exercises()`

[ex-12.2.2]: It would be better to join these tables usin the methods covered in the [Relational Data](https://r4ds.had.co.nz/relational-data.html). 
             We could use `inner_join(t2_cases, t2_population, by = c("country", "year"))`.

[non-syntactic]: See the [Creating Tibbles](https://r4ds.had.co.nz/tibbles.html#tibbles) section.
