# Tidy Data

## Introduction

```{r message=FALSE}
library(tidyverse)
```

## Tidy data

### Exercise <span class="exercise-number">12.2.1.1</span> {.unnumbered .exercise}

<div class="question">
Using prose, describe how the variables and observations are organized in each of the sample tables.
</div>

<div class="answer">

In `table1` each row is a (country, year) with variables `cases` and `population`.
```{r}
table1
```

In `table2`, each row is country, year , variable ("cases", "population") combination, and there is a `count` variable with the numeric value of the combination.
```{r}
table2
```

In `table3`, each row is a (country, year) combination with the column `rate` having the rate of cases to population as a character string in the format `"cases/rate"`.
```{r}
table3
```

Table 4 is split into two tables, one table for each variable: `table4a` is the table for cases, while `table4b` is the table for population. Within each table, each row is a country, each column is a year, and the cells are the value of the variable for the table.
```{r}
table4a
```
```{r}
table4b
```

</div>

### Exercise <span class="exercise-number">12.2.1.2</span> {.unnumbered .exercise}

<div class="question">
Compute the `rate` for `table2`, and `table4a` + `table4b`. You will need to perform four operations:

1.  Extract the number of TB cases per country per year.
1.  Extract the matching population per country per year.
1.  Divide cases by population, and multiply by 10000.
1.  Store back in the appropriate place.

Which representation is easiest to work with? Which is hardest? Why?

</div>

<div class="answer">

In order to calculate cases per person, we need to divide cases by population for each country, year.
This is easiest if country and population are two columns in a data frame with country and year rows.

For Table 2, we need to first create separate tables
for cases and population and ensure that they are sorted in the same order.
```{r}
t2_cases <- filter(table2, type == "cases") %>%
  rename(cases = count) %>%
  arrange(country, year)
t2_population <- filter(table2, type == "population") %>%
  rename(population = count) %>%
  arrange(country, year)
```
Calculate the cases per capita in a separate data frame.
```{r}
t2_cases_per_cap <- t2_cases %>%
  mutate(population = t2_population$population,
         cases_per_cap = (cases / population) * 10000) %>%
  select(country, year, cases_per_cap)
```
Since the question asks us to store it back in the appropriate location, we will add new rows with
`type = "cases_per_cap"` to `table2` and then
sort by country, year, and variable type as in the original table.
```{r}
t2_cases_per_cap <- t2_cases_per_cap %>%
  mutate(type = "cases_per_cap") %>%
  rename(count = cases_per_cap)
```
```{r}
bind_rows(table2, t2_cases_per_cap) %>%
  arrange(country, year, type, count)
```
Note that after adding the `cases_per_cap` rows,
the type of `count` is coerced to `numeric` (double) because `cases_per_cap` is not an integer.

For `table4a` and `table4b`, we will create a separate table for cases per capita (`table4c`), with country rows and year columns.
```{r}
table4c <-
  tibble(country = table4a$country,
         `1999` = table4a[["1999"]] / table4b[["1999"]] * 10000,
       `2000` = table4a[["2000"]] / table4b[["2000"]] * 10000)
table4c
```

Neither table is particularly easy to work with.
Since `table2` has separate rows for cases and population we needed to generate a table with columns for cases and population where we could
calculate cases per capita.
`table4a` and `table4b` split the cases and population variables into different tables which
made it easy to divide cases by population.
However, we had to repeat this calculation for each row.

The ideal format of a data frame to answer this question is one with columns `country`, `year`, `cases`, and `population`.
Then problem could be answered with a single `mutate()` call.

</div>

### Exercise <span class="exercise-number">12.2.1.3</span> {.unnumbered .exercise}

<div class="question">
Recreate the plot showing change in cases over time using `table2` instead of `table1`. What do you need to do first?
</div>

<div class="answer">

Before creating the plot with change in cases over time, we need to filter the data frame to only include rows representing cases of TB.
```{r}
table2 %>%
  filter(type == "cases") %>%
  ggplot(aes(year, count)) +
  geom_line(aes(group = country), colour = "grey50") +
  geom_point(aes(colour = country)) +
  scale_x_continuous(breaks = unique(table2$year)) +
  ylab("cases")

```

</div>

## Spreading and Gathering

This code is reproduced from the chapter because it is needed by the exercises:
```{r}
tidy4a <- table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases")
tidy4b <- table4b %>%
  gather(`1999`, `2000`, key = "year", value = "cases")
```

### Exercise <span class="exercise-number">12.3.3.1</span> {.unnumbered .exercise}

<div class="question">
Why are `gather()` and `spread()` not perfectly symmetrical?
Carefully consider the following example:

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>%
  spread(year, return) %>%
  gather(`2015`:`2016`, key = "year", value = "return")
```

</div>

<div class="answer">

The functions `spread()` and `gather()` are not perfectly symmetrical because column type information is not transferred between them.
When we use `gather()` on a data frame, it throws away all the information about the original column types. 
Additionally, it has to coerce all the variables that are being gathered into 
a single type, since they are going into a single vector.
Later if we `spread()` that data frame, the `spread()` function has no way to 
know what the original data types of the columns that were earlier gathered.

In this example, in the original table, the column `year` was numeric.
After running `spread()` and `gather()` it is a character vector.
Variable names are always converted to a character vector by `gather()`.

The functions `spread()` and `gather()` can be closer to symmetrical if we use the `convert` argument. It will try to convert character vectors to the appropriate type using `type.convert()`.
```{r}
stocks %>%
  spread(key = "year", value = "return") %>%
  gather(`2015`:`2016`, key = "year", value = "return", convert = TRUE)
```
However, since `convert = TRUE` is guessing the appropriate type it still may not work.

</div>

### Exercise <span class="exercise-number">12.3.3.2</span> {.unnumbered .exercise}

<div class="question">
Why does this code fail?

```{r error=TRUE}
table4a %>%
  gather(1999, 2000, key = "year", value = "cases")
```
</div>

<div class="answer">

The code fails because the column names `1999` and `2000` are not standard and thus needs to be quoted.
The tidyverse functions will interpret `1999` and `2000` without quotes as looking for the 1999th and 2000th column of the data frame.
This will work:
```{r}
table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases")
```

</div>

### Exercise <span class="exercise-number">12.3.3.3</span> {.unnumbered .exercise}

<div class="question">
Why does spreading this tibble fail? How could you add a new column to fix the problem?

```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
glimpse(people)
```

```{r error=TRUE}
spread(people, key, value)
```

</div>

<div class="answer">

Spreading the data frame fails because there are two rows with "age" for "Phillip Woods".
If we added another column with an indicator for the number observation it is,
the code will work.

```{r}
people <- tribble(
  ~name,             ~key,    ~value, ~obs,
  #-----------------|--------|------|------
  "Phillip Woods",   "age",       45, 1,
  "Phillip Woods",   "height",   186, 1,
  "Phillip Woods",   "age",       50, 2,
  "Jessica Cordero", "age",       37, 1,
  "Jessica Cordero", "height",   156, 1
)
spread(people, key, value)
```

</div>

### Exercise <span class="exercise-number">12.3.3.4</span> {.unnumbered .exercise}

<div class="question">
Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

</div>

<div class="answer">

To tidy `preg`, we need to use `gather()`. The variables in this data are

-   `sex` ("female", "male")
-   `pregnant` ("yes", "no")
-   `count`, which is a non-negative integer representing the number of observations.

The observations in this data are unique combinations of sex and pregnancy status.

```{r}
preg_tidy <- preg %>%
  gather(male, female, key = "sex", value = "count", na.rm = TRUE)
preg_tidy
```

However, we should consider the missing value in the male, non-pregnant row.
Is it missing due to missing data, or missing due to structural reasons? 
This will be discussed in the upcoming section on [Missing Values](http://r4ds.had.co.nz/tidy-data.html#missing-values-3).
Supposing that these data represent observations from a species in which it is
impossible for males to get pregnant (not seahorses), then that missing value
is structural. In the non-tidy data frame we had to include that structural missing value explicitly with an `NA` entry.
However, in the tidy version we can drop that row since it is an impossible 
combination.
We can do that by adding the argument `na.rm = TRUE` to `gather()`.
```{r}
preg_tidy2 <- preg %>%
  gather(male, female, key = "sex", value = "count", na.rm = TRUE)
preg_tidy2
```

Though not necessary, there is one more way in which we can improve this data.
If a variable takes two values, like `pregnant` and `sex` do, it is often preferable to store them as logical vectors.
The following vector uses logical vectors ro represent pregnancy and sex.
```{r}
preg_tidy3 <- preg_tidy2 %>%
  mutate(female = sex == "female",
         pregnant = pregnant == "yes") %>%
  select(female, pregnant, count)
preg_tidy3
```
In the previous data frame, I named the logical variable representing the sex `female`, not `sex`.
This makes the meaning of the variable self-documenting.
If the variable were named `sex` with values `TRUE` and `FALSE`, without reading the documentation, we wouldn't know whether `TRUE` means male or female.

Apart from some minor memory savings, representing these variables as logical
vectors results in more clear and concise code.
Compare the `filter()` calls to select non-pregnant females from `preg_tidy2` and `preg_tidy`.
```{r}
filter(preg_tidy2, sex == "female", pregnant == "no")
filter(preg_tidy3, female, !pregnant)
```

</div>

## Separating and Uniting

### Exercise <span class="exercise-number">12.4.3.1</span> {.unnumbered .exercise}

<div class="question">
What do the extra and fill arguments do in `separate()`? Experiment with the various  options for the following two toy datasets.
</div>

<div class="answer">

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))
```

The `extra` argument tells `separate()` what to do if there are too many pieces,
and the `fill` argument if there aren't enough.

```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"))
```
By default, `separate()` drops the extra values with a warning.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "drop")
```
This produces the same result as above, dropping extra values, but without the warning.
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>%
  separate(x, c("one", "two", "three"), extra = "merge")
```
In this, the extra values are not split, so `"f,g"` appears in column three.

In this, one of the entries for column, `"d,e"`, has too few elements.
The default for `fill` is similar to those in separate `separate()`; it fills with missing values but emits a warning. In this, row 2 of column "three", is `NA`.
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"))
```

Alternative options for `fill` are `"right"`, to fill with missing values from the right, but without a warning
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"), fill = "right")
```
The option `fill = "left"` also fills with missing values without a warning, but this time from the left side.
Now, column "one" of row 2 will be missing, and the other values in that row are shifted over.
```{r}
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>%
  separate(x, c("one", "two", "three"), fill = "left")
```

</div>

### Exercise <span class="exercise-number">12.4.3.2</span> {.unnumbered .exercise}

<div class="question">
Both `unite()` and `separate()` have a remove argument. What does it do? Why would you set it to `FALSE`?
</div>

<div class="answer">

You would set it to `FALSE` if you want to create a new variable, but keep the old one.

</div>

### Exercise <span class="exercise-number">12.4.3.3</span> {.unnumbered .exercise}

<div class="question">
Compare and contrast `separate()` and `extract()`, Why are there three variations of separation (by position, by separator, and with groups), but only one unite?
</div>

<div class="answer">

The function `separate()`, splits columns a column into multiple groups using
by separator, if the `sep` argument is a character vector, or character positions, if `sep` is numeric.
```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  separate(x, c("variable", "into"), sep = "_")

# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  separate(x, c("variable", "into"), sep = c(1))
```

The function `extract()` uses a regular expression to specify groups in character vector and split that single character vector into multiple columns.
This is more flexible than `separate()` because it does not require a common
separator or specific column positions.

```{r}
# example with separators
tibble(x = c("X_1", "X_2", "AA_1", "AA_2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])_([0-9])")

# example with position
tibble(x = c("X1", "X2", "Y1", "Y2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z])([0-9])")

# example that separate could not parse
tibble(x = c("X1", "X20", "AA11", "AA2")) %>%
  extract(x, c("variable", "id"), regex = "([A-Z]+)([0-9]+)")  

```

Both `separate()` and `extract()` convert a single column to many columns.
However, `unite()` converts many columns to one, with a choice of a separator to include between column values.

```{r}
tibble(variable = c("X", "X", "Y", "Y"), id = c(1, 2, 1, 2)) %>%
  unite(x, variable, id, sep = "_")

```

In other words, with `extract()` and `separate()` only one column can be chosen,
but there are many choices how to split that single column into different columns.
With `unite()`, there are many choices as to which columns to include, but only
choice as to how to combine their contents into a single vector.

</div>

## Missing Values

### Exercise <span class="exercise-number">12.5.1.1</span> {.unnumbered .exercise}

<div class="question">
Compare and contrast the `fill` arguments to `spread()` and `complete()`.
</div>

<div class="answer">

In `spread()`, the fill argument explicitly sets the value to replace `NA`s.
In `complete()`, the fill argument also sets a value to replace `NA`s but it is named list, allowing for different values for different variables.
Also, both cases replace both implicit and explicit missing values.

</div>

### Exercise <span class="exercise-number">12.5.1.2</span> {.unnumbered .exercise}

<div class="question">
What does the direction argument to `fill()` do?
</div>

<div class="answer">

With `fill`, it determines whether `NA` values should be replaced by the previous non-missing value (`"down"`) or the next non-missing value (`"up"`).

</div>

## Case Study

This code is repeated from the chapter because it is needed by the exercises.

```{r}
who1 <- who %>%
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm = TRUE)
glimpse(who1)
```

```{r}
who2 <- who1 %>%
 mutate(key = stringr::str_replace(key, "newrel", "new_rel"))

```

```{r}
who3 <- who2 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```

```{r}
who3 %>%
  count(new)
```

```{r}
who4 <- who3 %>%
  select(-new, -iso2, -iso3)
```

```{r}
who5 <- who4 %>%
  separate(sexage, c("sex", "age"), sep = 1)
who5
```

### Exercise <span class="exercise-number">12.6.1.1</span> {.unnumbered .exercise}

<div class="question">
In this case study I set `na.rm = TRUE` just to make it easier to check that we had the correct values. Is this reasonable? Think about how missing values are represented in this dataset. Are there implicit missing values? What’s the difference between an `NA` and zero?
</div>

<div class="answer">

Perhaps? I would need to know more about the data generation process.
There are zero's in the data, which means they may explicitly be indicating no cases.
```{r}
who1 %>%
  filter(cases == 0) %>%
  nrow()
```

So it appears that either a country has all its values in a year as non-missing if the WHO collected data for that country, or all its values are non-missing.
So it is okay to treat explicitly and implicitly missing values the same, and we don't lose any information by dropping them.
```{r}
gather(who, new_sp_m014:newrel_f65, key = "key", value = "cases") %>%
  group_by(country, year)  %>%
  mutate(missing = is.na(cases)) %>%
  select(country, year, missing) %>%
  distinct() %>%
  group_by(country, year) %>%
  filter(n() > 1)

```

</div>

### Exercise <span class="exercise-number">12.6.1.2</span> {.unnumbered .exercise}

<div class="question">
What happens if you neglect the `mutate()` step? (`mutate(key = stringr::str_replace(key, "newrel", "new_rel")`)
</div>

<div class="answer">

The `separate()` function emits the warning "too few values".
If we check the rows for keys beginning with `"newrel_"`, we see that `sexage` is missing,
and `type = m014`.

```{r}
who3a <- who1 %>%
  separate(key, c("new", "type", "sexage"), sep = "_")

filter(who3a, new == "newrel") %>% head()
```

</div>

### Exercise <span class="exercise-number">12.6.1.3</span> {.unnumbered .exercise}

<div class="question">
I claimed that `iso2` and `iso3` were redundant with country. Confirm this claim.
</div>

<div class="answer">

```{r}
select(who3, country, iso2, iso3) %>%
  distinct() %>%
  group_by(country) %>%
  filter(n() > 1)

```

</div>

### Exercise <span class="exercise-number">12.6.1.4</span> {.unnumbered .exercise}

<div class="question">
For each country, year, and sex compute the total number of cases of TB. Make an informative visualization of the data.
</div>

<div class="answer">

```{r}
who5 %>%
  group_by(country, year, sex) %>%
  filter(year > 1995) %>%
  summarise(cases = sum(cases)) %>%
  unite(country_sex, country, sex, remove = FALSE) %>%
  ggplot(aes(x = year, y = cases, group = country_sex, colour = sex)) +
  geom_line()

```

A small multiples plot faceting by country is difficult given the number of countries.
Focusing on those countries with the largest changes or absolute magnitudes after providing the context above is another option.

</div>

## Non-Tidy Data

No exercises
