---
output: html_document
editor_options:
  chunk_output_type: console
---
# Data Transformation

## Introduction

```{r message=FALSE,cache=FALSE}
library("nycflights13")
library("tidyverse")
```

## Filter rows with `filter()`

### Exercise 1 {.exercise}

`r BeginQuestion()`
Find all flights that

1.  Had an arrival delay of two or more hours
1.  Flew to Houston (IAH or HOU)
1.  Were operated by United, American, or Delta
1.  Departed in summer (July, August, and September)
1.  Arrived more than two hours late, but didn’t leave late
1.  Were delayed by at least an hour, but made up over 30 minutes in flight
1.  Departed between midnight and 6am (inclusive)

`r EndQuestion()`

`r BeginAnswer()`

*Had an arrival delay of two or more hours:* Since delay is in minutes, find
flights whose arrival was delayed more than 120 minutes.
```{r}
filter(flights, arr_delay > 120)
```

*Flew to Houston (IAH or HOU)*:
```{r}
filter(flights, dest %in% c("IAH", "HOU"))
```

*Were operated by United, American, or Delta* In the `flights` dataset,
the column `carrier` indicates the airline, but it uses two-character carrier codes.
We can find the carrier codes for the airlines in the `airlines` dataset.
Since the carrier code dataset only has `r nrow(airlines)` rows, and the names
of the airlines in that dataset are not exactly "United", "American", or "Delta",
it is easiest to manually lookup their carrier codes in that data.
```{r}
airlines
```
The carrier code for Delta is `"DL"`, for American is `"AA"`, and for United is `"UA"`.
Using these carriers codes, we check whether `carrier` is one of those.
```{r}
filter(flights, carrier %in% c("AA", "DL", "UA"))
```

*Departed in summer (July, August, and September)* The variable `month` has the month, and it is numeric.
```{r}
filter(flights, month >= 7, month <= 9)
```

*Arrived more than two hours late, but didn’t leave late*
```{r}
filter(flights, !is.na(dep_delay), dep_delay <= 0, arr_delay > 120)
```

*Were delayed by at least an hour, but made up over 30 minutes in flight*
```{r}
filter(flights, !is.na(dep_delay),
       dep_delay >= 60, dep_delay - arr_delay > 30)
```

*Departed between midnight and 6am (inclusive)*. Note that in `dep_time`, midnight is `2400`, not `0`.
```{r}
filter(flights, dep_time <= 600 | dep_time == 2400)
```

We could avoid using an "or" statement by using the [modulo operator](https://en.wikipedia.org/wiki/Modulo_operation), `%%`. The modulo operator returns the remainder of division.
What happens if we calculate `dep_time` mod `2400`? 
```{r}
c(2400, 600) %% 2400
```
Since `2400 %% 2400 == 0` and all other times are left unchanged, we will only need to compare the result of the modulo operation to `600`,
```{r}
filter(flights, dep_time %% 2400 <= 600)
```
This filter expression is more compact, but may or may not be more readable and 
intuitive to a reader.

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Another useful **dplyr** filtering helper is `between()`. What does it do? Can you use it to simplify the code needed to answer the previous challenges?
`r EndQuestion()`

`r BeginAnswer()`

The expression `between(x, left, right)` is equivalent to `x >= left & x <= right`.

Of the answers in the previous question, we could simplify the statement of *departed in summer* (`month >= 7 & month <= 9`) using `between()` as the following
```{r}
filter(flights, between(month, 7, 9))
```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?
`r EndQuestion()`

`r BeginAnswer()`

```{r}
filter(flights, is.na(dep_time))
```

Since `arr_time` is also missing, these are canceled flights.

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
Why is `NA ^ 0` not missing? Why is `NA | TRUE` not missing? Why is `FALSE & NA` not missing? Can you figure out the general rule? (`NA * 0` is a tricky counterexample!)
`r EndQuestion()`

`r BeginAnswer()`

`NA ^ 0 == 1` since for all numeric values $x ^ 0 = 1$.
```{r}
NA ^ 0
```

`NA | TRUE` is `TRUE` because the it doesn't matter whether the missing value is `TRUE` or `FALSE`, `x \lor T = T` for all values of `x`.
```{r}
NA | TRUE
```
Likewise, anything and `FALSE` is always `FALSE`.
```{r}
NA & FALSE
```
Because the value of the missing element matters in `NA | FALSE` and `NA & TRUE`, these are missing:
```{r}
NA | FALSE
NA & TRUE
```

Wut?? Since `x * 0 = 0` for all $x$ (except `Inf`) we might expect `NA * 0 = 0`, but that's not the case.
```{r}
NA * 0
```
The reason that `NA * 0` is not equal to `0` is that `x * 0 = NaN` is undefined when `x = Inf` or `x = -Inf`.
```{r}
Inf * 0
-Inf * 0
```

`r EndAnswer()`

## Arrange rows with `arrange()`

### Exercise 1 {.exercise}

`r BeginQuestion()`
How could you use `arrange()` to sort all missing values to the start? (Hint: use `is.na()`).
`r EndQuestion()`

`r BeginAnswer()`

This sorts by increasing `dep_time`, but with all missing values put first.
```{r}
arrange(flights, desc(is.na(dep_time)), dep_time)
```

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Sort flights to find the most delayed flights. Find the flights that left earliest.
`r EndQuestion()`

`r BeginAnswer()`

The most delayed flights are found by sorting by `dep_delay` in descending order.
```{r}
arrange(flights, desc(dep_delay))
```
If we sort `dep_delay` in ascending order, we get those that left earliest.
There was a flight that left 43 minutes early.
```{r}
arrange(flights, dep_delay)
```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
Sort flights to find the fastest flights.
`r EndQuestion()`

`r BeginAnswer()`

I assume that by by "fastest flights" it means the flights with the minimum air time.
So I sort by `air_time`. The fastest flights. The fastest flights area couple of flights between EWR and BDL with an air time of 20 minutes.
```{r}
arrange(flights, air_time)
```

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
Which flights traveled the longest? Which traveled the shortest?
`r EndQuestion()`

`r BeginAnswer()`

I'll assume hat traveled the longest or shortest refers to distance, rather than air-time.

The longest flights are the Hawaii Air (HA 51) between JFK and HNL (Honolulu) at 4,983 miles.
```{r}
arrange(flights, desc(distance))
```

Apart from an EWR to LGA flight that was canceled, the shortest flights are the Envoy Air Flights between EWR and PHL at 80 miles.
```{r}
arrange(flights, distance)
```

`r EndAnswer()`

## Select columns with `select()`

### Exercise 1 {.exercise}

`r BeginQuestion()`
Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.
`r EndQuestion()`

`r BeginAnswer()`

A few ways include:
```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, starts_with("dep_"), starts_with("arr_"))
select(flights, matches("^(dep|arr)_(time|delay)$"))
```
using `ends_with()` doesn't work well since it would return both `sched_arr_time` and `sched_dep_time`.

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
What happens if you include the name of a variable multiple times in a select() call?
`r EndQuestion()`

`r BeginAnswer()`

It ignores the duplicates, and that variable is only included once. No error, warning, or message is emitted.
```{r}
select(flights, year, month, day, year, year)
```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
What does the `one_of()` function do? Why might it be helpful in conjunction with this vector?
`r EndQuestion()`

`r BeginAnswer()`

The `one_of` vector allows you to select variables with a character vector rather than as unquoted variable names.
It's useful because then you can easily pass vectors to `select()`.

```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
```

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
`r EndQuestion()`

`r BeginAnswer()`

```{r}
select(flights, contains("TIME"))
```

The default behavior for contains is to ignore case.
Yes, it surprises me.
Upon reflection, I realized that this is likely the default behavior because `dplyr` is designed to deal with a variety of data backends, and some database engines don't differentiate case.

To change the behavior add the argument `ignore.case = FALSE`. Now no variables are selected.
```{r}
select(flights, contains("TIME", ignore.case = FALSE))
```

`r EndAnswer()`

## Add new variables with `mutate()`

### Exercise 1 {.exercise}

`r BeginQuestion()`
Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
`r EndQuestion()`

`r BeginAnswer()`

To get the departure times in the number of minutes, (integer) divide `dep_time` by 100 to get the hours since midnight and multiply by 60 and add the remainder of `dep_time` divided by 100.
```{r}
mutate(flights,
       dep_time_mins = dep_time %/% 100 * 60 + dep_time %% 100,
       sched_dep_time_mins = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %>%
  select(dep_time, dep_time_mins, sched_dep_time, sched_dep_time_mins)
```

This would be more cleanly done by first defining a function and reusing that:
```{r}
time2mins <- function(x) {
  x %/% 100 * 60 + x %% 100
}
mutate(flights,
       dep_time_mins = time2mins(dep_time),
       sched_dep_time_mins = time2mins(sched_dep_time)) %>%
  select(dep_time, dep_time_mins, sched_dep_time, sched_dep_time_mins)
```

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Compare `air_time` with `arr_time - dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?
`r EndQuestion()`

`r BeginAnswer()`

Since `arr_time` and `dep_time` may be in different time zones, the `air_time` doesn't equal the difference.
We would need to account for time-zones in these calculations.
```{r}
mutate(flights,
       air_time2 = arr_time - dep_time,
       air_time_diff = air_time2 - air_time) %>%
  filter(air_time_diff != 0) %>%
  select(air_time, air_time2, dep_time, arr_time, dest)
```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?
`r EndQuestion()`

`r BeginAnswer()`

I'd expect `dep_time`, `sched_dep_time`, and `dep_delay` to be related so that `dep_time - sched_dep_time = dep_delay`.
```{r}
mutate(flights,
       dep_delay2 = dep_time - sched_dep_time) %>%
  filter(dep_delay2 != dep_delay) %>%
  select(dep_time, sched_dep_time, dep_delay, dep_delay2)
```
Oops, I forgot to convert to minutes. I'll reuse the `time2mins` function I wrote earlier.
```{r}
mutate(flights,
       dep_delay2 = time2mins(dep_time) - time2mins(sched_dep_time)) %>%
  filter(dep_delay2 != dep_delay) %>%
  select(dep_time, sched_dep_time, dep_delay, dep_delay2)
```
Well, that solved most of the problems, but these two numbers don't match because we aren't accounting for flights where the departure time is the next day from the scheduled departure time.

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for `min_rank()`.
`r EndQuestion()`

`r BeginAnswer()`

I'd want to handle ties by taking the minimum of tied values. If three flights are have the same value and are the most delayed, we would say they are tied for first, not tied for third or second.
```{r}
mutate(flights,
       dep_delay_rank = min_rank(-dep_delay)) %>%
  arrange(dep_delay_rank) %>%
  filter(dep_delay_rank <= 10)
```

`r EndAnswer()`

### Exercise 5 {.exercise}

`r BeginQuestion()`
What does `1:3 + 1:10` return? Why?
`r EndQuestion()`

`r BeginAnswer()`

It returns `c(1 + 1, 2 + 2, 3 + 3, 1 + 4, 2 + 5, 3 + 6, 1 + 7, 2 + 8, 3 + 9, 1 + 10)`.
When adding two vectors recycles the shorter vector's values to get vectors of the same length.
We get a warning vector since the shorter vector is not a multiple of the longer one (this often, but not necessarily, means we made an error somewhere).

```{r}
1:3 + 1:10
```

`r EndAnswer()`

### Exercise 6 {.exercise}

`r BeginQuestion()`
What trigonometric functions does R provide?
`r EndQuestion()`

`r BeginAnswer()`

These are all described in the same help page,
```{r eval=FALSE}
help("Trig")
```

Cosine (`cos`), sine (`sin`), tangent (`tan`) are provided:
```{r}
tibble(
  x = seq(-3, 7, by = 1 / 2),
  cosx = cos(pi * x),
  sinx = cos(pi * x),
  tanx = tan(pi * x)
)
```
The convenience function `cospi(x)` is equivalent to `cos(pi * x)`, with `sinpi` and `tanpi` similarly defined,
```{r}
tibble(
  x = seq(-3, 7, by = 1 / 2),
  cosx = cospi(x),
  sinx = cos(x),
  tanx = tan(x)
)
```

The inverse function arc-cosine (`acos`), arc-sine (`asin`), and arc-tangent (`atan`) are provided,
```{r}
tibble(
  x = seq(-1, 1, by = 1 / 4),
  acosx = acos(x),
  asinx = asin(x),
  atanx = atan(x)
)
```

The function `atan2` is the angle between the x-axis and the the vector (0,0) to (`x`, `y`).
```{r}
atan2(c(1, 0, -1, 0), c(0, 1, 0, -1))
```

`r EndAnswer()`

## Grouped summaries with `summarise()`

### Exercise 1 {.exercise}

`r BeginQuestion()`
Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios:

-   A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.
-   A flight is always 10 minutes late.
-   A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.
-   99% of the time a flight is on time. 1% of the time it’s 2 hours late.

Which is more important: arrival delay or departure delay?

`r EndQuestion()`

`r BeginAnswer()`

What this question gets at is a fundamental question of data analysis: the cost function. 
As analysts, the reason we are interested in flight delay because it is costly to passengers. 
But it is worth thinking carefully about how it is costly and use that information in ranking and measuring these scenarios.

In many scenarios, arrival delay is more important. 
Presumably being late on arriving is more costly to the passenger since it could disrupt the next stages of their travel, such as connecting flights or meetings.  
If the departure is delayed without affecting the arrival time and the passenger arrived at the same time, this delay will not affect future plans nor does it affect the total time spent traveling. 
The delay could be a positive, if less time is spent on the airplane itself, or a negative, if that extra time is spent on the plane in the runway.

Variation in arrival time is worse than consistency. 
If a flight is always 30 minutes late and that delay is know, then it is as if the arrival time is that delayed time.
The traveler could easily plan for this. If the delay of the flight is more variable, then it is harder for the traveler to plan for it.

**TODO** (Add a better explanation and some examples)

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Come up with another approach that will give you the same output as `not_cancelled %>% count(dest)` and `not_cancelled %>% count(tailnum, wt = distance)` (without using `count()`).
`r EndQuestion()`

`r BeginAnswer()`

The data frame `not_cancelled` is defined in the chapter as,
```{r not_cancelled}
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

Count will group a dataset on the given variable and then determine the number of instances within each group.
This can be done by by first grouping by the given variable, and then finding the number of observations in each group.
The number of observations in each group can be found by calling the `length()` function on any variable.
To make the result match `count()`, the value should go in a new column `n`.
```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarise(n = length(dest))
```
A more concise way to get the number of observations in a data frame, or a group, is the function `n()`,
```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarise(n = n())
```

For a weighted count, take the sum of the weight variable in each group.
```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  summarise(n = sum(distance))
```

Alternatively, we could have used `group_by` followed by `tally()`,
since `count()` itself is a shortcut for calling `group_by()` then `tally()`,
```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  tally()
```
and
```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  tally(distance)
```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
Our definition of canceled flights `(is.na(dep_delay) | is.na(arr_delay))` is slightly suboptimal. Why? Which is the most important column?
`r EndQuestion()`

`r BeginAnswer()`

If a flight never departs, then it won't arrive.
A flight could also depart and not arrive if it crashes, or if it is redirected and lands in an airport other than its intended destination.

The more important column is `arr_delay`, which indicates the amount of delay in arrival.
```{r}
filter(flights, !is.na(dep_delay), is.na(arr_delay)) %>%
  select(dep_time, arr_time, sched_arr_time, dep_delay, arr_delay)
```
Okay, I'm not sure what's going on in this data. `dep_time` can be non-missing and `arr_delay` missing but `arr_time` not missing.
They may be combining different flights?

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
Look at the number of canceled flights per day. Is there a pattern? Is the proportion of canceled flights related to the average delay?
`r EndQuestion()`

`r BeginAnswer()`

```{r}
canceled_delayed <-
  flights %>%
  mutate(canceled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(prop_canceled = mean(canceled),
            avg_dep_delay = mean(dep_delay, na.rm = TRUE))

ggplot(canceled_delayed, aes(x = avg_dep_delay, prop_canceled)) +
  geom_point() +
  geom_smooth()
```

`r EndAnswer()`

### Exercise 5 {.exercise}

`r BeginQuestion()`
Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights %>% group_by(carrier, dest) %>% summarise(n())`)
`r EndQuestion()`

`r BeginAnswer()`

```{r}
flights %>%
  group_by(carrier) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(arr_delay))
```

```{r}
filter(airlines, carrier == "F9")
```

Frontier Airlines (FL) has the worst delays.

You can get part of the way to disentangling the effects of airports vs. carriers by
comparing each flight's delay to the average delay of destination airport.
However, you'd really want to compare it to the average delay of the destination airport, *after* removing other flights from the same airline.

FiveThirtyEight conducted a [similar analysis](http://fivethirtyeight.com/features/the-best-and-worst-airlines-airports-and-flights-summer-2015-update/).

`r EndAnswer()`

### Exercise 6 {.exercise}

`r BeginQuestion()`
For each plane, count the number of flights before the first delay of greater than 1 hour.
`r EndQuestion()`

`r BeginAnswer()`

I think this requires grouped mutate (but I may be wrong):
```{r}
flights %>%
  arrange(tailnum, year, month, day) %>%
  group_by(tailnum) %>%
  mutate(delay_gt1hr = dep_delay > 60) %>%
  mutate(before_delay = cumsum(delay_gt1hr)) %>%
  filter(before_delay < 1) %>%
  count(sort = TRUE)
```

`r EndAnswer()`

### Exercise 7 {.exercise}

`r BeginQuestion()`
What does the sort argument to `count()` do. When might you use it?
`r EndQuestion()`

`r BeginAnswer()`

The sort argument to `count` sorts the results in order of `n`.
You could use this anytime you would do `count` followed by `arrange`.

`r EndAnswer()`

## Grouped mutates (and filters)

### Exercise 1 {.exercise}

`r BeginQuestion()`
Refer back to the table of useful mutate and filtering functions. Describe how each operation changes when you combine it with grouping.
`r EndQuestion()`

`r BeginAnswer()`

They operate within each group rather than over the entire data frame. E.g. `mean` will calculate the mean within each group.

`r EndAnswer()`

### Exercise 2 {.exercise}

`r BeginQuestion()`
Which plane (`tailnum`) has the worst on-time record?
`r EndQuestion()`

`r BeginAnswer()`

```{r}
flights %>%
  group_by(tailnum) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  filter(rank(desc(arr_delay)) <= 1)

```

`r EndAnswer()`

### Exercise 3 {.exercise}

`r BeginQuestion()`
What time of day should you fly if you want to avoid delays as much as possible?
`r EndQuestion()`

`r BeginAnswer()`

Let's group by hour. The earlier the better to fly. This is intuitive as delays early in the morning are likely to propagate throughout the day.
```{r}
flights %>%
  group_by(hour) %>%
  summarise(arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(arr_delay)

```

`r EndAnswer()`

### Exercise 4 {.exercise}

`r BeginQuestion()`
For each destination, compute the total minutes of delay. For each, flight, compute the proportion of the total delay for its destination.
`r EndQuestion()`

`r BeginAnswer()`

```{r}
flights %>%
  filter(!is.na(arr_delay), arr_delay > 0) %>%  
  group_by(dest) %>%
  mutate(total_delay = sum(arr_delay),
         prop_delay = arr_delay / total_delay) %>%
  select(
    (year:day),
    dep_delay,
    total_delay,
    prop_delay
  )
```

Alternatively, consider the delay as relative to the *minimum* delay for any flight to that destination. Now all non-canceled flights have a proportion.
```{r}
flights %>%
  filter(!is.na(arr_delay), arr_delay > 0) %>%  
  group_by(dest) %>%
  mutate(total_delay = sum(arr_delay - min(arr_delay)),
         prop_delay = arr_delay / sum(arr_delay)) %>%
  select(
    (year:day),
    dep_delay,
    total_delay,
    prop_delay
  )
```

`r EndAnswer()`

### Exercise 5 {.exercise}

`r BeginQuestion()`
Delays are typically temporally correlated: even once the problem that caused the initial delay has been resolved, later flights are delayed to allow earlier flights to leave. Using `lag()` explore how the delay of a flight is related to the delay of the immediately preceding flight.
`r EndQuestion()`

`r BeginAnswer()`

We want to group by day to avoid taking the lag from the previous day.
Also, I want to use departure delay, since this mechanism is relevant for departures.
Also, I remove missing values both before and after calculating the lag delay.
However, it would be interesting to ask the probability or average delay after a cancellation.
```{r}
flights %>%
  group_by(year, month, day) %>%
  filter(!is.na(dep_delay)) %>%
  mutate(lag_delay = lag(dep_delay)) %>%
  filter(!is.na(lag_delay)) %>%
  ggplot(aes(x = dep_delay, y = lag_delay)) +
  geom_point() +
  geom_smooth()
```

`r EndAnswer()`

### Exercise 6 {.exercise}

`r BeginQuestion()`
Look at each destination. Can you find flights that are suspiciously fast? (i.e. flights that represent a potential data entry error). Compute the air time a flight relative to the shortest flight to that destination. Which flights were most delayed in the air?
`r EndQuestion()`

`r BeginAnswer()`

The shorter BOS and PHL flights that are 20 minutes for 30+ minutes flights seem plausible - though maybe entries of +/- a few minutes can easily create large changes.
I assume that departure time has a standardized definition, but I'm not sure; if there is some discretion, that could create errors that are small in absolute time, but large in relative time for small flights.
The ATL, GSP, and BNA flights look suspicious as they are almost half the time of longer flights.
```{r}
flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest) %>%
  mutate(med_time = median(air_time),
         fast = (air_time - med_time) / med_time) %>%
  arrange(fast) %>%
  select(air_time, med_time, fast, dep_time, sched_dep_time, arr_time, sched_arr_time) %>%
  head(15)
```

I could also try a z-score. Though the standard deviation and mean will be affected by large delays.
```{r}
flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest) %>%
  mutate(air_time_mean = mean(air_time),
         air_time_sd = sd(air_time),
         z_score = (air_time - air_time_mean) / air_time_sd) %>%
  arrange(z_score) %>%
  select(z_score, air_time_mean, air_time_sd, air_time, dep_time, sched_dep_time, arr_time, sched_arr_time)
```

```{r}
flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest) %>%
  mutate(air_time_diff = air_time - min(air_time)) %>%
  arrange(desc(air_time_diff)) %>%
  select(dest, year, month, day, carrier, flight, air_time_diff, air_time, dep_time, arr_time) %>%
  head()
```

`r EndAnswer()`

### Exercise 7 {.exercise}

`r BeginQuestion()`
Find all destinations that are flown by at least two carriers. Use that information to rank the carriers.
`r EndQuestion()`

`r BeginAnswer()`

The carrier that flies to the most locations is ExpressJet Airlines (EV).
ExpressJet is a regional airline and partner for major airlines, so its one of those that flies small planes to close airports

```{r}
flights %>%
  group_by(dest, carrier) %>%
  count(carrier) %>%
  group_by(carrier) %>%
  count(sort = TRUE)

```

```{r}
filter(airlines, carrier == "EV")
```

`r EndAnswer()`
