---
output: html_document
editor_options:
  chunk_output_type: console
---
# Data Visualisation

## Introduction

```{r, message=FALSE}
library("tidyverse")
```

No exercises.

## First Steps

### Exercise <span class="exercise-number">3.2.1</span>. {.unnumbered .exercise}

<div class="question">
Run `ggplot(data = mpg)` what do you see?
</div>

<div class="answer">

```{r}
ggplot(data = mpg)
```

An empty plot. The background of the plot is created by `ggplot()`, but nothing else is displayed.

</div>

### Exercise <span class="exercise-number">3.2.2</span>. {.unnumbered .exercise}

<div class="question">
How many rows are in `mtcars`? How many columns?
</div>

<div class="answer">

There are `r nrow(mtcars)` rows and `r ncol(mtcars)` columns in the `mtcars` data frame.
```{r}
nrow(mtcars)
ncol(mtcars)
```
The number of rows and columns is also displayed by `glimpse()`:
```{r}
glimpse(mtcars)
```

</div>

### Exercise <span class="exercise-number">3.2.3</span>. {.unnumbered .exercise}

<div class="question">
What does the `drv` variable describe? Read the help for `?mpg` to find out.
</div>

<div class="answer">

The `drv` categorizes cars by which wheels the engine provides torque to, or drives: the front two wheels, the rear two wheels, or all four wheels.[^layout]

| Value      | Description                                                                                   |
|------------|-----------------------------------------------------------------------------------------------|
| `"f"`      | [front-wheel drive](https://en.wikipedia.org/wiki/Front-wheel_drive)                          |
| `"r"`      | [rear-wheel drive](https://en.wikipedia.org/wiki/Automobile_layout#Rear-wheel-drive_layouts)  |
| `"4"`      | [four-wheel drive](https://en.wikipedia.org/wiki/Four-wheel_drive)                            |

[^layout]: See the Wikipedia article on [Automobile layout](https://en.wikipedia.org/wiki/Automobile_layout).

</div>

### Exercise <span class="exercise-number">3.2.4</span>. {.unnumbered .exercise}

<div class="question">
Make a scatter plot of `hwy` vs `cyl`.
</div>

<div class="answer">

```{r}
ggplot(mpg, aes(x = hwy, y = cyl)) +
  geom_point()
```

</div>

### Exercise <span class="exercise-number">3.2.5</span>. {.unnumbered .exercise}

<div class="question">
What happens if you make a scatter plot of `class` vs `drv`. Why is the plot not useful?
</div>

<div class="answer">
```{r}
ggplot(mpg, aes(x = class, y = drv)) +
  geom_point()
```

A scatter plot is not a useful way to plot these variables, since both `drv` and `class` are factor variables taking a limited number of values.
```{r}
count(mpg, drv, class)
```
The scatter plot cannot show which points are overlapping.

This is covered in a later section, but may be of interest. Section [7.5.2](https://r4ds.had.co.nz/exploratory-data-analysis.html#two-categorical-variables) discusses other methods for plotting two categorical variables.
The first is `geom_count()` which uses the size of the points to show the number of observations at that point.
```{r}
ggplot(mpg, aes(x = class, y = drv)) +
  geom_count()
```
The second is `geom_tile()` which uses the color aesthetic to show the number o observations at each point.
```{r}
mpg %>%
  count(class, drv) %>%
  ggplot(aes(x = class, y = drv)) +
    geom_tile(mapping = aes(fill = n))
```
Since there are many combinations of `class` and `drv` that have zero observations,
they are shown as missing.
The `complete()` function in **tidyr** adds new rows to a data frame for missing combinations of columns.
The following code add rows for missing combinations of `class` and `drv` and uses 
the `fill` argument to set `n = 0` for those new rows.

```{r}
mpg %>%
  count(class, drv) %>%
  complete(class, drv, fill = list(n = 0L)) %>%
  ggplot(aes(x = class, y = drv)) +
    geom_tile(mapping = aes(fill = n))
```

</div>

## Aesthetic mappings

### Exercise <span class="exercise-number">3.3.1</span> {.unnumbered .exercise}

<div class="question">
What’s gone wrong with this code? Why are the points not blue?

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = "blue"))
```
</div>

<div class="answer">

Since `colour = "blue"` was included within the `mapping` argument, it was treated as an aesthetic (a mapping between a variable and a value).
The expression, `color="blue"`, treats `"blue"` as a variable with only one value: `"blue"`. If this is confusing, consider how `colour = 1:234` or `colour = 1` would be interpreted by `aes()`.

</div>

### Exercise <span class="exercise-number">3.3.2</span> {.unnumbered .exercise}

<div class="question">

Which variables in `mpg` are categorical?
Which variables are continuous?
(Hint: type `?mpg` to read the documentation for the dataset).
How can you see this information when you run `mpg`?

</div>

<div class="answer">

```{r eval=FALSE}
?mpg
```

When printing the data frame, this information is given at the top of each column within angled brackets. Categorical variables have a class of "character" (`<chr>`).
```{r}
mpg
```
Alternatively, `glimpse()` displays the type of each column:
```{r}
glimpse(mpg)
```
</div>

### Exercise <span class="exercise-number">3.3.3</span> {.unnumbered .exercise}

<div class="question">
Map a continuous variable to color, size, and shape.
How do these aesthetics behave differently for categorical vs. continuous variables?
</div>

<div class="answer">

The variable `cty`, city highway miles per gallon, is a continuous variable:
```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = cty)) +
  geom_point()
```

Instead of using discrete colors, the continuous variable uses a scale that varies from a light to dark blue color.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = cty)) +
  geom_point()
```

When mapped to size, the sizes of the points vary continuously with respect to the size (although the legend shows a few representative values)

```{r error=TRUE}
ggplot(mpg, aes(x = displ, y = hwy, shape = cty)) +
  geom_point()
```

When a continuous value is mapped to shape, it gives an error.
Though we could split a continuous variable into discrete categories and use a shape aesthetic, this would conceptually not make sense.
A continuous numeric variable is ordered, but shapes have no natural order.
It is clear that smaller points correspond to smaller values, or once the color scale is given, which colors correspond to larger or smaller values. But it is not clear whether a square is greater or less than a circle.

</div>

### Exercise <span class="exercise-number">3.3.4</span> {.unnumbered .exercise}

<div class="question">
What happens if you map the same variable to multiple aesthetics?
</div>

<div class="answer">

```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = hwy, size = displ)) +
  geom_point()
```

In the above plot, `hwy` is mapped to both location on the y-axis and color, and `displ` is mapped to both location on the x-axis and size.
The code works and produces a plot, even if it is a bad one.
Mapping a single variable to multiple aesthetics is redundant.
Because it is redundant information, in most cases avoid mapping a single variable to multiple aesthetics.

</div>

### Exercise <span class="exercise-number">3.3.5</span> {.unnumbered .exercise}

<div class="question">
What does the stroke aesthetic do? What shapes does it work with? (Hint: use `?geom_point`)
</div>

<div class="answer">

Stroke changes the size of the border for shapes (21-25).
These are filled shapes in which the color and size of the border can differ from that of the filled interior of the shape.

For example
```{r ex.3.3.1.5}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(shape = 21, colour = "black", fill = "white", size = 5, stroke = 5)
```

</div>

### Exercise <span class="exercise-number">3.3.6</span>. {.unnumbered .exercise}

<div class="question">
What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`?
</div>

<div class="answer">

```{r ex.3.3.1.6}
ggplot(mpg, aes(x = displ, y = hwy, colour = displ < 5)) +
  geom_point()
```

Aesthetics can also be mapped to expressions (code like `displ < 5`).
It will create a temporary variable which takes values from  the result of the expression.
In this case, it is logical variable which is `TRUE` or `FALSE`.
This also explains exercise 1, `colour = "blue"` created a categorical variable that only had one category: "blue".

</div>

## Common problems

No exercises

## Facets

### Exercise <span class="exercise-number">3.5.1</span> {.unnumbered .exercise}

<div class="question">
What happens if you facet on a continuous variable?
</div>

<div class="answer">

Let's see.
```{r ex.3.5.1.1}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(. ~ cty)
```

It converts the continuous variable to a factor and creates facets for **all** unique values of it.

</div>

### Exercise <span class="exercise-number">3.5.2</span> {.unnumbered .exercise}

<div class="question">
What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean? How do they relate to this plot?
</div>

<div class="answer">

They are cells in which there are no values of the combination of `drv` and `cyl`.

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = cyl))
```

The locations in the above plot without points are the same cells in `facet_grid(drv ~ cyl)` that have no points.

</div>

### Exercise <span class="exercise-number">3.5.3</span> {.unnumbered .exercise}

<div class="question">
What plots does the following code make? What does `.` do?
</div>

<div class="answer">

The symbol `.` ignores that dimension for faceting.

This plot facets by values of `drv` on the y-axis:
```{r ex.3.5.1.4.a}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```
This plot facets by values of `cyl` on the x-axis:
```{r ex.3.5.1.4.b}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

</div>

### Exercise <span class="exercise-number">3.5.4</span> {.unnumbered .exercise}

<div class="question">
Take the first faceted plot in this section:

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~class, nrow = 2)
```
What are the advantages to using faceting instead of the colour aesthetic?
What are the disadvantages?
How might the balance change if you had a larger dataset?

</div>

<div class="answer">

This is what the plot looks like when `class` is represented by the colour
the color aesthetic instead of faceting.
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Advantages of encoding `class` with facets instead of color include the
ability to encode more distinct categories.
For me, it is difficult to distinguish color of `"midsize"` and the teal of `"minivan"` points are difficult to distinguish.
Given human visual perception, the max number of colors to use when encoding
unordered categorical (qualitative) data is nine, and in practice, often much less
than that.
Also, while placing points in different categories in different scales makes
it difficult to directly compare values of individual points in different categories, it can make it easier to compare patterns between categories.

Disadvantages of encoding `class` with facets instead of color are that different
the different class is that the points for each category are on different plots,
making it more difficult to directly compare the locations of individual points.
Using the same x- and y-scales for all facets lessens this disadvantage.
Since encoding class within color also places all points on the same plot,
it visualizes the unconditional relationship between the x and y variables;
with facets, the unconditional relationship is no longer visualized since the
points are spread across multiple plots.

The benefits encoding a variable through facetting over color become more advantageous as either the number of points or the number of categories increase.
In the former, as the number of points increase, there is likely to be more
overlap.

It is difficult to handle overlapping points with color.
Jittering will still work with color.
But jittering will only work well if there are few points and the classes do not overlap much, otherwise the colors of areas will no longer be distinct and it will be hard to visually pick out the patterns of different categories.
Transparency (`alpha`) does not work well with colors since the mixing of overlapping transparent colors will no longer represent the colors of the categories.
Binning methods use already color to encode density, so color cannot be used to encode categories.

As noted before, as the number of categories increases, the difference between
colors decreases, to the point that the color of categories will no longer be
visually distinct.

</div>

### Exercise <span class="exercise-number">3.5.5</span> {.unnumbered .exercise}

<div class="question">
Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? What other options control the layout of the individual panels? Why doesn’t `facet_grid()` have `nrow` and `ncol` variables?
</div>

<div class="answer">

The arguments `nrow` (`ncol`) determines the number of rows (columns) to use when laying out the facets.
It is necessary since `facet_wrap()` only facets on one variable.
These arguments are unnecessary for `facet_grid()` since the number of rows and columns are determined by the number of unique values of the variables specified.

</div>

### Exercise <span class="exercise-number">3.5.6</span> {.unnumbered .exercise}

<div class="question">
When using `facet_grid()` you should usually put the variable with more unique levels in the columns. Why?
</div>

<div class="answer">

IF the plot is laid out horizontally, there will be more space for columns.
You should put the variable with more unique levels in the columns if the plot is laid out landscape.
It is easier to compare relative levels of y by scanning horizontally, so it may be easier to visually compare these levels.

</div>

## Geometric Objects

### Exercise <span class="exercise-number">3.6.1</span> {.unnumbered .exercise}

<div class="question">
What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
</div>

<div class="answer">

-   line chart: `geom_line()`
-   boxplot: `geom_boxplot()`
-   histogram: `geom_hist()`
-   area chart: `geom_area()`

</div>

### Exercise <span class="exercise-number">3.6.2</span> {.unnumbered .exercise}

<div class="question">
Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
</div>

<div class="answer">

```{r eval=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

This will produce a scatter plot with `displ` on the x-axis, `hwy` on the y-axis.
The points will be colored by `drv`.
There will be a smooth line, without standard errors, fit through each `drv` group.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

</div>

### Exercise <span class="exercise-number">3.6.3</span> {.unnumbered .exercise}

<div class="question">
What does `show.legend = FALSE` do? What happens if you remove it?
Why do you think I used it earlier in the chapter?
</div>

<div class="answer">

Show legend hides the legend box. In this code, without show legend, there is a legend.
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, colour = drv),
  )
```
But there is no legend in this code:
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, colour = drv),
    show.legend = FALSE
  )
```

In the example earlier in the chapter,
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))

ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, colour = drv),
    show.legend = FALSE
  )
```
the legend is suppressed because there are three plots, and adding a legend that only appears in the last one would make the presentation asymmetric.
Additionally, the purpose of this plot is to illustrate the difference between not grouping, using a `group` aesthetic, and using a `color` aesthetic (with implicit grouping).
In that example, the legend isn't necessary since looking up the values associated with each color isn't necessary to make that point.

</div>

### Exercise <span class="exercise-number">3.6.4</span> {.unnumbered .exercise}

<div class="question">
What does the `se` argument to `geom_smooth()` do?
</div>

<div class="answer">

It adds standard error bands to the lines.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = TRUE)
```

By default `se = TRUE`:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth()
```

</div>

### Exercise <span class="exercise-number">3.6.5</span> {.unnumbered .exercise}

<div class="question">
Will these two graphs look different? Why/why not?
</div>

<div class="answer">

No. Because both `geom_point()` and `geom_smooth()` use the same data and mappings. They will inherit those options from the `ggplot()` object, and thus don't need to specified again (or twice).

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```

```{r}
ggplot() +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

</div>

### Exercise <span class="exercise-number">3.6.6</span> {.unnumbered .exercise}

<div class="question">
Recreate the R code necessary to generate the following graphs.
</div>

<div class="answer">

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(group = drv), se = FALSE) +
  geom_point()
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(se = FALSE)
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4, color = "white") +
  geom_point(aes(colour = drv))
```
</div>

## Statistical Transformations

### Exercise <span class="exercise-number">3.1</span> {.unnumbered .exercise}

<div class="question">
What is the default geom associated with `stat_summary()`? How could you rewrite the previous plot to use that geom function instead of the stat function?
</div>

<div class="answer">

The default geom for [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html) is `geom_pointrange()` (see the `stat`) argument.

But, the default `stat` for [`geom_pointrange()`](https://ggplot2.tidyverse.org/reference/geom_linerange.html) is `identity()`, so use `geom_pointrange(stat = "summary")`.
```{r}
ggplot(data = diamonds) +
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = "summary",
  )
```

The default message says that `stat_summary()` uses the `mean` and `sd` to calculate the point, and range of the line. So lets use the previous values of `fun.ymin`, `fun.ymax`, and `fun.y`:
```{r}
ggplot(data = diamonds) +
  geom_pointrange(
    mapping = aes(x = cut, y = depth),
    stat = "summary",
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )
```

</div>

### Exercise <span class="exercise-number">3.2</span>. {.unnumbered .exercise}

<div class="question">
What does `geom_col()` do? How is it different to `geom_bar()`?
</div>

<div class="answer">

The `geom_col()` function has different default than `geom_bar()`.
The default stat of `geom_col() is `identity()` stat.
This means that `geom_col()` expects that the data is already preprocessed into `x` values and `y` values representing the bar height.
The defult stat of `geom_bar()` is `count()`.
This means that `geom_bar()` expects the `x` variable to contain multiple observations for each values, and it will handle counting the number of observations
for each value of `x` in order to create the bar heights.

</div>

### Exercise <span class="exercise-number">3.3</span>. {.unnumbered .exercise}

<div class="question">
Most geoms and stats come in pairs that are almost always used in concert.
Read through the documentation and make a list of all the pairs.
What do they have in common?
</div>

<div class="answer">

See the [ggplot2 documentation](https://ggplot2.tidyverse.org/reference/).

**TODO**

</div>

### Exercise <span class="exercise-number">3.4</span>. {.unnumbered .exercise}

<div class="question">
What variables does `stat_smooth()` compute? What parameters control its behavior?
</div>

<div class="answer">

The function `stat_smooth()` calculates the following statistics:

-   `y`: predicted value
-   `ymin`: lower value of the confidence interval
-   `ymax`: upper value of the confidence interval
-   `se`: standard error

There's parameters such as `method` which determines which method is used to calculate the predictions and confidence interval, and some other arguments that are passed to that.

</div>

### Exercise <span class="exercise-number">3.5</span>. {.unnumbered .exercise}

<div class="question">
In our proportion bar chart, we need to set `group = 1` Why?
In other words what is the problem with these two graphs?
</div>

<div class="answer">

If `group` is not set to 1, then all the bars have `prop == 1`.
The function `geom_bar()` assumes that the groups are equal to the `x` values, since the stat computes the counts within the group.

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop..))
```

The problem with these two plots is that the proportions are calculated within the groups.
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop..))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```

This is more likely what was intended:
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop.., group = color))
```

</div>

## Position Adjustments

### Exercise <span class="exercise-number">3.8.1</span>. {.unnumbered .exercise}

<div class="question">
What is the problem with this plot?
How could you improve it?
</div>

<div class="answer">

There is overplotting because there are multiple observations for each combination of `cty` and `hwy`.
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point()
```
I'd fix it by using a jitter position adjustment.
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point(position = "jitter")
```

</div>

### Exercise <span class="exercise-number">3.8.2</span>. {.unnumbered .exercise}

<div class="question">
What parameters to `geom_jitter()` control the amount of jittering?
</div>

<div class="answer">

From the [`geom_jitter()`](https://ggplot2.tidyverse.org/reference/geom_jitter.html) documentation, there are two arguments to jitter:

-   `width` controls the amount of vertical displacement, and
-   `height` controls the amount of horizontal displacement.

The defaults values of `width` and `height` will introduce noise in both directions.
Here is what the plot looks like with the default values of `height` and `width`.
```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point(position = position_jitter(width = 0))
```

However, we can adjust them. Here are few examples to understand how adjusting
these parameters affects the look of the plot.

With `width = 0` there is no horizontal jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(width = 0)
```

With `width = 20`, there is too much horizontal jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(width = 20)
```

With `height = 0`, there is no vertical  horizontal jitter:

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(height = 0)
```

With `height = 15`, there is too much vertical jitter.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter(height = 15)
```

Note that the `height` and `width` arguments are in the units of the data.
Thus `height = 1` corresponds to different relative amounts of jittering depending on the scale of the `y` variable.
The default values of `height` and `width` are defined to be 80% of the `resolution()` of the data, which is the smallest non-zero distance between adjacent values of a variable. 
This means that if `x` and `y` are discrete variables, their resolutions are both equal to 1, and `height = 0.4` and `width = 0.4`,
since the jitter moves the points in both positive and negative directions.

</div>

### Exercise <span class="exercise-number">3.8.3</span>. {.unnumbered .exercise}

<div class="question">
Compare and contrast `geom_jitter()` with `geom_count()`.
</div>

<div class="answer">

`geom_jitter()` adds random variation to the locations points of the graph.
In other words, it "jitters" the locations of points slightly.
This method reduces overplotting since no two points are likely to have the same location after this random variation is added to their locations.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_jitter()
```

However, the reduction in overlapping comes at the cost of slightly changing the `x` and `y` values of the points.

`geom_count()` resizes the points relative to the number of observations at each location.
In other words, points with more observations will be larger than those with fewer observations.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_count()
```

This method does not change the `x` and `y` coordinates of the points.
However, if the points are close together and counts are large, the size of some
points can itself introduce overplotting.
For example, in the following example a third variable mapped to color is added to the plot. In this case, `geom_count()` is less readable than `geom_jitter()` when adding a third variable as color aesthetic.

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = class)) +
  geom_jitter()
```

```{r}  
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = class)) +
  geom_count()
```

Unfortunately, there is no universal solution to overplotting. 
The costs and benefits of different approaches will depend on the structure of the data and the goal of the data scientist.

</div>

### Exercise <span class="exercise-number">3.8.4</span>. {.unnumbered .exercise}

<div class="question">
What’s the default position adjustment for `geom_boxplot()`? Create a visualization of the mpg dataset that demonstrates it.
</div>

<div class="answer">

The default position for `geom_boxplot()` is `"dodge2"`, which is a shortcut for `position_dodge2`.
This does not change the vertical position of a geom, but moves a geom horizontally in order to avoid overlapping another geom.
See the documentation for [`position_dodge2()`](https://ggplot2.tidyverse.org/reference/position_dodge.html) for additional discussion on how it works.

When we add `colour = class` to the box plot, the different classes within `drv` are placed side by side, i.e. dodged. 
```{r}
ggplot(data = mpg, aes(x = drv, y = hwy, colour = class)) +
  geom_boxplot()
```
If `position_identity()` is used the boxplots overlap.
```{r}
ggplot(data = mpg, aes(x = drv, y = hwy, colour = class)) +
  geom_boxplot(position = "identity")
```

</div>

## Coordinate Systems

### Exercise <span class="exercise-number">3.9.1</span> {.unnumbered .exercise}

<div class="question">
Turn a stacked bar chart into a pie chart using `coord_polar()`.
</div>

<div class="answer">
This is a stacked bar chart with a single category
```{r}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar()
```

See the documentation for [coord_polar](https://ggplot2.tidyverse.org/reference/coord_polar.html) for an example of making a pie chart. In particular, `theta = "y"`, meaning that the angle of the chart is the `y` variable which has to be specified.

```{r}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
```

If `theta = "y"` is not specified, then you get a bull’s-eye chart
```{r}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar()
```

If you had a multiple stacked bar chart,
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

and apply polar coordinates to it, you end up with a multi-doughnut chart,
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") +
  coord_polar(theta = "y")
```

</div>

### Exercise <span class="exercise-number">3.9.2</span> {.unnumbered .exercise}

<div class="question">
What does `labs()` do? Read the documentation.
</div>

<div class="answer">

The `labs` function allows for adding axes titles, plot titles, and a caption to the plot.

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = "Highway MPG", 
       x = "Year", 
       title = "Highway MPG by car class",
       subtitle = "1999-2008",
       caption = "Source: http://fueleconomy.gov")
```

The arguments to `labs()` are optional, so you can add as many or as few of these as are needed.
```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip() +
  labs(y = "Highway MPG", 
       x = "Year", 
       title = "Highway MPG by car class")
```

This is not the only function you can use for adding titles.
You can also add axis labels with `xlab()`, `ylab()`, and the scale functions, and titles with `ggtitle()`.

</div>

### Exercise <span class="exercise-number">3.9.3</span> {.unnumbered .exercise}

<div class="question">
What’s the difference between `coord_quickmap()` and `coord_map()`?
</div>

<div class="answer">

`coord_map()` uses map projection to project 3-dimensional Earth onto a 2-dimensional plane.
By default, `coord_map()` uses the [Mercator projection](https://en.wikipedia.org/wiki/Mercator_projection).
However, this projection must be applied to all geoms in the plot.
`coord_quickmap()` uses a faster, but approximate map projection.
This approximation ignores the curvature of Earth and adjusts the map for the  latitude/longitude ratio.
This transformation is quicker than `coord_map()` because the coordinates of the individual geoms do not need to be transformed.

The **ggplot2** [documentation](https://ggplot2.tidyverse.org/reference/coord_map.html)
contains more information on and examples for these two functions.

</div>

### Exercise <span class="exercise-number">3.9.4</span> {.unnumbered .exercise}

<div class="question">
What does the plot below tell you about the relationship between city and highway mpg? Why is `coord_fixed()` important?
What does `geom_abline()` do?
</div>

<div class="answer">

The function `coord_fixed()` ensures that the line produced by `geom_abline()` is at a 45 degree angle.
The 45 degree line makes it easy to compare the highway and city mileage to the case in which city and highway MPG were equal.

```{r}
p <- ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline()
p + coord_fixed()
```

If we didn't include `geom_coord()`, then the line would no longer have an angle of 45 degrees.
```{r}
p
```

On average, humans are best able to perceive differences in angles relative to 45 degrees.
See @Cleveland1993, @Cleveland1994,@Cleveland1993a, @ClevelandMcGillMcGill1988,  @HeerAgrawala2006 for discussion on how the aspect ratio of a plot affects perception of the values it encodes, evidence that 45 degrees is generally optimal, and methods to calculate the an aspect ratio to achieve it.
The function `ggthemes::bank_slopes()` will calculate the optimal aspect ratio to bank slopes to 45 degrees.

</div>

## The Layered Grammar of Graphics

No exercises
