---
output: html_document
editor_options:
  chunk_output_type: console
---
# Workflow Basics

## Coding basics

No exercises

## What's in a name?

No exercises

## Calling functions

No exercises

## Practice

### Exercise 1 {.unnumbered .exercise}

`r BeginQuestion()`
Why does this code not work?
```{r error=TRUE}
my_variable <- 10
my_varıable
```
`r EndQuestion()`

The variable being printed is `my_varıable`, not `my_variable`:
the seventh character is "ı" ("LATIN SMALL LETTER DOTLESS I"), not "i".

While it wouldn't have helped much in this case, the importance of
distinguishing characters in code is reasons why fonts which clearly
distinguish similar characters are preferred in programming: especially
important are distinguishing between zero (0), Latin small letter O (o), and
Latin capital letter O (O); and the numeral one (1), Latin small letter I (i),
Latin capital letter I (I), and Latin small letter L (l). In these fonts, zero
and the Latin letter O are often distinguished by using a glyph for zero that
uses either a dot in the interior or a slash through it.

Error messages like `"object '...' not found"` mean exactly what they say. R
cannot find an object with that name.  Unfortunately, the error does not tell
you why that object cannot be found, because R does not know the reason that
the object does not exist. This is usually because you either (1) forgot to
define the function (or had an error that prevented it from being defined
earlier), (2) didn't load a package with the object, or (3) made a typo in the
object's name (either when using it or when you originally defined it).

### Exercise 2 {.unnumbered .exercise}

`r BeginQuestion()`
Tweak each of the following R commands so that they run correctly:
`r EndQuestion()`

```{r error=TRUE}
library("tidyverse")

ggplot(dota = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```
The error message is `argument "data" is missing, with no default`.

It looks like a typo, `dota` instead of `data`.
```{r error=TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r error=TRUE}
fliter(mpg, cyl = 8)
```

R could not find the function `fliter` because we made a typo: `fliter` instead of `filter`.

```{r error=TRUE}
filter(mpg, cyl = 8)
```

We aren't done yet. But the error message gives a suggestion. Let's follow it.

```{r error=TRUE}
filter(mpg, cyl == 8)
```

```{r error=TRUE}
filter(diamond, carat > 3)
```

R says it can't find the object `diamond`.
This is a typo; the data frame is named `diamonds`.
```{r error=TRUE}
filter(diamonds, carat > 3)
```

How did I know? I started typing in `diamond` and RStudio completed it to `diamonds`.
Since `diamonds` includes the variable `carat` and the code works, that appears to have been the problem.

### Exercise 3 {.unnumbered .exercise}

`r BeginQuestion()`
Press *Alt + Shift + K*. What happens? How can you get to the same place using the menus?
`r EndQuestion()`

This gives a menu with keyboard shortcuts. This can be found in the menu under `Tools -> Keyboard Shortcuts Help`.
