---
title: "Diamond sizes"
date: 2018-07-15
output: html_document
---

```{r knitr_opts, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r setup, message = FALSE}
library("ggplot2")
library("dplyr")

smaller <- diamonds %>%
  filter(carat <= 2.5)
```

```{r include = FALSE}
# Hide objects and functions ONLY used inline
n_larger <- nrow(diamonds) - nrow(smaller)
pct_larger <- n_larger / nrow(diamonds) * 100

comma <- function(x) {
  format(x, digits = 2, big.mark = ",")
}
```

## Size and Cut, Color, and Clarity

Diamonds with lower quality cuts (cuts are ranked from "Ideal" to "Fair") tend 
to be be larger.
```{r}
ggplot(diamonds, aes(y = carat, x = cut)) +
  geom_boxplot()
```
Likewise, diamonds with worse color (diamond colors are ranked from J (worst)
to D (best)) tend to be larger:

```{r}
ggplot(diamonds, aes(y = carat, x = color)) +
  geom_boxplot()
```

The pattern present in cut and color is also present in clarity. Diamonds with 
worse clarity  (I1 (worst), SI1, SI2, VS1, VS2, VVS1, VVS2, IF (best)) tend to
be larger:

```{r}
ggplot(diamonds, aes(y = carat, x = clarity)) +
  geom_boxplot()
```

These patterns are consistent with there being a profitability threshold for 
retail diamonds that is a function of carat, clarity, color, cut and other 
characteristics. A diamond may be profitable to sell if a poor value of one
feature, for example, poor clarity, color, or cut, is be offset by a good value
of another feature, such as a large size. This can be considered an example
of [Berkson's paradox](https://en.wikipedia.org/wiki/Berkson%27s_paradox).

## Largest Diamonds

We have data about `r comma(nrow(diamonds))` diamonds. Only
`r n_larger` (`r round(pct_larger, 1)`%) are larger
than 2.5 carats. The distribution of the remainder is shown below:

```{r}
smaller %>%
  ggplot(aes(carat)) +
  geom_freqpoly(binwidth = 0.01)
```

The frequency distribution of diamond sizes is marked by spikes at
whole-number and half-carat values, as well as several other carat values
corresponding to fractions.

The largest twenty diamonds (by carat) in the datasets are,

```{r results = "asis"}
diamonds %>%
  arrange(desc(carat)) %>%
  slice(1:20) %>%
  select(carat, cut, color, clarity) %>%
  knitr::kable(
    caption = "The largest 20 diamonds in the `diamonds` dataset."
  )
```

Most of the twenty largest datasets are in the lowest clarity category ("I1"),
with one being in the second best category ("VVS2") The top twenty diamonds 
have colors ranging from the worst, "J", to best, "D",categories, though most
are in the lower categories "J" and "I". The top twenty diamonds are more evenly
distributed among the cut categories, from "Fair" to "Ideal", although the worst
category (Fair) is the most common.
